# GO semantic similarity analysis {#GOSemSim}

```{r}
#| label: load-packages
#| echo: false
#| message: false
#| cache: false

source("_common.R")
loadNamespace("Rcpp")
```


`r Biocpkg("GOSemSim")` implemented all methods described in [Chapter 1](#semantic-similarity-overview), including four IC-based methods and one graph-based method. 

## Semantic data {#semantic-data}

To measure semantic similarity, we need to prepare GO annotations including GO structure (i.e. GO term relationships) and gene to GO mapping. For IC-based methods, information of GO term is species specific. We need to calculate `IC` for all GO terms of a species before we measure semantic similarity.


`r Biocpkg("GOSemSim")` provides the `godata()` function to prepare semantic data to support measuring GO and gene similarity. It internally uses the `r Biocpkg("GO.db")` package to obtain GO structure and `OrgDb` for gene to GO mapping. 


```{r godata}
library(AnnotationDbi)
library(GOSemSim)
hsGO <- godata(annoDb = 'org.Hs.eg.db', ont="MF")
```

User can set `computeIC=FALSE` if they only want to use Wang's method.


## Supported organisms {#gosemsim-supported-organisms}

`r Biocpkg("GOSemSim")` supports all organisms that have an `OrgDb` object available.

Bioconductor have already provided `OrgDb` for [about 20 species](http://bioconductor.org/packages/release/BiocViews.html#___OrgDb).

We can query `OrgDb` online via the `r Biocpkg("AnnotationHub")` package. For example:

```{r eval=FALSE}
#| label: query-orgdb
library(AnnotationHub)
hub <- AnnotationHub()
q <- query(hub, "Cricetulus")
id <- q$ah_id[length(q)]
Cgriseus <- hub[[id]]
```

If an organism is not supported by `r Biocpkg("AnnotationHub")`, user can use the `r Biocpkg("AnnotationForge")` package to build `OrgDb` manually.

Once we have `OrgDb`, we can build annotation data needed by `r Biocpkg("GOSemSim")` via `godata()` function described previously.


## GO semantic similarity measurement {#go-semantic-similarity}

The `goSim()` function calculates semantic similarity between two GO terms, while the `mgoSim()` function calculates semantic similarity between two sets of GO terms.


```{r gosemsim-gosim}
goSim("GO:0004022", "GO:0005515", semData=hsGO, measure="Jiang")
goSim("GO:0004022", "GO:0005515", semData=hsGO, measure="Wang")
go1 = c("GO:0004022","GO:0004024","GO:0004174")
go2 = c("GO:0009055","GO:0005515")
mgoSim(go1, go2, semData=hsGO, measure="Wang", combine=NULL)
mgoSim(go1, go2, semData=hsGO, measure="Wang", combine="BMA")
```

## Gene semantic similarity measurement {#gene-go-semantic-similarity}

On the basis of semantic similarity between GO terms, [GOSemSim](https://www.bioconductor.org/packages/GOSemSim) can
also compute semantic similarity among sets of GO terms, gene products, and gene clusters.

Suppose we have gene $g_1$ annotated by GO terms sets $GO_{1}=\{go_{11},go_{12} \cdots go_{1m}\}$
and $g_2$ annotated by $GO_{2}=\{go_{21},go_{22} \cdots go_{2n}\}$, `r Biocpkg("GOSemSim")` implemented four combining methods, including __*max*__, __*avg*__, __*rcmax*__, and __*BMA*__, to aggregate semantic similarity scores of multiple GO terms (see also @sec-combine-methods). The similarities
among gene products and gene clusters which are annotated by multiple GO
terms are also calculated by these combining methods.


`r Biocpkg("GOSemSim")` provides `geneSim()` to calculate semantic similarity between two gene products, and `mgeneSim()` to calculate semantic similarity among multiple gene products.

```{r gosemsim-genesim}
GOSemSim::geneSim("241", "251", semData=hsGO, measure="Wang", combine="BMA")
mgeneSim(genes=c("835", "5261","241", "994"),
         semData=hsGO, measure="Wang",verbose=FALSE)
mgeneSim(genes=c("835", "5261","241", "994"),
       semData=hsGO, measure="Rel",verbose=FALSE)
```

By default, `godata` function use `ENTREZID` as keytype, and the input ID type is `ENTREZID`. User can use other ID types such as `ENSEMBL`, `UNIPROT`, `REFSEQ`, `ACCNUM`, `SYMBOL` _et al_.

Here as an example, we use `SYMBOL` as `keytype` and calculate semantic similarities among several genes by using their gene symbol as input.

```{r gosemsim-mgeneSim}
hsGO2 <- godata('org.Hs.eg.db', keytype = "SYMBOL", ont="MF", computeIC=FALSE) 
genes <- c("CDC45", "MCM10", "CDC20", "NMU", "MMP1")
mgeneSim(genes, semData=hsGO2, measure="Wang", combine="BMA", verbose=FALSE)
```

Users can also use [`clusterProfiler::bitr`](#bitr) to translate biological IDs.

## Gene cluster semantic similarity measurement {#gene-cluster-go-semantic-similarity}


`r Biocpkg("GOSemSim")` also supports calculating semantic similarity between two gene clusters using `clusterSim()` function and measuring semantic similarity among multiple gene clusters using `mclusterSim()` function.


```{r echo=FALSE}
#| label: clusterSim-aliases
clusterSim <- GOSemSim::clusterSim
mclusterSim <- GOSemSim::mclusterSim
```


```{r gosemsim-clusterSim}
gs1 <- c("835", "5261","241", "994", "514", "533")
gs2 <- c("578","582", "400", "409", "411")
clusterSim(gs1, gs2, semData=hsGO, measure="Wang", combine="BMA")

library(org.Hs.eg.db)
x <- org.Hs.egGO
hsEG <- mappedkeys(x)
set.seed <- 123
clusters <- list(a=sample(hsEG, 20), b=sample(hsEG, 20), c=sample(hsEG, 20))
mclusterSim(clusters, semData=hsGO, measure="Wang", combine="BMA")
```


## Cross-species gene similarity calculation {#cross-species-gosemsim}

While `GOSemSim` primarily focuses on semantic similarity within a single species, researchers often need to calculate GO semantic similarity for orthologous gene pairs between different species. This is particularly relevant for comparative genomics studies between well-annotated species like human and mouse.

### Current limitations and workarounds

By default, `godata()` function only accepts a single `OrgDb` object, meaning it supports semantic similarity calculation within one species. However, there are several approaches to address cross-species similarity calculation:

1. **Graph-based methods**: Use Wang's method (graph-based) with `mgoSim()` function by manually providing GO term sets for genes from different species
2. **Manual GO term mapping**: Extract GO annotations for orthologous genes and use `mgoSim()` directly
3. **Future development**: A potential `merge()` function for combining `GOSemSimDATA` objects from multiple species

```{r eval=FALSE}
#| label: cross-species-sim
# Example workflow for cross-species similarity using mgoSim
# Extract GO terms for human gene using clusterProfiler
library(clusterProfiler)
human_go_terms <- bitr("human_gene_id", fromType = "ENTREZID", 
                       toType = "GO", OrgDb = org.Hs.eg.db)$GO

# Extract GO terms for mouse ortholog  
mouse_go_terms <- bitr("mouse_gene_id", fromType = "ENTREZID",
                       toType = "GO", OrgDb = org.Mm.eg.db)$GO

# Calculate similarity using Wang's method (graph-based)
similarity <- mgoSim(human_go_terms, mouse_go_terms, 
                    semData=hsGO, measure="Wang", combine="BMA")
```

For species not supported by Bioconductor `OrgDb` packages, users can:
- Use `AnnotationHub` to query and retrieve organism-specific annotations
- Build custom `OrgDb` objects using `AnnotationForge::makeOrgPackage()` or `makeOrgPackageFromNCBI()`
- Perform manual GO annotation followed by semantic similarity calculation

## Applications in functional genomics {#gosemsim-applications}

`GOSemSim` has been widely applied in various research domains, with over 200 citations. One particularly valuable application is in prioritizing genes for experimental validation from large sets of differentially expressed genes.

### Functional similarity for target prioritization

When faced with numerous differentially expressed genes, researchers can use GO semantic similarity to identify functionally related "hub" genes that may serve as better candidates for experimental validation. The approach involves:

1. Calculating pairwise semantic similarity among all differentially expressed genes
2. Constructing a similarity network where edges represent functional relationships
3. Identifying genes with high betweenness centrality as potential key regulators

### Case study: FMNL1 interactome analysis

In a collaboration with Technische Universität München, we used GO semantic similarity to analyze proteins identified through Co-IP and LC-MS/MS experiments targeting FMNL1 in hematopoietic cells. The workflow was:

1. **Similarity metric definition**: Functional similarity was defined as the geometric mean of molecular function and cellular component GO semantic similarities
2. **Network construction**: Created a similarity network of all co-purified proteins
3. **Hub identification**: Identified AHNAK1, SIPA1, and FLII as central nodes in the network
4. **Experimental validation**: Validated the novel interaction between FMNL1 and AHNAK1, leading to discovery of FMNL1's role in calcium-dependent membrane plasticity

This approach successfully identified previously unreported protein interactions that were subsequently validated experimentally, demonstrating the power of GO semantic similarity in guiding biological discovery.

### Implementation considerations

When using semantic similarity for target prioritization:
- Combine multiple GO aspects (MF, BP, CC) for comprehensive functional assessment
- Consider both direct interactions and functional relationships
- Use appropriate similarity measures based on biological context
- Validate computational predictions with experimental evidence

## References {-}
