# GO enrichment analysis {#clusterprofiler-go}

```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
loadNamespace("Rcpp")
```


GO comprises three orthogonal ontologies, i.e. molecular function (MF), biological
process (BP), and cellular component (CC).




```{r include=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, cache=TRUE)
library(clusterProfiler)
library(org.Hs.eg.db)
```


## Supported organisms {#clusterProfiler-go-supported-organisms}

GO analyses (`groupGO()`, `enrichGO()` and `gseGO()`) support organisms that have an `OrgDb` object available (see also [session 2.2](#gosemsim-supported-organisms)).


If a user has GO annotation data (in a `data.frame` format with the first column as gene ID and the second column as GO ID), they can use the `enricher()` and `GSEA()` functions to perform an over-representation test and gene set enrichment analysis. Users can also read GO annotation from GMT files using the `read.gmt()` function, which parses the file into a `data.frame` suitable for these functions.

If the genes are annotated by direct annotation, they should also be annotated by their ancestor GO nodes (indirect annotation). If a user only has direct annotation, they can pass their annotation to the `buildGOmap` function, which will infer indirect annotation and generate a `data.frame` that is suitable for both `enricher()` and `GSEA()`.

## Direct vs indirect GO annotation

One common issue users encounter is the discrepancy between direct gene-to-GO mapping and the results from enrichment analysis. For instance, a user might find that a specific GO term is enriched, but when they check the input gene list against the direct annotations, the term doesn't appear to be associated with many genes.

This happens because GO structure is a Directed Acyclic Graph (DAG). If a gene is annotated to a specific term (e.g., "glucose metabolic process"), it is implicitly annotated to all its ancestor terms (e.g., "metabolic process"). `enrichGO` and `gseGO` account for this **indirect annotation** by propagating annotations up the graph.

If you are providing your own annotation file (e.g., using `enricher` or `GSEA`), you must ensure that it includes these indirect annotations. If your annotation file only contains direct annotations, you can use the `buildGOmap()` function to generate the full annotation set (direct + indirect) before performing enrichment analysis.

```{r eval=FALSE}
# Assuming 'gomap' is a data.frame with direct annotations (GeneID, GOID)
# Build indirect annotations
gomap_with_ancestors <- buildGOmap(gomap)

# Use the result for enrichment
enricher(gene, TERM2GENE = gomap_with_ancestors, ...)
```

This step is crucial for accurate GO enrichment analysis when not using `OrgDb` objects.

## GO classification

In `r Biocpkg("clusterProfiler")`, the `groupGO()` function is designed for gene classification based on GO distribution at a specific level. Here we use the dataset `geneList` provided by `r Biocpkg("DOSE")`. 

```{r clusterprofiler-groupgo, warning=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)

data(geneList, package="DOSE")
gene <- names(geneList)[abs(geneList) > 2]

# Entrez gene ID
head(gene)

ggo <- groupGO(gene     = gene,
               OrgDb    = org.Hs.eg.db,
               ont      = "CC",
               level    = 3,
               readable = TRUE)

head(ggo)
```

The `gene` parameter is a vector of gene IDs (can be any ID type that is supported by the corresponding `OrgDb`, see also @sec-id-convert). If `readable` is set to `TRUE`, the input gene IDs will be converted to gene symbols.


## GO over-representation analysis {#clusterprofiler-go-ora}


The `r Biocpkg("clusterProfiler")` package implements `enrichGO()` for gene ontology over-representation test.


```{r}
ego <- enrichGO(gene          = gene,
                universe      = names(geneList),
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
		readable      = TRUE)
head(ego)
```

Any gene ID type that is supported in `OrgDb` can be directly used in GO analyses. Users need to specify the `keyType` parameter to specify the input gene ID type.

However, the GO annotation in `OrgDb` is indexed by ENTREZID. If the input ID type is not ENTREZID, `clusterProfiler` will first map the input IDs to ENTREZID using the `bitr()` function. This mapping step may lead to a loss of information, as some IDs might not have a corresponding ENTREZID or multiple IDs might map to the same ENTREZID (many-to-one mapping).

To address this issue, `enrichGO` and `gseGO` perform the mapping and use the mapped ENTREZIDs for enrichment analysis. The results are then mapped back to the original input ID type. This ensures that the gene counts and gene ratios in the final result accurately reflect the input IDs, avoiding discrepancies caused by the intermediate mapping step.

For example, if we use ENSEMBL IDs as input:

```{r}
gene.df <- bitr(gene, fromType = "ENTREZID",
	   	toType = c("ENSEMBL", "SYMBOL"),
		OrgDb = org.Hs.eg.db)

ego2 <- enrichGO(gene         = gene.df$ENSEMBL,
                OrgDb         = org.Hs.eg.db,
                keyType       = 'ENSEMBL',
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05)
head(ego2, 3)                
```

Gene IDs can be mapped to gene Symbols by using the parameter `readable=TRUE` or [`setReadable()` function](#sec-setReadable).


## GO Gene Set Enrichment Analysis

The `r Biocpkg("clusterProfiler")` package provides the `gseGO()` function for [gene set enrichment analysis](#gsea-algorithm) using gene ontology.

```{r eval=FALSE}
ego3 <- gseGO(geneList     = geneList,
              OrgDb        = org.Hs.eg.db,
              ont          = "CC",
              minGSSize    = 100,
              maxGSSize    = 500,
              pvalueCutoff = 0.05,
              verbose      = FALSE)
```

The format of input data, `geneList`, is documented in the @sec-id-convert. Beware that only gene sets with size in `[minGSSize, maxGSSize]` will be tested.



## GO analysis for non-model organisms {#clusterprofiler-go-non-model}

Both the `enrichGO()` and `gseGO()` functions require an `OrgDb` object as the background annotation. For organisms that don't have `OrgDb` provided by `Bioconductor`, users can query one (if available) online via `r Biocpkg("AnnotationHub")`. If there is no `OrgDb` available, users can obtain GO annotation from other sources, e.g. from `r Biocpkg("biomaRt")`, or annotate the genes using [Blast2GO](https://www.blast2go.com/) or the [Trinotate](https://rnabio.org/module-07-trinotate/0007/02/01/Trinotate/) pipeline. Then the `enricher()` or `GSEA()` functions can be used to perform GO analysis for these organisms, similar to the examples using WikiPathways and MSigDB. Another solution is to create an `OrgDb` on your own using the `r Biocpkg("AnnotationForge")` package.


Here is an example of querying GO annotation from Ensembl using `r Biocpkg("biomaRt")`. 


```{r eval=FALSE}
library(biomaRt)
ensembl <- useEnsemblGenomes(biomart = "plants_mart", dataset = "nattenuata_eg_gene")
gene2go <- getBM(attributes =c("ensembl_gene_id", "go_id"), mart=ensembl)
```

Alternatively, you can use `AnnotationHub` to query and retrieve `OrgDb` objects for a wide range of organisms, including non-model species.

**Example: GO enrichment analysis for Maize (Zea mays)**

```{r eval=FALSE}
library(AnnotationHub)
hub <- AnnotationHub()

# Query for Zea mays OrgDb
query(hub, "zea")
# Retrieve the OrgDb (e.g., AH55736, assume it's the correct record)
maize <- hub[['AH55736']]

# Check keys
length(keys(maize))
columns(maize)

# Perform enrichment analysis
# Assuming 'sample_genes' is a vector of Entrez IDs
res <- enrichGO(sample_genes, OrgDb=maize, pvalueCutoff=0.05, qvalueCutoff=0.05)
```

This approach allows you to use the standard `enrichGO` workflow with organisms not included in the default Bioconductor `OrgDb` packages.

## Visualization of GO enrichment results

The `clusterProfiler` package provides several visualization methods to help interpret enrichment results, including `barplot`, `dotplot`, `cnetplot`, `emapplot`, `goplot` and `plotGOgraph`.

Please refer to the @sec-enrichplot chapter for details.


## Filtering and simplifying GO terms

GO enrichment analysis often results in many redundant terms due to the hierarchical structure of GO. Parent terms and child terms often share a large proportion of genes, leading to multiple significant results that represent similar biological processes.

### Removing specific terms or levels

The `dropGO()` function can be used to remove specific GO terms or GO levels from the results. The `enrichGO()` function tests the whole GO corpus and enriched result may contains very general terms. If users want to restrict the result to a specific GO level (e.g., level 3 or 4), they can use the `gofilter()` function. Both functions work with results obtained from `enrichGO()`, `gseGO()` and `compareCluster()`.

### Simplifying enriched GO terms

To reduce redundancy, `clusterProfiler` provides a `simplify` method. It uses semantic similarity (via `GOSemSim`) to calculate the similarity between enriched terms and removes highly similar terms, keeping the most significant one.

The `simplify` method applies `select_fun` (which can be a user-defined function) to the feature specified by `by` (e.g., p.adjust) to select one representative term from redundant terms (which have similarity higher than `cutoff`). This results in a cleaner and more interpretable visualization.

In @fig-simplify-emapplot(A), we can found that there are many redundant terms form a highly condense network. After removing redundant terms using the `simplify()` method, the result is more clear to view the whole story.

```{r}
#| label: fig-simplify-emapplot
#| fig-width: 16
#| fig-height: 8
#| fig-scap: |
#|   Visualize enriched terms by EnrichmentMap using the `emapplot()` function.
#| fig-cap: |
#|   **Visualize enriched terms by EnrichmentMap using the `emapplot()` function.**
#|   (A) original result. (B) simplify result.
#|
library(enrichplot)
data(geneList, package="DOSE")
de <- names(geneList)[abs(geneList) > 2]
bp <- enrichGO(de, ont="BP", OrgDb = 'org.Hs.eg.db')
bp <- pairwise_termsim(bp)
bp2 <- simplify(bp, cutoff=0.7, by="p.adjust", select_fun=min)
p1 <- emapplot(bp)
p2 <- emapplot(bp2)
plot_list(p1, p2, ncol=2, tag_levels = 'A')
```

Alternatively, users can use slim version of GO and use the `enricher()` or `gseGO()` functions to analyze.

**Note on `enricher` results**: The `simplify()` method is designed for `enrichGO()` results where the ontology (BP, CC, or MF) is known. If you use `enricher()` to perform GO analysis (e.g., with custom annotation), the result object does not strictly define the ontology. To use `simplify()`, you must explicitly set the ontology slot in the result object:

```{r eval=FALSE}
# Assuming 'res' is the result from enricher() using GO annotation
# Set the ontology (must be one of "BP", "MF", "CC")
res@ontology <- "BP" 
res_simplified <- simplify(res)
```


## Troubleshooting

### No results found

It is common to receive a message saying "No gene set have size > 10 ... return NULL" or finding no significant terms.

1.  **Gene Set Size**: The default `minGSSize` is 10 and `maxGSSize` is 500. If your background annotations for a specific term have fewer than 10 genes or more than 500, they are excluded. You can adjust these parameters.
2.  **Universe**: If you provide a custom `universe` (background gene list), ensure it is large enough. A small universe can make it statistically difficult to find enrichment.
3.  **P-value Adjustment**: `clusterProfiler` uses multiple testing correction (e.g., BH) by default. Some web tools might use raw p-values or different thresholds, leading to more (but potentially false positive) results. `clusterProfiler` tends to be more conservative and reliable.

### Annotation quality

Sometimes, the `OrgDb` provided by Bioconductor might be outdated compared to the latest online databases (e.g., TAIR for Arabidopsis).

*   **Check Dates**: Always check the metadata of the `OrgDb` object to see the source and date of the annotation.
*   **Custom Annotation**: If the `OrgDb` is too old, consider downloading the latest annotation (e.g., GAF or GOSLIM files) from the organism's primary database and using `enricher()`/`GSEA()` with `buildGOmap()` as described in the [Direct vs indirect GO annotation] section.

## Summary {#clusterprofiler-go-summary}

GO semantic similarity can be calculated by `r Biocpkg("GOSemSim")` [@yu2010]. We can use it to cluster genes/proteins into different clusters based on their functional similarity and can also use it to measure the similarities among GO terms to reduce the redundancy of GO enrichment results.

## References {-}
