# Analysis with Other Knowledge Bases

```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
library(clusterProfiler)
```

## WikiPathways analysis

[WikiPathways](https://www.wikipathways.org) is a continuously updated pathway database curated by a community of researchers and pathway enthusiasts. WikiPathways produces monthly releases of GMT files for supported organisms at [data.wikipathways.org](http://data.wikipathways.org/current/gmt/). The `r Biocpkg("clusterProfiler")` package [@yu2012] supports enrichment analysis (either ORA or GSEA) for WikiPathways using the `enrichWP()` and `gseWP()` functions. These functions will automatically download and parse the latest WikiPathways GMT file for the selected organism.


Supported organisms can be listed by:

```{r get-wp-organisms}
library(clusterProfiler)
get_wp_organisms()
```


<!--


Download the appropriate gmt file and then generate `TERM2GENE` and `TERM2NAME` to use `enricher()` and `GSEA()` functions. 

Users can download WikiPathways manually:

```r
###################################
# download file manually
# 1. visit the website, http://data.wikipathways.org/current/gmt/, to get the url
# 2. use the following code to download it
####################################
url <- "http://data.wikipathways.org/current/gmt/wikipathways-20200810-gmt-Homo_sapiens.gmt"
wpgmtfile <- "wikiPathways-HS.gmt"
download.file(url, destfile = wpgmtfile)
```

As an alternative to manually downloading gmt files, install the `r Biocpkg("rWikiPathways")` to gain scripting access to the latest gmt files using the `downloadPathwayArchive()` function.


```r
## supported organisms can be accessed via the following command:
## rWikiPathways::listOrganisms()

wpgmtfile <- rWikiPathways::downloadPathwayArchive(organism="Homo sapiens", format = "gmt")
```

Once the GMT file is downloaded, we can use `read.gmt.wp()` to parse it. Note that the `read.gmt.wp()` function is designed for WikiPathways GMT files. For ordinary GMT files, please use the `read.gmt()` function.

```r
wp2gene <- read.gmt.wp(wpgmtfile)
#TERM2GENE
wpid2gene <- wp2gene %>% dplyr::select(wpid, gene) 
#TERM2NAME
wpid2name <- wp2gene %>% dplyr::select(wpid, name) 

ewp <- enricher(gene, TERM2GENE = wpid2gene, TERM2NAME = wpid2name)
head(ewp)

ewp2 <- GSEA(geneList, TERM2GENE = wpid2gene, TERM2NAME = wpid2name, verbose=FALSE)
head(ewp2)
```
-->


### WikiPathways over-representation analysis {#clusterprofiler-wikipathway-ora}

```{r enrichwp}
data(geneList, package="DOSE")
gene <- names(geneList)[abs(geneList) > 2]

enrichWP(gene, organism = "Homo sapiens") 
```

### WikiPathways gene set enrichment analysis {#clusterprofiler-wikipathway-gsea}


```{r gsewp}
gseWP(geneList, organism = "Homo sapiens")
```


If your input gene ID type is not Entrez gene ID, you can use the [`bitr()`](#bitr) function to convert gene ID. If you want to convert the gene IDs in output result to gene symbols, you can use the `setReadable()` function, see also @sec-setReadable.


## DAVID functional analysis

[DAVID](https://david.ncifcrf.gov/) is a popular bioinformatics resource for functional annotation and enrichment analysis. Although we recommend using `enrichGO` and `enrichKEGG` which use up-to-date data maintained by Bioconductor, `r Biocpkg("clusterProfiler")` provides `enrichDAVID` to support DAVID analysis for users who prefer it.

Users need to register an account on DAVID website and use the email address to access the web service.

```{r}
#| eval: false
require(clusterProfiler)
data(geneList, package="DOSE")
gene = names(geneList)[abs(geneList) > 2]
david = enrichDAVID(gene = gene, 
                    idType="ENTREZ_GENE_ID", 
                    listType="Gene", 
                    annotation="KEGG_PATHWAY",
                    david.user = "clusterProfiler@hku.hk")
```

The result can be visualized using the same functions as other enrichment results.

```{r}
#| eval: false
barplot(david)
cnetplot(david, foldChange=geneList)
```

### DAVID Web Service Limitations

DAVID Web Service has the following limitations:

+ A job with more than 3000 genes to generate gene or term cluster report will not be handled by DAVID due to resource limit.
+ No more than 200 jobs in a day from one user or computer.
+ DAVID Team reserves right to suspend any improper uses of the web service without notice.

For more details, please refer to [http://david.abcc.ncifcrf.gov/content.jsp?file=WS.html](http://david.abcc.ncifcrf.gov/content.jsp?file=WS.html).

### Custom Background

The `enrichDAVID` function also supports a custom background (universe) using the `universe` parameter.

```{r}
#| eval: false
enrichDAVID(gene = gene,
            universe = names(geneList),
            idType = "ENTREZ_GENE_ID",
            listType = "Gene",
            annotation = "KEGG_PATHWAY")
```

### Comparing DAVID Functional Profiles

Comparison of functional profiles among different gene clusters is also supported via `compareCluster`.

```{r}
#| eval: false
data(gcSample)
x=compareCluster(gcSample, fun="enrichDAVID", annotation="KEGG_PATHWAY")
plot(x)
```

### Note on Data Updates


As highlighted in a Nature Methods study (@wadi_impact_2016), DAVID has a very low update frequency, often going more than 5 years between updates, leading to a continuous decline in knowledge completeness. This means analysis results may only reflect outdated biological knowledge, which could introduce biases in current research. As shown in previous comparisons, using `clusterProfiler` with Bioconductor annotation data often yields more comprehensive results (more annotated genes and enriched pathways) compared to DAVID.

## References {-}
