# Universal enrichment analysis {#universal-api}



```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
library(clusterProfiler)
```



The `r Biocpkg("clusterProfiler")` package [@yu2012] supports both hypergeometric test and gene set enrichment analyses of many ontologies/pathways, but it's still not enough as users may want to analyze their data with unsupported organisms, slim versions of GO, novel functional annotations (e.g. GO via BlastGO or KEGG via KAAS), unsupported ontologies/pathways, or customized annotations.

The `r Biocpkg("clusterProfiler")` package provides `enricher()` function for hypergeometric test and `GSEA()` function for gene set enrichment analysis that are designed to accept user-defined annotations. They accept two additional parameters `TERM2GENE` and `TERM2NAME`. As indicated in the parameter names, `TERM2GENE` is a data.frame with the first column of term ID and the second column of corresponding mapped genes, and `TERM2NAME` is a `data.frame` with the first column of term ID and the second column of corresponding term names. `TERM2NAME` is optional.


## Input data

For over representation analysis, all we need is a gene vector, that is a vector of gene IDs. These gene IDs can be obtained by differential expression analysis (*e.g.* with the `r Biocpkg("DESeq2")` package). 

For gene set enrichment analysis, we need a ranked list of genes. `r Biocpkg("DOSE")` provides an example dataset `geneList` which was derived from `R` package `r Biocpkg("breastCancerMAINZ")` that contained 200 samples, including 29 samples in grade I, 136 samples in grade II and 35 samples in grade III. We computed the ratios of geometric means of grade III samples versus geometric means of grade I samples. Logarithm of these ratios (base 2) were stored in `geneList` dataset. If you want to prepare your own `geneList`, please refer to the [FAQ](#sec-genelist).


We can load the sample data into R via:

```{r}
#| label: load-genelist
data(geneList, package="DOSE")
head(geneList)
```

Suppose we define fold change greater than 2 as DEGs:

```{r}
#| label: define-de-genes
gene <- names(geneList)[abs(geneList) > 2]
head(gene)
```




## Cell Marker	

<!--     # tidyr::unite("cellMarker", tissueType, cancerType, cellName, sep=", ")  -->


```{r eval=FALSE}
#| label: cellmarker-data-prep
# not executed: it downloads an ~8 MB file, and the URL changes between
# CellMarker releases (see the note below).

## The original download endpoint no longer resolves:
##   http://bio-bigdata.hrbmu.edu.cn/CellMarker/download/Human_cell_markers.txt
## The file is now served from the CellMarker download page
##   http://117.50.127.228/CellMarker/CellMarker_download.html
## This direct link worked when this chapter was last checked:
url <- "http://www.bio-bigdata.center/CellMarker_download_files/file/Cell_marker_Human.xlsx"
f <- tempfile(fileext = ".xlsx")
download.file(url, f, mode = "wb")

cell_marker_data <- readxl::read_excel(f, 1)

## Column names differ between CellMarker releases (`Cell name` in older
## exports, `cell_name` in the current one), so check before selecting:
##   names(cell_marker_data)

## instead of `cell_name`, users can use other features (e.g. `cancer_type`)
cells <- cell_marker_data %>%
    dplyr::select(cell_name, GeneID) %>%
    dplyr::mutate(GeneID = strsplit(GeneID, ", ")) %>%
    tidyr::unnest(cols = c(GeneID))
```

::: {.callout-note}
**The CellMarker download URL is not stable.** The host has moved more than once
(`bio-bigdata.hrbmu.edu.cn` now refuses connections; the mirror that used to be
cited here, `yikedaxue.slwshop.cn`, no longer resolves), and the column headers
have been renamed between releases — older exports used `Cell name` and
`cellType`, the current one uses `cell_name` and `cell_type`. If the code above
fails, download the file from the CellMarker download page by hand and run
`names(cell_marker_data)` to see which names your copy uses, then adjust the
`dplyr::select()` call accordingly.
:::

### Cell Marker over-representation analysis {#cell-marker-ora}


```{r eval=FALSE}
#| label: cellmarker-ora
x <- enricher(gene, TERM2GENE = cells)
head(x)
```

### Cell Marker gene set enrichment analysis {#cell-marker-gsea}

```{r eval=FALSE}
#| label: cellmarker-gsea
y <- GSEA(geneList, TERM2GENE = cells)
head(y)
```

## MSigDb analysis


[Molecular Signatures Database](http://software.broadinstitute.org/gsea/msigdb) is a collection of annotated gene sets. It contains 8 major collections:

* H: hallmark gene sets
* C1: positional gene sets
* C2: curated gene sets
* C3: motif gene sets
* C4: computational gene sets
* C5: GO gene sets
* C6: oncogenic signatures
* C7: immunologic signatures

Users can download [GMT files](www.broadinstitute.org/cancer/software/gsea/wiki/index.php/Data_formats#GMT:_Gene_Matrix_Transposed_file_format_.28.2A.gmt.29) from [Broad Institute](http://software.broadinstitute.org/gsea/msigdb) and use the `read.gmt()` function to parse the file to be used in `enricher()` and `GSEA()`.

There is an R package, [msigdbr](https://cran.r-project.org/package=msigdbr), that already packed the MSigDB gene sets in tidy data format that can be used directly with `r Biocpkg("clusterProfiler")` [@yu2012].

It supports several species:

```{r}
#| label: msigdbr-species
library(msigdbr)
msigdbr_species()
```

We can retrieve human gene sets with `msigdbr()`. Since the complete human MSigDB table is large, it is usually better to request only the collection you need in analysis workflows.


```{r}
#| label: msigdbr-human
#| cache: false
msigdbr(species = "Homo sapiens", collection = "C6") |>
  head(2) |>
  as.data.frame()
```


Or specific collection. Here we use C6, oncogenic gene sets as an example:

```{r}
#| label: msigdbr-c6
#| cache: false
m_t2g <- msigdbr(species = "Homo sapiens", collection = "C6") |>
  dplyr::select(gs_name, ncbi_gene)
head(m_t2g)
```


### MSigDb over-representation analysis {#msigdb-ora}


```{r}
#| label: msigdb-ora-example
em <- enricher(gene, TERM2GENE=m_t2g)
head(em)
```


### MSigDb gene set enrichment analysis {#msigdb-gsea}

In over-representation analysis, we use oncogenic gene sets (i.e. C6) to test whether the DE genes are involved in the process that leads to cancer. In this example, we will use the C3 category to test whether genes are up/down-regulated by sharing specific motifs using the GSEA approach.


```{r}
#| label: msigdb-gsea-example
#| cache: false
C3_t2g <- msigdbr(species = "Homo sapiens", collection = "C3") |>
  dplyr::select(gs_name, ncbi_gene)
head(C3_t2g)

em2 <- GSEA(geneList, TERM2GENE = C3_t2g)
head(em2)
```


## Proteomics data and UniProt identifiers {#uniprot-enrichment}

Proteomics experiments normally report **UniProt accessions** rather than Entrez gene IDs, and the organism is frequently not one of the model organisms that have an `OrgDb` package. This is the situation where the usual `enrichGO()`-style entry points cannot be used directly, so it is worth spelling out the two ways out.

### Model organisms: convert the identifiers

If an `OrgDb` exists for your organism and carries the `UNIPROT` keytype, the simplest solution is to translate the accessions and then use the standard pipeline:

```{r}
#| label: uniprot-to-entrez
library(org.Hs.eg.db)

up <- c("P04637", "P00533", "Q9NQ11", "P38398", "O14746")
bitr(up, fromType = "UNIPROT", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
```

Use `keytypes(org.Hs.eg.db)` to check what is available for your organism — not every `OrgDb` provides `UNIPROT`.

### Any organism: build `TERM2GENE` from the EBI GOA files

For a non-model organism there is no `OrgDb` to convert with, but you do not need one. The Gene Ontology Annotation ([GOA](https://www.ebi.ac.uk/GOA)) group at EMBL-EBI publishes GO annotation for **complete proteomes** in the standard GAF format, keyed by UniProt accession:

* Browse the available proteomes: <https://www.ebi.ac.uk/GOA/proteomes>
* Download directory: <http://ftp.ebi.ac.uk/pub/databases/GO/goa/proteomes/>

A GAF file is plain tab-delimited text, so it can be read with `read.delim()`. In the GAF 2.x layout the columns we need are: 2 = UniProt accession, 4 = qualifier, 5 = GO ID, and 9 = ontology aspect (`P`/`C`/`F`).

```{r}
#| label: goa-download
#| eval: false

## example: Solanum lycopersicum (tomato), proteome UP000004994
url <- "http://ftp.ebi.ac.uk/pub/databases/GO/goa/proteomes/284179.S_lycopersicum_genome_cv_Heinz_1706.goa"
goa_file <- "tomato.goa"
download.file(url, goa_file)
```

```{r}
#| label: goa-parse
#| eval: false

library(GO.db)
library(AnnotationDbi)

gaf <- read.delim(goa_file, header = FALSE, comment.char = "!", sep = "\t",
                  quote = "", fill = TRUE, stringsAsFactors = FALSE)

colnames(gaf) <- c("DB", "UniProt", "Symbol", "Qualifier", "GO", "Reference",
                   "Evidence", "With", "Aspect", "Name", "Synonym", "Type",
                   "Taxon", "Date", "AssignedBy", "Extension", "GeneProductForm")

## drop negative annotations, which are written as "NOT|..."
gaf <- gaf[!grepl("^NOT", gaf$Qualifier), ]

## TERM2GENE: (GO id, UniProt accession)
term2gene <- unique(gaf[, c("GO", "UniProt")])

## TERM2NAME: GO term names come from GO.db
goterm <- AnnotationDbi::select(GO.db, keys = unique(term2gene$GO),
                                columns = c("TERM", "ONTOLOGY"), keytype = "GOID")
term2name <- goterm[, c("GOID", "TERM")]
```

These two data frames plug straight into `enricher()` and `GSEA()`, exactly like the MSigDb example above — the only difference is that the gene identifiers are UniProt accessions:

```{r}
#| label: uniprot-enricher
#| eval: false

res <- enricher(gene     = my_proteins,
                universe = all_proteins,
                TERM2GENE = term2gene,
                TERM2NAME = term2name)
```

Two details are easy to get wrong:

* **`universe` should be the proteins detected in your experiment**, not the entire proteome. The GOA file annotates the full proteome, but the relevant background for a mass-spectrometry experiment is the set of proteins you could have quantified. Passing the whole proteome instead inflates the significance of everything.
* **Aspect filtering.** If you only want one ontology, filter `term2gene` by the `ONTOLOGY` column of `goterm` before calling `enricher()`:

```{r}
#| label: uniprot-aspect
#| eval: false

bp_ids  <- goterm$GOID[goterm$ONTOLOGY == "BP"]
term2gene_bp <- term2gene[term2gene$GO %in% bp_ids, ]
```

If you need curated annotations only, filter on the evidence code as well — the GOA files include electronically inferred (`IEA`) annotations, and `gaf[gaf$Evidence != "IEA", ]` keeps just the manually reviewed ones.

### KEGG analysis with UniProt identifiers

KEGG does accept UniProt accessions directly, so no conversion is needed for organisms that KEGG covers:

```{r}
#| label: kegg-uniprot
#| eval: false

kk <- enrichKEGG(gene = my_proteins, organism = "sly", keyType = "uniprot")
```

Use `bitr_kegg()` to check the coverage first — for most proteomes a substantial fraction of accessions has no KEGG entry, and those proteins are silently dropped:

```{r}
#| label: bitr-kegg-uniprot
#| eval: false

uni2kegg <- bitr_kegg(unique(gaf$UniProt), fromType = "uniprot",
                      toType = "kegg", organism = "sly")
```

## References {-}
