# Biological theme comparison {#clusterprofiler-comparecluster}



```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
library(clusterProfiler)
library(org.Hs.eg.db)
library(DOSE)
```


The `r Biocpkg("clusterProfiler")` package was developed for biological theme comparison [@yu2012; @wu_clusterprofiler_2021], and it provides a function, `compareCluster`, to automatically calculate enriched functional profiles of each gene clusters and aggregate the results into a single object. Comparing functional profiles can reveal functional consensus and differences among different experiments and helps in identifying differential functional modules in omics datasets.

## Comparing multiple gene lists

The `compareCluster()` function applies selected function (via the `fun` parameter) to perform enrichment analysis for each gene list. 

```{r}
#| label: compare-multiple-gene-lists
data(gcSample)
str(gcSample) 
```

Users can use a named list of gene IDs as the input that passed to the `geneCluster` parameter. 

```{r}
#| label: compare-multiple-gene-lists-enrichGO
ck <- compareCluster(geneCluster = gcSample, fun = enrichGO, 
                    OrgDb = org.Hs.eg.db, ont="BP")
ck <- setReadable(ck, OrgDb = org.Hs.eg.db, keyType="ENTREZID")
head(ck) 
```


## Formula interface of compareCluster

As an alternative to using named list, the `compareCluster()` function also supports passing a formula to describe more complicated experimental designs (*e.g.*, $Gene \sim time + treatment$).


```{r}
#| label: compare-formula
data(geneList, package='DOSE')
mydf <- data.frame(Entrez=names(geneList), FC=geneList)
mydf <- mydf[abs(mydf$FC) > 1,]
mydf$group <- "upregulated"
mydf$group[mydf$FC < 0] <- "downregulated"
mydf$othergroup <- "A"
mydf$othergroup[abs(mydf$FC) > 2] <- "B"

formula_res <- compareCluster(Entrez~group+othergroup, data=mydf, 
                            fun=enrichGO, OrgDb = org.Hs.eg.db, ont="BP")
  
head(formula_res)
```

::: {.callout-note}
**`compareCluster()` takes gene lists, not enrichment tables.** This is the
single most common misunderstanding about the formula interface. Both the named
list and the formula form expect **genes**, and `compareCluster()` runs the
enrichment test itself, once per group:

- `data` must be a data frame with **one row per gene** — a gene identifier
  column, plus the grouping columns on the right-hand side of the formula.
- The statistics in the result are computed by `fun` (`enrichGO()` here) from
  each group's gene list. They are **not** carried over from any input table.

So if you already have two tables of enrichment results (`ID`, `Description`,
`GeneRatio`, `pvalue`, `p.adjust`, ...), there is nothing to join: those numbers
were produced by whatever run created them, on whatever background, and
`compareCluster()` has no way to reuse them. To compare the two conditions,
combine the **gene lists** that produced those tables and let
`compareCluster()` re-run the test:

```{r eval=FALSE}
## one row per gene, with a column naming the group it belongs to
genes <- rbind(
    data.frame(gene = condition_A_genes, group = "A"),
    data.frame(gene = condition_B_genes, group = "B")
)
res <- compareCluster(gene ~ group, data = genes, fun = "enrichGO",
                      OrgDb = org.Hs.eg.db, ont = "BP")
```

Re-running the test is not a limitation to work around — it is what makes the
two columns comparable. Enrichment p-values depend on the gene list, the
universe/background and the multiple-testing correction, so p-values computed
separately for A and B are not on a common footing and plotting them side by
side would be misleading.

If you want to carry a *pre-computed* gene-level statistic into the comparison
instead (a ranked list rather than a cut-off list), use the GSEA route:
`compareCluster(..., fun = "gseGO")`.
:::

## Functional analysis of single-cell marker genes

The `compareCluster()` function is highly versatile and can be directly integrated into single-cell analysis workflows. For example, after identifying marker genes for each cluster using `Seurat`, the results can be directly used for functional enrichment analysis.

### Using Seurat results

Typically, `FindAllMarkers()` from `Seurat` returns a data frame where rows are genes and columns include cluster information and statistical metrics.

This example is wrapped in `retry()` (defined in `_common.R`) because its `fun` reaches the remote KEGG service, so a transient failure is retried a few times before giving up; the `enrichGO` examples further on stay unwrapped because they run locally.

```{r eval=FALSE}
#| label: sc-seurat-example
library(Seurat)
library(SeuratData)
library(dplyr)
library(clusterProfiler)

# Load example data
data("pbmc3k")
sce <- pbmc3k.final

# Identify marker genes
sce.markers <- FindAllMarkers(object = sce, only.pos = TRUE, 
                              min.pct = 0.25, 
                              thresh.use = 0.25)

# Filter markers
markers <- sce.markers |> 
    group_by(cluster) |> 
    filter(p_val_adj < 0.001) |> 
    ungroup()

# ID conversion (if needed)
gid <- bitr(unique(markers$gene), 'SYMBOL', 'ENTREZID', OrgDb = 'org.Hs.eg.db')
markers <- full_join(markers, gid, by = c('gene' = 'SYMBOL'))

# Perform comparison using formula interface
x <- retry(compareCluster(ENTREZID ~ cluster, data = markers, fun = 'enrichKEGG'))

# Visualization
dotplot(x, label_format = 40) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.callout-note}
**Enrichment plots need enrichment result objects, not a table of pathway
statistics.** A frequent question is how to hand `enrichplot` a data frame of
per-pathway statistics (`PValue`, `FDR`, `NES`, `logFC`, ...) obtained elsewhere.
There is no way to do this directly: `dotplot()`, `gseaplot2()`, `ridgeplot()`
and friends dispatch on `enrichResult` / `gseaResult` / `compareClusterResult`
objects, and those carry the gene sets, the gene-to-set mapping and the
background that the plots need. A bare table has none of that.

The way in is to re-run the enrichment inside `clusterProfiler` on the same
gene list and gene sets, which gives you an object the plotting functions
understand:

- **Cut-off gene list per cluster** (what `FindAllMarkers()` gives you) →
  `compareCluster(ENTREZID ~ cluster, data = markers, fun = "enrichGO")`, then
  `dotplot()` / `cnetplot()` / `emapplot()`, as above.
- **A ranked list** (a statistic for every gene, not a cut-off list) →
  `gseGO()` / `GSEA()`, then `gseaplot2()` / `ridgeplot()` / `gseaplot()`.

If the statistics came from another tool (fgsea, GSEA desktop, a vendor
report), re-running the equivalent analysis in `clusterProfiler` with the same
gene list and gene sets is usually a couple of lines, and is the only way to get
the object the visualisation layer requires.
:::

### Using COSG results

Methods like `COSG` return marker genes as a list or data frame where columns represent clusters. Since a data frame is essentially a list of equal-length vectors, it can be passed directly to `compareCluster()`.

```{r eval=FALSE}
#| label: sc-cosg-example
library(COSG)

# Identify markers using COSG
marker_cosg <- cosg(
    sce,
    groups = 'all',
    assay = 'RNA',
    slot = 'data',
    mu = 1,
    n_genes_user = 100
)

# The first element is a data frame of gene symbols
# Columns correspond to clusters
head(marker_cosg[[1]])

# Directly use the data frame for enrichment
y <- compareCluster(marker_cosg[[1]], 
                    fun = 'enrichGO', 
                    OrgDb = 'org.Hs.eg.db', 
                    keyType = 'SYMBOL', 
                    ont = "MF")

# Visualization
dotplot(y, label_format = 60) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This demonstrates that `compareCluster()` can seamlessly handle various data structures commonly produced by single-cell analysis tools, simplifying the downstream functional interpretation.

## Visualization of functional profile comparison {#vis-compare-profile}

### Dot plot {#compare-dotplot}

We can visualize the result using the `dotplot()` method. 

```{r eval=FALSE}
#| label: compare-dotplot
dotplot(ck)
dotplot(formula_res)
```


```{r}
#| label: fig-ccn  
#| echo: false
#| fig-keep: last
#| dev: 'png'
#| fig-width: 20
#| fig-height: 14
#| out-width: '100%'
#| fig-scap: |
#|   Comparing enrichment results of multiple gene lists.
#| fig-cap: |
#|   **Comparing enrichment results of multiple gene lists.**
#|   (A) Using a named list of gene clusters, the results were 
#|   displayed as multiple columns with each one represents 
#|   an enrichment result of a gene cluster. 
#|   (B) Using formula interface, the columns represent gene clusters defined by the formula.  

library(ggplot2)
library(aplot)

p1 <- dotplot(ck) + 
    scale_y_discrete(labels=function(x) yulab.utils::str_wrap(x, width=50)) + 
    scale_size(range = c(3, 10)) + 
    theme_dose(14)

p2 <- dotplot(formula_res) + 
    scale_y_discrete(labels=function(x) yulab.utils::str_wrap(x, width=30)) + 
    scale_size(range = c(5, 15)) + 
    theme_dose(14) +
    theme(axis.text.x=element_text(angle=30, hjust=1))

plot_list(p1, p2, ncol=2, tag_levels='A')
```

The formula interface allows more complicated gene cluster definition. In @fig-ccn(B), the gene clusters were defined by two variables (i.e. `group` that divides genes into `upregulated` and `downregulated` and `othergroup` that divides the genes into two categories of `A` and `B`.). The `dotplot()` function allows us to use one variable to divide the result into different facet and plot the result with other variables in each facet panel (@fig-ccf).



```{r}
#| label: fig-ccf  
#| fig-keep: last
#| crop: true
#| dev: 'png'
#| fig-width: 12
#| fig-height: 7
#| out-width: '90%'
#| fig-scap: |
#|   Comparing enrichment results of multiple gene lists defined by multiple variable.
#| fig-cap: |
#|   **Comparing enrichment results of multiple gene lists defined by multiple variable.** 
#|   The results were visualized as a dot plot with an x-axis representing one level of conditions
#|   (the *group* variable) and a facet panel indicating another level of conditions (the *othergroup* variable).
dotplot(formula_res, x="group") + facet_grid(~othergroup)
```


By default, only the top 5 (most significant) categories of each cluster are
shown. Users can change `showCategory` to specify how many categories should be
displayed for each cluster, and if `showCategory` is set to `NULL`, the whole
result will be plotted. The `showCategory` parameter also accepts a character
vector of selected categories to [plot pathways of
interest](#showing-specific-pathways).

`dotplot()` tries to make the comparison among different clusters more
informative and reasonable. After extracting, for example, 10 categories for
each cluster, `r Biocpkg("clusterProfiler")` collects overlap of these
categories among clusters. For example, `term A` may be enriched in all gene
clusters (*e.g.*, `g1` and `g2`) and ranked among the 10 most significant
categories of `g1` but not of `g2`. In this case,
`r Biocpkg("clusterProfiler")` will still include `term A` in the `g2` cluster
to make the comparison more reasonable.

This feature ensures that if a term is significant and selected for one cluster,
its statistics in other clusters are also displayed, allowing for a valid
comparison. Consequently, the number of categories shown for some clusters might
exceed the `showCategory` limit (e.g., 15 categories shown for *g2* while
`showCategory` is 10). Disabling this by setting `includeAll = FALSE` in
`dotplot()` may result in a plot that suggests zero overlap between clusters,
which can be misleading (see @fig-includeAll A vs B) and is not recommended.

The same selection logic is used by `cnetplot()` for `compareCluster()`
results. In both functions, a numeric `showCategory` selects the top terms
within each cluster by default, while `includeAll` controls whether matched
terms from other clusters should be retained for comparison.

```{r}
#| label: fig-includeAll
#| fig-width: 14
#| fig-height: 10
#| fig-cap: |
#|   **Comparison of dotplot with and without includeAll parameter.** 
#|   (A) `includeAll=FALSE` produces a plot that strictly follows `showCategory` but may hide overlapping terms. 
#|   (B) The default behavior (`includeAll=TRUE`) recovers the missing overlap information, making the comparison more reasonable.
p1 <- dotplot(ck, showCategory=5, includeAll=FALSE) + ggtitle("includeAll=FALSE")
p2 <- dotplot(ck, showCategory=5) + ggtitle("default")
plot_list(p1, p2, ncol=2, tag_levels='A')
``` 

The `dotplot()` function accepts a parameter `size` for setting the scale of dot sizes. The default parameter `size` is set to `geneRatio`, which corresponds to the `GeneRatio` column of the output. If it is set to `count`, the comparison will be based on gene counts, while if set to `rowPercentage`, the dot sizes will be normalized by `count/(sum of each row)`. Users can also map the dot size to other variables or derived variables (see [Chapter 16](#clusterProfiler-dplyr)).

To provide the full information, we also provide number of identified genes in each category (numbers in parentheses) when `by` is set to `rowPercentage` and number of gene clusters in each cluster label (numbers in parentheses) when `by` is set to `geneRatio`, as shown in @fig-ccn. 

The p-values indicate which categories are more likely to have biological meanings. The dots in the plot are color-coded based on their corresponding adjusted p-values. Color gradient ranging from red to blue corresponds to the order of increasing adjusted p-values. That is, red indicates low p-values (high enrichment), and blue indicates high p-values (low enrichment). Adjusted p-values were filtered out by the threshold given by the parameter `pvalueCutoff`, and FDR can be estimated by `qvalue`.

### Gene-Concept Network {#compare-cnetplot}

The [cnetplot](#cnetplot) also works with `compareCluster()` result. For
`compareClusterResult` objects, enriched terms are drawn as pies so that the
composition of each term across gene clusters can be displayed directly in the
network. Like `dotplot()`, a numeric `showCategory` selects the top enriched
terms within each cluster by default, and `includeAll = TRUE` keeps comparable
terms from other clusters in the plot.

```{r}
#| label: fig-mcnetplot  
#| fig-width: 16
#| fig-height: 16
#| fig-scap: |
#|   `cnetplot()` for comparing functional profiles of multiple gene clusters.
#| fig-cap: |
#|   **`cnetplot()` for comparing functional profiles of multiple gene clusters.** 
#|   Default per-cluster selection (A), `includeAll = FALSE` (B),
#|   `categorySizeBy = ~ -log10(p.adjust)` (C), and custom pie colors (D).
cluster_cols <- setNames(
    ## `length.out` keeps the palette the same length as the number of clusters:
    ## indexing a two-colour vector by the cluster count would leave every
    ## cluster past the second one without a colour (`NA`), which shows up as an
    ## empty legend entry and an unpainted pie.
    rep(c("#0072B2", "#D55E00"), length.out = length(unique(as.data.frame(ck)$Cluster))),
    unique(as.data.frame(ck)$Cluster)
)

p1 <- cnetplot(ck, showCategory = 3)
p2 <- cnetplot(ck, showCategory = 3, includeAll = FALSE)
p3 <- cnetplot(ck, showCategory = 3, categorySizeBy = ~ -log10(p.adjust))
p4 <- cnetplot(ck, showCategory = 3, categorySizeBy = 2) +
    scale_fill_manual(values = cluster_cols)

plot_list(p1, p2, p3, p4, ncol = 2, tag_levels = "A")
```

Compared with ordinary enrichment results, there are several useful parameters
to keep in mind for `compareCluster()` output:

- `showCategory = 3` means that the top 3 enriched terms are selected within each cluster by default.
- `showCategory` can also be a character vector if you want to display a manually selected set of terms.
- `includeAll = TRUE` keeps comparable terms that also appear in other clusters; set `includeAll = FALSE` if you only want the directly selected terms.
- `categorySizeBy` controls the size of category pies and supports expressions such as `~itemNum`, `~p.adjust`, or `~ -log10(p.adjust)`.
- Pie colors are mapped to `Cluster`, so custom cluster colors can be supplied with `ggplot2::scale_fill_manual()`.

Because the default cluster colors are taken from the set of clusters actually
present in the plot, the scale can **shift when a cluster has no enriched terms**
(i.e. when it disappears from the network). Supplying an explicit fixed palette
with `scale_fill_manual(values = ...)` (as in @fig-mcnetplot) keeps the same
cluster consistently colored across plots, which makes interpretation easier when
comparing multiple networks or when some clusters lack enriched terms.

Data attributes carried by `compareCluster()` that are shared with ordinary
enrichment results can also be mapped to node / edge aesthetics — `size_category`
/ `size_item` / `size_edge` control the relative node and edge sizes, and
`color_edge = "category"` colors edges by their term. (Note that `foldChange`
coloring of gene nodes, demonstrated in the [cnetplot section](#cnetplot), is
designed for ordinary `enrichResult` / `gseaResult` objects rather than for the
pie-composed `compareCluster` network.)

As in `dotplot()`, the `split` parameter can be used when term selection should
be carried out within another grouping variable rather than only by cluster. For
example, when GO enrichment is performed with `ont = "ALL"`, users can select
the top terms within each ontology:

```{r}
#| label: fig-mcnetplot-split
#| fig-width: 12
#| fig-height: 8
#| eval: false
ck_all <- compareCluster(
    geneCluster = gcSample,
    fun = enrichGO,
    OrgDb = org.Hs.eg.db,
    ont = "ALL"
)
ck_all <- setReadable(ck_all, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

cnetplot(ck_all, showCategory = 3, split = "ONTOLOGY")
```
## Summary


The comparison function was designed as a framework for comparing gene clusters of any kind of ontology associations, not only [groupGO](#go-classification), [enrichGO](#clusterprofiler-go-ora), [enrichKEGG](#clusterprofiler-kegg-pathway-ora), [enrichMKEGG](#clusterprofiler-kegg-module-ora), [enrichWP](#clusterprofiler-wikipathway-ora) and [enricher](#universal-api) that were provided in this package, but also other biological and biomedical ontologies, including but not limited to [enrichPathway](#reactomepa-ora), [enrichDO](#dose-do-ora), [enrichNCG](#dose-ncg-ora) and [enrichMeSH](#meshes-ora). 

In [@yu2012], we analyzed the publicly available expression dataset of breast tumor tissues from 200 patients (GSE11121, Gene Expression Omnibus) [@schmidt2008]. We identified 8 gene clusters from differentially expressed genes, and used the `compareCluster()` function to compare these gene clusters by their enriched biological process. In [@wu_clusterprofiler_2021], we analyzed the GSE8057 dataset which contains expression data from ovarian cancer cells at multiple time points and under two treatment conditions. Eight groups of DEG lists were analyzed simultaneously using `compareCluster()` with WikiPathways. The result indicates that the two drugs have distinct effects at the beginning but consistent effects in the later stages (Fig. 4 of [@wu_clusterprofiler_2021]).


## References {-}
