# GO semantic similarity analysis {#GOSemSim}



## Supported organisms

For IC-based methods, information of GO term is species specific. We
need to calculate `IC` for all GO terms of a species before we measure
semantic
similarity. [GOSemSim](https://www.bioconductor.org/packages/GOSemSim)
support all organisms that have an `OrgDb` object available.

Bioconductor have already provided `OrgDb` for about 20 species, see <http://bioconductor.org/packages/release/BiocViews.html#___OrgDb>.

We can build query `OrgDb` online via [AnnotationHub](https://www.bioconductor.org/packages/AnnotationHub). For example:

```{r eval=FALSE}
library(AnnotationHub)
hub <- AnnotationHub()
q <- query(hub, "Cricetulus")
id <- q$ah_id[length(q)]
Cgriseus <- hub[[id]]
```


If organism is not supported by [AnnotationHub](https://www.bioconductor.org/packages/AnnotationHub), user can use [AnnotationForge](https://www.bioconductor.org/packages/AnnotationForge) to build `OrgDb`.

Once we have `OrgDb`, we can build annotation data needed by [GOSemSim](https://www.bioconductor.org/packages/GOSemSim) via `godata` function.

```{r}
library(GOSemSim)
hsGO <- godata('org.Hs.eg.db', ont="MF")
```

User can set `computeIC=FALSE` if they only want to use Wang's method.


## goSim and mgoSim function

In [GOSemSim](https://www.bioconductor.org/packages/GOSemSim), we implemented all these IC-based and graph-based methods. __*goSim*__ function calculates semantic similarity between two GO terms, while __*mgoSim*__ function calculates semantic similarity between two sets of GO terms.


```{r}
goSim("GO:0004022", "GO:0005515", semData=hsGO, measure="Jiang")
goSim("GO:0004022", "GO:0005515", semData=hsGO, measure="Wang")
go1 = c("GO:0004022","GO:0004024","GO:0004174")
go2 = c("GO:0009055","GO:0005515")
mgoSim(go1, go2, semData=hsGO, measure="Wang", combine=NULL)
mgoSim(go1, go2, semData=hsGO, measure="Wang", combine="BMA")
```

## Gene Semantic Similarity Measurement

On the basis of semantic similarity between GO terms, [GOSemSim](https://www.bioconductor.org/packages/GOSemSim) can
also compute semantic similarity among sets of GO terms, gene products, and gene clusters.

Suppose we have gene $g_1$ annotated by GO terms sets $GO_{1}=\{go_{11},go_{12} \cdots go_{1m}\}$
and $g_2$ annotated by $GO_{2}=\{go_{21},go_{22} \cdots go_{2n}\}$, [GOSemSim](https://www.bioconductor.org/packages/GOSemSim) implemented four methods which called __*max*__, __*avg*__, __*rcmax*__, and __*BMA*__ to combine
semantic similarity scores of multiple GO terms. The similarities
among gene products and gene clusters which annotated by multiple GO
terms were also calculated by the same combine methods mentioned
above.



## geneSim and mgeneSim

In [GOSemSim](https://www.bioconductor.org/packages/GOSemSim), we implemented __*geneSim*__ to calculate semantic similarity between two gene products, and __*mgeneSim*__ to calculate semantic similarity among multiple gene products.

```{r}
geneSim("241", "251", semData=hsGO, measure="Wang", combine="BMA")
mgeneSim(genes=c("835", "5261","241", "994"),
         semData=hsGO, measure="Wang",verbose=FALSE)
mgeneSim(genes=c("835", "5261","241", "994"),
       semData=hsGO, measure="Rel",verbose=FALSE)
```

By default, `godata` function use `ENTREZID` as keytype, and the input ID type is `ENTREZID`. User can use other ID types such as `ENSEMBL`, `UNIPROT`, `REFSEQ`, `ACCNUM`, `SYMBOL` _et al_.

Here as an example, we use `SYMBOL` as `keytype` and calculate semantic similarities among several genes by using their gene symbol as input.

```{r}
hsGO2 <- godata('org.Hs.eg.db', keytype = "SYMBOL", ont="MF", computeIC=FALSE)
genes <- c("CDC45", "MCM10", "CDC20", "NMU", "MMP1")
mgeneSim(genes, semData=hsGO2, measure="Wang", combine="BMA", verbose=FALSE)
```

Users can also use [`clusterProfiler::bitr`](https://bioconductor.org/packages/release/bioc/vignettes/clusterProfiler/inst/doc/clusterProfiler.html#bitr-biological-id-translator) to translate biological IDs.


## clusterSim and mclusterSim

We also implemented __*clusterSim*__ for calculating semantic similarity between two gene clusters
and __*mclusterSim*__ for calculating semantic similarities among multiple gene clusters.

```{r}
gs1 <- c("835", "5261","241", "994", "514", "533")
gs2 <- c("578","582", "400", "409", "411")
clusterSim(gs1, gs2, semData=hsGO, measure="Wang", combine="BMA")

library(org.Hs.eg.db)
x <- org.Hs.egGO
hsEG <- mappedkeys(x)
set.seed <- 123
clusters <- list(a=sample(hsEG, 20), b=sample(hsEG, 20), c=sample(hsEG, 20))
mclusterSim(clusters, semData=hsGO, measure="Wang", combine="BMA")
```


<!--


## Applications

[GOSemSim](https://www.bioconductor.org/packages/GOSemSim) was cited by more than [200 papers](https://scholar.google.com.hk/scholar?oi=bibs&hl=en&cites=9484177541993722322,17633835198940746971,18126401808149291947) and had been applied to many research domains, including:

+ [Disease or Drug analysis](https://guangchuangyu.github.io/software/GOSemSim/featuredArticles/#diease-or-drug-analysis)
+ [Gene/Protein functional analysis](https://guangchuangyu.github.io/software/GOSemSim/featuredArticles/#geneprotein-functional-analysis)
+ [Protein-Protein interaction](https://guangchuangyu.github.io/software/GOSemSim/featuredArticles/#protein-protein-interaction)
+ [miRNA-mRNA interaction](https://guangchuangyu.github.io/software/GOSemSim/featuredArticles/#mirna-mrna-interaction)
+ [sRNA regulation](https://guangchuangyu.github.io/software/GOSemSim/featuredArticles/#srna-regulation)
+ [Evolution](https://guangchuangyu.github.io/software/GOSemSim/featuredArticles/#evolution)

Find out more on <https://guangchuangyu.github.io/software/GOSemSim/featuredArticles/>.


# GO enrichment analysis

GO enrichment analysis can be supported by our package [clusterProfiler](https://www.bioconductor.org/packages/clusterProfiler)[@yu2012], which supports hypergeometric test and Gene Set Enrichment Analysis (GSEA). Enrichment results across different gene clusters can be compared using __*compareCluster*__ function.

# Disease Ontology Semantic and Enrichment analysis

Disease Ontology (DO) annotates human genes in the context of disease. DO is an important annotation in translating molecular findings from high-throughput data to clinical relevance.
[DOSE](https://www.bioconductor.org/packages/DOSE)[@yu_dose_2015] supports semantic similarity computation among DO terms and genes.
Enrichment analysis including hypergeometric model and GSEA are also implemented to support discovering disease associations of high-throughput biological data.

# MeSH enrichment and semantic analyses

MeSH (Medical Subject Headings) is the NLM controlled vocabulary used to manually index articles for MEDLINE/PubMed. [meshes](https://www.bioconductor.org/packages/meshes) supports enrichment (hypergeometric test and GSEA) and semantic similarity analyses for more than 70 species.


-->