# GO enrichment analysis {#clusterprofiler-go}


GO comprises of three orthogonal ontologies, i.e. molecular function (MF), biological
process (BP), and cellular component (CC).




```{r include=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, cache=TRUE)
library(clusterProfiler)
```


## Supported organisms {#clusterProfiler-go-supported-organisms}

GO analyses (`groupGO()`, `enrichGO()` and `gseGO()`) support organisms that have an `OrgDb` object available (see also [session 2.2](#gosemsim-supported-organisms)).


If user have GO annotation data (in data.frame format with first column of gene ID and second column of GO ID), they can use `enricher()` and `gseGO()` functions to perform over-representation test and gene set enrichment analysis.

If genes are annotated by direction annotation, it should also annotated by its ancestor GO nodes (indirect annation). If user only has direct annotation, they can pass their annotation to `buildGOmap` function, which will infer indirection annotation and generate a `data.frame` that suitable for both `enricher()` and `gseGO()`.

## GO classification

In `r Biocpkg("clusterProfiler")`, the `groupGO()` is designed for gene classification based on GO distribution at a specific level. Here we use the dataset `geneList` provided by `r Biocpkg("DOSE")`. 

```{r clusterprofiler-groupgo, warning=FALSE}
library(clusterProfiler)
data(geneList, package="DOSE")
gene <- names(geneList)[abs(geneList) > 2]

# Entrez gene ID
head(gene)

ggo <- groupGO(gene     = gene,
               OrgDb    = org.Hs.eg.db,
               ont      = "CC",
               level    = 3,
               readable = TRUE)

head(ggo)
```

The `gene` parameter is a vector of gene IDs (can be any ID type that supported by corresponding `OrgDb`, see also [session 16.1](#id-convert)). If `readable` is set to `TRUE`, the input gene IDs will be converted to gene symbols.


## GO over-representation test {#clusterprofiler-go-ora}


The `r Biocpkg("clusterProfiler")` package implemented `enrichGO()` function for gene ontology over-representation test.


```{r}
ego <- enrichGO(gene          = gene,
                universe      = names(geneList),
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
		readable      = TRUE)
head(ego)
```

Any gene ID type that supported in `OrgDb` can be directly used in GO analyses. User need to specify the `keyType` parameter to specify the input gene ID type.

```{r}
gene.df <- bitr(gene, fromType = "ENTREZID",
	   	toType = c("ENSEMBL", "SYMBOL"),
		OrgDb = org.Hs.eg.db)

ego2 <- enrichGO(gene         = gene.df$ENSEMBL,
                OrgDb         = org.Hs.eg.db,
                keyType       = 'ENSEMBL',
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05)
head(ego2, 3)                
```

Gene ID can be mapped to gene Symbol by using paramter `readable=TRUE` or [`setReadable()` function](#setReadable).


## GO Gene Set Enrichment Analysis

The `r Biocpkg("clusterProfiler")` package provides `gseGO()` function for [gene set enrichment analysis](#gsea-algorithm) using gene ontology.

```{r eval=FALSE}
ego3 <- gseGO(geneList     = geneList,
              OrgDb        = org.Hs.eg.db,
              ont          = "CC",
              minGSSize    = 100,
              maxGSSize    = 500,
              pvalueCutoff = 0.05,
              verbose      = FALSE)
```

The format of input data, `geneList`, was documented in the [FAQ](#genelist). Beware that only gene Set size in `[minGSSize, maxGSSize]` will be tested.



## GO analysis for non-model organisms {#clusterprofiler-go-non-model}

Both `enrichGO()` and `gseGO()` functions require an `OrgDb` object as background annotation. For organisms that don't have `OrgDb` provided by Bioconductor, users can query one (if available) online via `r Biocpkg("AnnotationHub")`. If there is no `OrgDb` available, users can obtain GO annotation from other sources, e.g. from `r Biocpkg("biomaRt")` or [Blast2GO](https://www.blast2go.com/). Then the `enricher()` or `GSEA()` functions can be used to perform GO analysis for these organisms, similar to the examples using wikiPathways and MSigDB. Another solution is to create `OrgDb` by your own using `r Biocpkg("AnnotationForge")` package.


## Summary {#clusterprofiler-go-summary}

GO semantic similarity can be calculated by `r Biocpkg("GOSemSim")` [@yu2010]. We can use it to cluster genes/proteins into different clusters based on their functional similarity and can also use it to measure the similarities among GO terms to reduce the redundancy of GO enrichment results.
