# DO enrichment analysis {#dose-enrichment}


```{r include=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, cache=TRUE)
library(clusterProfiler)
```


[DOSE](https://www.bioconductor.org/packages/DOSE)[@yu_dose_2015] supports Disease Ontology (DO) Semantic and Enrichment analysis. The `enrichDO` function is very useful for identifying disease association of interesting genes, and function `gseDO` function is designed for gene set enrichment analysis of _DO_.

In addition, [DOSE](https://www.bioconductor.org/packages/DOSE) also supports enrichment analysis of [Network of Cancer Gene](http://ncg.kcl.ac.uk/) (NCG)[@omer_ncg] and [Disease Gene Network](http://disgenet.org/)[@janet_disgenet], please refer to the [DOSE](https://www.bioconductor.org/packages/DOSE) vignettes.



## `enrichDO` function

In the following example, we selected fold change above 1.5 as the differential genes and analyzing their disease association.

```{r}
library(DOSE)
data(geneList)
gene <- names(geneList)[abs(geneList) > 1.5]
head(gene)
x <- enrichDO(gene          = gene,
              ont           = "DO",
              pvalueCutoff  = 0.05,
              pAdjustMethod = "BH",
              universe      = names(geneList),
              minGSSize     = 5,
              maxGSSize     = 500,
              qvalueCutoff  = 0.05,
              readable      = FALSE)
head(x)
```

The `enrichDO` function requires an entrezgene ID vector as input, mostly is the differential gene list of gene expression profile studies. If user needs to convert other gene ID type to entrezgene ID, we recommend using `bitr` function provided by `r Biocpkg("clusterProfiler")`.

The `ont` parameter can be "DO" or "DOLite", DOLite[@Du15062009] was constructed to aggregate the redundant DO terms. The DOLite data is not updated, we recommend user use `ont="DO"`. `pvalueCutoff` setting the cutoff value of *p* value and *p* value adjust; `pAdjustMethod` setting the *p* value correction methods, include the Bonferroni correction ("bonferroni"), Holm ("holm"), Hochberg ("hochberg"), Hommel ("hommel"), Benjamini \& Hochberg ("BH") and Benjamini \& Yekutieli ("BY") while `qvalueCutoff` is used to control *q*-values.

The `universe` setting the background gene universe for testing. If user do not explicitly setting this parameter, `enrichDO` will set the universe to all human genes that have DO annotation.

The `minGSSize` (and `maxGSSize`) indicates that only those DO terms that have more than `minGSSize` (and less than `maxGSSize`) genes annotated will be tested.

The `readable` is a logical parameter, indicates whether the entrezgene IDs will mapping to gene symbols or not.

We also implement `setReadable` function that helps the user to convert entrezgene IDs to gene symbols.

```{r}
x <- setReadable(x, 'org.Hs.eg.db')
head(x)
```

## `enrichNCG` function

[Network of Cancer Gene](http://ncg.kcl.ac.uk/) (NCG)[@omer_ncg] is a manually curated repository of cancer genes. NCG release 5.0 (Aug. 2015) collects 1,571 cancer genes from 175 published studies. `r Biocpkg("DOSE")` supports analyzing gene list and determine whether they are enriched in genes known to be mutated in a given cancer type.

```{r}
gene2 <- names(geneList)[abs(geneList) < 3]
ncg <- enrichNCG(gene2)
head(ncg)
```

## `enrichDGN` and `enrichDGNv` functions

[DisGeNET](http://disgenet.org/)[@janet_disgenet] is an integrative and comprehensive resources of gene-disease associations from several public data sources and the literature. It contains gene-disease associations and snp-gene-disease associations.

The enrichment analysis of disease-gene associations is supported by the `enrichDGN` function and analysis of snp-gene-disease associations is supported by the `enrichDGNv` function.

```{r}
dgn <- enrichDGN(gene)
head(dgn)

snp <- c("rs1401296", "rs9315050", "rs5498", "rs1524668", "rs147377392",
         "rs841", "rs909253", "rs7193343", "rs3918232", "rs3760396",
         "rs2231137", "rs10947803", "rs17222919", "rs386602276", "rs11053646",
         "rs1805192", "rs139564723", "rs2230806", "rs20417", "rs966221")
dgnv <- enrichDGNv(snp)
head(dgnv)
```


## `gseDO` fuction

In the following example, in order to speedup the compilation of this document, only gene sets with size above 120 were tested and only 100 permutations were performed.

```{r}
library(DOSE)
data(geneList)
y <- gseDO(geneList,
           nPerm         = 100,
           minGSSize     = 120,
           pvalueCutoff  = 0.2,
           pAdjustMethod = "BH",
           verbose       = FALSE)
head(y, 3)
```

## `gseNCG` fuction

```{r}
ncg <- gseNCG(geneList,
              nPerm         = 100,
              minGSSize     = 120,
              pvalueCutoff  = 0.2,
              pAdjustMethod = "BH",
              verbose       = FALSE)
ncg <- setReadable(ncg, 'org.Hs.eg.db')
head(ncg, 3)
```

## `gseDGN` fuction

```{r}
dgn <- gseDGN(geneList,
              nPerm         = 100,
              minGSSize     = 120,
              pvalueCutoff  = 0.2,
              pAdjustMethod = "BH",
              verbose       = FALSE)
dgn <- setReadable(dgn, 'org.Hs.eg.db')
head(dgn, 3)
```
