# Visualization of functional enrichment result {#enrichplot}


The `r Biocpkg("enrichplot")` package implements several visualization methods to help interpreting enrichment results. It supports visualizing enrichment results obtained from `r Biocpkg("DOSE")` [@yu_dose_2015], `r Biocpkg("clusterProfiler")` [@yu2012],
`r Biocpkg("ReactomePA")` [@yu_reactomepa_2016] and `r Biocpkg("meshes")` [@yu_meshes_2018]. Both over representation analysis (ORA) and gene set enrichment analysis (GSEA) are supported.

Many of these visualization methods were first implemented in `r Biocpkg("DOSE")` and rewrote from scratch using `r CRANpkg("ggplot2")`. If you want to use [old methods](https://www.biostars.org/p/375555), you can use the [doseplot](https://github.com/GuangchuangYu/doseplot) package.


## Bar Plot


Bar plot is the most widely used method to visualize enriched terms. It depicts the enrichment scores (*e.g.* p values) and gene count or ratio as bar height
and color.



```{r}
library(DOSE)
data(geneList)
de <- names(geneList)[abs(geneList) > 2]

edo <- enrichDGN(de)
```

(ref:Barplotscap) Bar plot of enriched terms.

(ref:Barplotcap) **Bar plot of enriched terms.** 

```{r Barplot, fig.height=5, fig.width=8, fig.cap="(ref:Barplotcap)", fig.scap="(ref:Barplotscap)"}
library(enrichplot)
barplot(edo, showCategory=20) 
```


## Dot plot


Dot plot is similar to bar plot with the capability to encode another score as dot size.

```r
edo2 <- gseDO(geneList)
dotplot(edo, showCategory=30) + ggtitle("dotplot for ORA")
dotplot(edo2, showCategory=30) + ggtitle("dotplot for GSEA")
```

(ref:Dotplotscap) Dot plot of enriched terms.

(ref:Dotplotcap) **Dot plot of enriched terms.** 

```{r Dotplotcap, fig.width=12, fig.height=8, fig.cap="(ref:Dotplotcap)", fig.scap="(ref:Dotplotscap)", echo=FALSE}
edo2 <- gseDO(geneList)
p1 <- dotplot(edo, showCategory=30) + ggtitle("dotplot for ORA")
p2 <- dotplot(edo2, showCategory=30) + ggtitle("dotplot for GSEA")
plot_grid(p1, p2, ncol=2, labels = LETTERS[1:2])
```



## Gene-Concept Network

Both the `barplot` and `dotplot` only displayed most significant enriched terms,
while users may want to know which genes are involved in these significant
terms. 
In order to consider the potentially biological complexities in which a gene may belong to multiple annotation categories and provide information of numeric changes if available, we developed `cnetplot` function to extract the complex association. 
The `cnetplot` depicts the linkages of genes and biological concepts (*e.g.* GO terms or KEGG pathways) as a network. GSEA result is also supported
with only core enriched genes displayed.

(ref:Networkplotscap) Network plot of enriched terms.

(ref:Networkplotcap) **Network plot of enriched terms.** 

```{r Networkplot, fig.width=20, fig.height=8, fig.cap="(ref:Networkplotcap)", fig.scap="(ref:Networkplotscap)"}
## convert gene ID to Symbol
edox <- setReadable(edo, 'org.Hs.eg.db', 'ENTREZID')
p1 <- cnetplot(edox, foldChange=geneList)
## categorySize can be scaled by 'pvalue' or 'geneNum'
p2 <- cnetplot(edox, categorySize="pvalue", foldChange=geneList)
p3 <- cnetplot(edox, foldChange=geneList, circular = TRUE, colorEdge = TRUE)
cowplot::plot_grid(p1, p2, p3, ncol=3, labels=LETTERS[1:3], rel_widths=c(.8, .8, 1.2))
```




If you would like label subset of the nodes, you can use the `node_label` parameter, which supports 4 possible selections (i.e. "category", "gene", "all" and "none"), as demonstrated in Figure \@ref(fig:cnetNodeLabel).


(ref:cnetNodeLabelscap) Labelling nodes by selected subset.

(ref:cnetNodeLabelcap) **Labelling nodes by selected subset.** gene category (A), gene name (B), both gene category and gene name (C, default) and not to label at all (D).

```{r cnetNodeLabel, fig.height=12, fig.width=16, fig.cap="(ref:cnetNodeLabelcap)", fig.scap="(ref:cnetNodeLabelscap)"}
p1 <- cnetplot(edox, node_label="category") 
p2 <- cnetplot(edox, node_label="gene") 
p3 <- cnetplot(edox, node_label="all") 
p4 <- cnetplot(edox, node_label="none") 
cowplot::plot_grid(p1, p2, p3, p4, ncol=2, labels=LETTERS[1:4])
```


## Heatmap-like functional classification


The `heatplot` is similar to `cnetplot`, while displaying the relationships as a
heatmap. The gene-concept network may become too complicated if user want to
show a large number significant terms. The `heatplot` can simplify the result
and more easy to identify expression patterns.

(ref:Heatplotscap) Heatmap plot of enriched terms.

(ref:Heatplotcap) **Heatmap plot of enriched terms.** default (A), `foldChange=geneList` (B)

```{r Heatplot, fig.width=16, fig.height=8, fig.cap="(ref:Heatplotcap)", fig.scap="(ref:Heatplotscap)"}
p1 <- heatplot(edox)
p2 <- heatplot(edox, foldChange=geneList)
cowplot::plot_grid(p1, p2, ncol=1, labels=LETTERS[1:2])
```



## Enrichment Map


Enrichment map organizes enriched terms into a network with edges connecting
overlapping gene sets. In this way, mutually overlapping gene sets are tend to
cluster together, making it easy to identify functional module.

The `emapplot` function supports results obtained from hypergeometric test and gene set enrichment analysis. The `node_scale` parameter can be used to resize nodes, as demonstrated in Figure \@ref(fig:Enrichment) B, and the `layout` parameter can adjust the layout, as demonstrated in Figure \@ref(fig:Enrichment) C and D.

(ref:Enrichmentscap) Plot for results obtained from hypergeometric test and gene set enrichment analysis.

(ref:Enrichmentcap) **Plot for results obtained from hypergeometric test and gene set enrichment analysis.** default (A), `node_scale=1.5` (B), `layout="kk"` (C) and `node_scale=1.5,layout="kk"` (D).

```{r Enrichment, fig.height=14, fig.width=16, fig.cap="(ref:Enrichmentcap)", fig.scap="(ref:Enrichmentscap)"}
p1 <- emapplot(edo)
p2 <- emapplot(edo, node_scale=1.5)
p3 <- emapplot(edo, layout="kk")
p4 <- emapplot(edo, node_scale=1.5,layout="kk") 
cowplot::plot_grid(p1, p2, p3, p4, ncol=2, labels=LETTERS[1:4])
```


## Biological theme comparison 

The `emapplot` function also supports results obtained from `compareCluster` function of `clusterProfiler` package. In addition to `node_scale` and `layout` parameters, the number of circles in the bottom left corner can be adjusted  using the `legend_n` parameteras, as demonstrated in Figure \@ref(fig:Enrichment2) B. And proportion of clusters in the pie chart can be adjusted using the `pie` parameter, when `pie="count"`, the proportion of  clusters in the pie chart is determined by the number of genes, as demonstrated in Figure \@ref(fig:Enrichment2) C and D.


(ref:Enrichment2scap) Plot for results obtained from `compareCluster` function of `clusterProfiler` package.

(ref:Enrichment2cap) **Plot for results obtained from `compareCluster` function of `clusterProfiler` package.** default (A), `legend_n=2` (B), `pie="count"` (C) and `pie="count", node_scale=1.5, layout="kk"` (D).

```{r Enrichment2, fig.height=18, fig.width=16, fig.cap="(ref:Enrichment2cap)", fig.scap="(ref:Enrichment2scap)"}
library(clusterProfiler)
data(gcSample)
xx <- compareCluster(gcSample, fun="enrichKEGG",
                     organism="hsa", pvalueCutoff=0.05)
p1 <- emapplot(xx)
p2 <- emapplot(xx, legend_n=2) 
p3 <- emapplot(xx, pie="count")
p4 <- emapplot(xx, pie="count", node_scale=1.5, layout="kk")
cowplot::plot_grid(p1, p2, p3, p4, ncol=2, labels=LETTERS[1:4])
```


## UpSet Plot


The `upsetplot` is an alternative to `cnetplot` for visualizing the complex
association between genes and gene sets. It emphasizes the gene overlapping
among different gene sets.



(ref:upsetORAscap) Upsetplot for over-representation analysis.

(ref:upsetORAcap) **Upsetplot for over-representation analysis.** 

```{r upsetORA, fig.width=12, fig.height=5, fig.cap="(ref:upsetORAcap)", fig.scap="(ref:upsetORAscap)"}
upsetplot(edo)
```

For over-representation analysis, `upsetplot` will calculate the overlaps among different gene sets as demonstrated in Figure \@ref(fig:upsetORA). For GSEA result, it will plot the fold change distributions of different categories (e.g. unique to pathway, overlaps among different pathways).


(ref:upsetGSEAscap) Upsetplot for gene set enrichment analysis.

(ref:upsetGSEAcap) **Upsetplot for gene set enrichment analysis.** 

```{r upsetGSEA, fig.height=5, fig.width=12, fig.cap="(ref:upsetGSEAcap)", fig.scap="(ref:upsetGSEAscap)"}
upsetplot(kk2) 
```


## ridgeline plot for expression distribution of GSEA result

The `ridgeplot` will visualize expression distributions of core enriched genes
for GSEA enriched categories. It helps users to interpret up/down-regulated pathways.

(ref:ridgeplotscap) Ridgeplot for gene set enrichment analysis.

(ref:ridgeplotcap) **Ridgeplot for gene set enrichment analysis.** 

```{r ridgeplot, fig.width=12, fig.height=8, message=FALSE, fig.cap="(ref:ridgeplotcap)", fig.scap="(ref:ridgeplotscap)"}
ridgeplot(edo2)
```


## running score and preranked list of GSEA result

Running score and preranked list are traditional methods for visualizing GSEA
result. The `r Biocpkg("enrichplot")` package supports both of them to visualize
the distribution of the gene set and the enrichment score.

(ref:gseaplotscap) gseaplot for GSEA result(`by = "runningScore"`).

(ref:gseaplotcap) **gseaplot for GSEA result(`by = "runningScore"`).** `by = "runningScore"` (A), `by = "preranked"` (B), default (C)

```{r gseaplot, fig.width=12, fig.height=16, fig.cap="(ref:gseaplotcap)", fig.scap="(ref:gseaplotscap)"}
p1 <- gseaplot(edo2, geneSetID = 1, by = "runningScore", title = edo2$Description[1])
p2 <- gseaplot(edo2, geneSetID = 1, by = "preranked", title = edo2$Description[1])
p3 <- gseaplot(edo2, geneSetID = 1, title = edo2$Description[1])
cowplot::plot_grid(p1, p2, p3, ncol=1, labels=LETTERS[1:3])
```


Another method to plot GSEA result is the `gseaplot2` function:


(ref:gseaplot2scap) Gseaplot2 for GSEA result.

(ref:gseaplot2cap) **Gseaplot2 for GSEA result.** 

```{r gseaplot2, fig.width=12, fig.height=8, fig.cap="(ref:gseaplot2cap)", fig.scap="(ref:gseaplot2scap)"}
gseaplot2(edo2, geneSetID = 1, title = edo2$Description[1])
```

The `gseaplot2` also supports multile gene sets to be displayed on the same figure:

(ref:gseaplot22scap) Gseaplot2 for GSEA result of multile gene sets.

(ref:gseaplot22cap) **Gseaplot2 for GSEA result of multile gene sets.** 

```{r gseaplot22, fig.width=12, fig.height=8, fig.cap="(ref:gseaplot22cap)", fig.scap="(ref:gseaplot22scap)"}
gseaplot2(edo2, geneSetID = 1:3)
```

User can also displaying the pvalue table on the plot via `pvalue_table`
parameter:

(ref:gseaplot23scap) Gseaplot2 for GSEA result of multile gene sets(add pvalue_table).

(ref:gseaplot23cap) **Gseaplot2 for GSEA result of multile gene sets(add pvalue_table).** 

```{r gseaplot23, fig.width=12, fig.height=8, fig.cap="(ref:gseaplot23cap)", fig.scap="(ref:gseaplot23scap)"}
gseaplot2(edo2, geneSetID = 1:3, pvalue_table = TRUE,
          color = c("#E495A5", "#86B875", "#7DB0DD"), ES_geom = "dot")
```


User can specify `subplots` to only display a subset of plots:

(ref:gseaplot24scap) Gseaplot2 for GSEA result of multile gene sets(add subplots).

(ref:gseaplot24cap) **Gseaplot2 for GSEA result of multile gene sets(add subplots).** `subplots = 1` (A),`subplots = 1:2` (B)

```{r gseaplot24, fig.width=12, fig.height=8, fig.cap="(ref:gseaplot24cap)", fig.scap="(ref:gseaplot24scap)"}
p1 <- gseaplot2(edo2, geneSetID = 1:3, subplots = 1)
p2 <- gseaplot2(edo2, geneSetID = 1:3, subplots = 1:2)
cowplot::plot_grid(p1, p2, ncol=1, labels=LETTERS[1:2])
```



The `gsearank` function plot the ranked list of genes belong to the specific
gene set.

(ref:gsearankscap) Ranked list of genes belong to the specific gene set.

(ref:gsearankcap) **Ranked list of genes belong to the specific gene set.**

```{r gsearank, fig.width=8, fig.height=4, fig.cap="(ref:gsearankcap)", fig.scap="(ref:gsearankscap)"}
gsearank(edo2, 1, title = edo2[1, "Description"])
```

Multiple gene sets can be aligned using `cowplot`:

(ref:gsearank2scap) Gsearank for multiple gene sets.

(ref:gsearank2cap) **Gsearank for multiple gene sets.**

```{r gsearank2, fig.width=8, fig.height=12, fig.cap="(ref:gsearank2cap)", fig.scap="(ref:gsearank2scap)"}
library(ggplot2)
library(cowplot)

pp <- lapply(1:3, function(i) {
    anno <- edo2[i, c("NES", "pvalue", "p.adjust")]
    lab <- paste0(names(anno), "=",  round(anno, 3), collapse="\n")

    gsearank(edo2, i, edo2[i, 2]) + xlab(NULL) +ylab(NULL) +
        annotate("text", 0, edo2[i, "enrichmentScore"] * .9, label = lab, hjust=0, vjust=0)
})
plot_grid(plotlist=pp, ncol=1)
```



## pubmed trend of enriched terms

One of the problem of enrichment analysis is to find pathways for further
investigation. Here, we provide `pmcplot` function to plot the number/proportion
of publications trend based on the query result from PubMed Central. Of course,
users can use `pmcplot` in other scenarios. All text that can be queried on PMC
is valid as input of `pmcplot`.

(ref:pmcplotscap) Pmcplot of enrichment analysis.

(ref:pmcplotcap) **Pmcplot of enrichment analysis.**

```{r pmcplot, fig.width=14, fig.height=6, fig.cap="(ref:pmcplotcap)", fig.scap="(ref:pmcplotscap)"}
terms <- edo$Description[1:5]
p <- pmcplot(terms, 2010:2020)
p2 <- pmcplot(terms, 2010:2020, proportion=FALSE)
plot_grid(p, p2, ncol=2)
```



<!--

using ggplot2 api

`[` accessor


ggplot(egobp[1:20], aes(x=reorder(Description, -pvalue), y=Count, fill=-p.adjust)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_continuous(low="blue", high="red") +
    labs(x = "", y = "", fill = "p.adjust") +
    theme(axis.text=element_text(size=11))

-->

