# (PART\*) Part III: Miscellaneous topics {-}

# dplyr verbs for manipulating enrichment result {#clusterProfiler-dplyr}


```{r include=F}
library(DOSE)
library(ggplot2)
library(forcats)
library(enrichplot)
theme_set(theme_grey())
select <- clusterProfiler:::select.enrichResult
```


```{r}
library(DOSE)
data(geneList)
de = names(geneList)[1:100]
x = enrichDO(de)
```

## filter

```{r}

filter(x, p.adjust < .05, qvalue < 0.2)
```

## arrange
```{r}
mutate(x, geneRatio = parse_ratio(GeneRatio)) %>%
  arrange(desc(geneRatio))
```

## select

```{r}
select(x, -geneID) %>% head
```

## mutate

(ref:richfactorplotscap) Visualizing rich factor of enriched terms using lolliplot.

(ref:richfactorplotcap) **Visualizing rich factor of enriched terms using lolliplot.** 


```{r dot-richfactor, fig.width=7, fig.height=7, fig.cap="(ref:richfactorplotcap)", fig.scap="(ref:richfactorplotscap)"}

# k/M
y <- mutate(x, richFactor = Count / as.numeric(sub("/\\d+", "", BgRatio)))
y

library(ggplot2)
library(forcats)
library(enrichplot)

ggplot(y, showCategory = 20, 
  aes(richFactor, fct_reorder(Description, richFactor))) + 
  geom_segment(aes(xend=0, yend = Description)) +
  geom_point(aes(color=p.adjust, size = Count)) +
  scale_color_viridis_c(guide=guide_colorbar(reverse=TRUE)) +
  scale_size_continuous(range=c(2, 10)) +
  theme_minimal() + 
  xlab("rich factor") +
  ylab(NULL) + 
  ggtitle("Enriched Disease Ontology")
```


A very similar concept is Fold Enrichment, which is defined as the ratio of two proportions, (k/n) / (M/N). Using `mutate` to add the fold enrichment variable is also easy:

```r
mutate(x, FoldEnrichment = parse_ratio(GeneRatio) / parse_ratio(BgRatio))
```

## slice

We can use `slice` to choose rows by their ordinal position in the enrichment result. Grouped result use the ordinal position with the group.

In the following example, a GSEA result of Reactome pathway was sorted by the absolute values of NES and the result was grouped by the sign of NES. We then extracted first 5 rows of each groups. The result was displayed in Figure \@ref(fig:sliceNES).


(ref:sliceNESscap) Choose pathways by ordinal positions.

(ref:sliceNEScap) **Choose pathways by ordinal positions.** 


```{r sliceNES, fig.width=9, fig.height=5, fig.cap="(ref:sliceNEScap)", fig.scap="(ref:sliceNESscap)"}
library(ReactomePA)
data(geneList)
x <- gsePathway(geneList)


y <- arrange(x, abs(NES)) %>% 
        group_by(sign(NES)) %>% 
        slice(1:5)

library(forcats)
library(ggplot2)
library(ggstance)
library(enrichplot)

ggplot(y, aes(NES, fct_reorder(Description, NES), fill=qvalues), showCategory=10) + 
    geom_barh(stat='identity') + 
    scale_fill_continuous(low='red', high='blue', guide=guide_colorbar(reverse=TRUE)) + 
    theme_minimal() + ylab(NULL)
```


## summarise

(ref:pBarplotscap) Distribution of adjusted p-values.

(ref:pBarplotcap) **Distribution of adjusted p-values.** 


```{r p-bar, fig.width=8, fig.height=6, fig.cap="(ref:pBarplotcap)", fig.scap="(ref:pBarplotscap)"}
pbar <- function(x) {
  pi=seq(0, 1, length.out=11)

  mutate(x, pp = cut(p.adjust, pi)) %>%
    group_by(pp) %>% 
    summarise(cnt = n()) %>% 
    ggplot(aes(pp, cnt)) + geom_col() + 
    theme_minimal() +
    xlab("p value intervals") +
    ylab("Frequency") + 
    ggtitle("p value distribution")
}    

x <- enrichDO(de, pvalueCutoff=1, qvalueCutoff=1)
set.seed(2020-09-10)
random_genes <- sample(names(geneList), 100)
y <- enrichDO(random_genes, pvalueCutoff=1, qvalueCutoff=1)
p1 <- pbar(x)
p2 <- pbar(y)
cowplot::plot_grid(p1, p2, ncol=1, labels = LETTERS[1:2])
```

