# bitr: Biological Id TranslatoR {#bitr}


[clusterProfiler](https://www.bioconductor.org/packages/clusterProfiler) provides `bitr` and `bitr_kegg` for converting ID types. Both `bitr` and `bitr_kegg` support many species including model and many non-model organisms.

```{r}
x <- c("GPX3",  "GLRX",   "LBP",   "CRYAB", "DEFB1", "HCLS1",   "SOD2",   "HSPA2",
       "ORM1",  "IGFBP1", "PTHLH", "GPC3",  "IGFBP3","TOB1",    "MITF",   "NDRG1",
       "NR1H4", "FGFR3",  "PVR",   "IL6",   "PTPRM", "ERBB2",   "NID2",   "LAMB1",
       "COMP",  "PLS3",   "MCAM",  "SPP1",  "LAMC1", "COL4A2",  "COL4A1", "MYOC",
       "ANXA4", "TFPI2",  "CST6",  "SLPI",  "TIMP2", "CPM",     "GGT1",   "NNMT",
       "MAL",   "EEF1A2", "HGD",   "TCN2",  "CDA",   "PCCA",    "CRYM",   "PDXK",
       "STC1",  "WARS",  "HMOX1", "FXYD2", "RBP4",   "SLC6A12", "KDELR3", "ITM2B")
eg = bitr(x, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
head(eg)
```

User should provides an annotation package, both _fromType_ and _toType_ can accept any types that supported.

User can use _keytypes_ to list all supporting types.

```{r}
library(org.Hs.eg.db)
keytypes(org.Hs.eg.db)
```

We can translate from one type to other types.
```{r}
ids <- bitr(x, fromType="SYMBOL", toType=c("UNIPROT", "ENSEMBL"), OrgDb="org.Hs.eg.db")
head(ids)
```

For GO analysis, user don't need to convert ID, all ID type provided by `OrgDb` can be used in `groupGO`, `enrichGO` and `gseGO` by specifying `keyType` parameter.

### `bitr_kegg`: converting biological IDs using KEGG API {#bitr_kegg}


```{r}
data(gcSample)
hg <- gcSample[[1]]
head(hg)

eg2np <- bitr_kegg(hg, fromType='kegg', toType='ncbi-proteinid', organism='hsa')
head(eg2np)
```

The ID type (both `fromType` & `toType`) should be one of 'kegg', 'ncbi-geneid', 'ncbi-proteinid' or 'uniprot'. The 'kegg' is the primary ID used in KEGG database. The data source of KEGG was from NCBI. A rule of thumb for the 'kegg' ID is `entrezgene` ID for eukaryote species and `Locus` ID for prokaryotes.

Many prokaryote species don't have entrezgene ID available. For example we can check the gene information of `ece:Z5100` in <http://www.genome.jp/dbget-bin/www_bget?ece:Z5100>, which have `NCBI-ProteinID` and `UnitProt` links in the `Other DBs` Entry, but not `NCBI-GeneID`.


If we try to convert `Z5100` to `ncbi-geneid`, `bitr_kegg` will throw error of `ncbi-geneid is not supported`.

```{r eval=FALSE}
bitr_kegg("Z5100", fromType="kegg", toType='ncbi-geneid', organism='ece')
```

```
## Error in KEGG_convert(fromType, toType, organism) :
## ncbi-geneid is not supported for ece ...
```

We can of course convert it to `ncbi-proteinid` and `uniprot`:

```{r}
bitr_kegg("Z5100", fromType="kegg", toType='ncbi-proteinid', organism='ece')
bitr_kegg("Z5100", fromType="kegg", toType='uniprot', organism='ece')
```

# Useful utilities {#other-utilities}



## `setReadable`: translating gene IDs to human readable symbols {#setReadable}

Some of the functions, especially those internally supported for [DO](#chapter4), [GO](#chapter5), and [Reactome Pathway](#chapter8), support a parameter, `readable`.  If `readable = TRUE`, all the gene IDs will be translated to gene symbols. The `readable` parameter is not available for enrichment analysis of KEGG or using user's own annotation. KEGG analysis using `enrichKEGG` and `gseKEGG`, internally query annotation information from KEEGG database and thus support all species if it is available in the KEGG database. However, KEGG database doesn't provide gene ID to symbol mapping information. For analysis using user's own annotation data, we even don't know what species is in analyzed. Translating gene IDs to gene symbols is partly supported using the `setReadable` function if and only if there is an `OrgDb` available. 


```{r}
library(org.Hs.eg.db)
library(clusterProfiler)

data(geneList, package="DOSE")
de <- names(geneList)[1:100]
x <- enrichKEGG(de)
## The geneID column is ENTREZID
head(x, 3)

y <- setReadable(x, OrgDb = org.Hs.eg.db, keyType="ENTREZID")
## The geneID column is translated to symbol
head(y, 3)
```

For those functions that internally support `readable` parameter, user can also use `setReadable` for translating gene IDs.


