# Comparing biological themes from multiple gene lists {#clusterprofiler-comparecluster}

This chapter answers the comparison question: which biological themes differ among gene lists, clusters, or experimental groups? `compareCluster()` reruns the same enrichment function for each group and returns one `compareClusterResult`, so the groups share an explicit analysis design and can be compared without treating independently generated p-values as if they were a common table.

```{r include=FALSE}
#| message: false
#| cache: false
source("_common.R")
library(clusterProfiler)
library(org.Hs.eg.db)
library(DOSE)
```

## Chapter overview

| Aspect | Comparison analysis |
|---|---|
| **Input** | A named list of gene vectors, or one gene-per-row data frame with formula grouping columns. |
| **Engine** | One shared enrichment function (`enrichGO()`, `enrichKEGG()`, `GSEA()`, or another supported function) applied per group. |
| **Output** | A `compareClusterResult` retaining group labels, terms, gene mappings, and per-group statistics. |
| **Key limitation** | Groups must be compared under an explicit identifier namespace, universe/background, annotation source, and multiple-testing design. |

The `r Biocpkg("clusterProfiler")` package was developed for biological theme comparison [@yu2012; @wu_clusterprofiler_2021], and it provides a function, `compareCluster`, to automatically calculate enriched functional profiles of each gene clusters and aggregate the results into a single object. Comparing functional profiles can reveal functional consensus and differences among different experiments and helps in identifying differential functional modules in omics datasets.

## Comparing multiple gene lists

The `compareCluster()` function applies selected function (via the `fun` parameter) to perform enrichment analysis for each gene list. 

```{r}
#| label: compare-multiple-gene-lists
data(gcSample)
str(gcSample) 
```

Users can use a named list of gene IDs as the input that passed to the `geneCluster` parameter. 

```{r}
#| label: compare-multiple-gene-lists-enrichGO
ck <- compareCluster(geneCluster = gcSample, fun = enrichGO, 
                    OrgDb = org.Hs.eg.db, ont="BP")
ck <- setReadable(ck, OrgDb = org.Hs.eg.db, keyType="ENTREZID")
head(ck) 
```


## Formula interface of compareCluster

As an alternative to using named list, the `compareCluster()` function also supports passing a formula to describe more complicated experimental designs (*e.g.*, $Gene \sim time + treatment$).


```{r}
#| label: compare-formula
data(geneList, package='DOSE')
mydf <- data.frame(Entrez=names(geneList), FC=geneList)
mydf <- mydf[abs(mydf$FC) > 1,]
mydf$group <- "upregulated"
mydf$group[mydf$FC < 0] <- "downregulated"
mydf$othergroup <- "A"
mydf$othergroup[abs(mydf$FC) > 2] <- "B"

formula_res <- compareCluster(Entrez~group+othergroup, data=mydf, 
                            fun=enrichGO, OrgDb = org.Hs.eg.db, ont="BP")
  
head(formula_res)
```

::: {.callout-note}
**`compareCluster()` takes gene lists, not enrichment tables.** This is the
single most common misunderstanding about the formula interface. Both the named
list and the formula form expect **genes**, and `compareCluster()` runs the
enrichment test itself, once per group:

- `data` must be a data frame with **one row per gene** — a gene identifier
  column, plus the grouping columns on the right-hand side of the formula.
- The statistics in the result are computed by `fun` (`enrichGO()` here) from
  each group's gene list. They are **not** carried over from any input table.

So if you already have two tables of enrichment results (`ID`, `Description`,
`GeneRatio`, `pvalue`, `p.adjust`, ...), there is nothing to join: those numbers
were produced by whatever run created them, on whatever background, and
`compareCluster()` has no way to reuse them. To compare the two conditions,
combine the **gene lists** that produced those tables and let
`compareCluster()` re-run the test:

```{r eval=FALSE}
## one row per gene, with a column naming the group it belongs to
genes <- rbind(
    data.frame(gene = condition_A_genes, group = "A"),
    data.frame(gene = condition_B_genes, group = "B")
)
res <- compareCluster(gene ~ group, data = genes, fun = "enrichGO",
                      OrgDb = org.Hs.eg.db, ont = "BP")
```

Re-running the test is not a limitation to work around — it is what makes the
two columns comparable. Enrichment p-values depend on the gene list, the
universe/background and the multiple-testing correction, so p-values computed
separately for A and B are not on a common footing and plotting them side by
side would be misleading.

If you want to carry a *pre-computed* gene-level statistic into the comparison
instead (a ranked list rather than a cut-off list), use the GSEA route:
`compareCluster(..., fun = "gseGO")`.
:::

## Functional analysis of single-cell marker genes

The `compareCluster()` function is highly versatile and can be directly integrated into single-cell analysis workflows. For example, after identifying marker genes for each cluster using `Seurat`, the results can be directly used for functional enrichment analysis.

### Using Seurat results

Typically, `FindAllMarkers()` from `Seurat` returns a data frame where rows are genes and columns include cluster information and statistical metrics.

This example is wrapped in `retry()` (defined in `_common.R`) because its `fun` reaches the remote KEGG service, so a transient failure is retried a few times before giving up; the `enrichGO` examples further on stay unwrapped because they run locally.

```{r eval=FALSE}
#| label: sc-seurat-example
library(Seurat)
library(SeuratData)
library(dplyr)
library(clusterProfiler)

# Load example data
data("pbmc3k")
sce <- pbmc3k.final

# Identify marker genes
sce.markers <- FindAllMarkers(object = sce, only.pos = TRUE, 
                              min.pct = 0.25, 
                              thresh.use = 0.25)

# Filter markers
markers <- sce.markers |> 
    group_by(cluster) |> 
    filter(p_val_adj < 0.001) |> 
    ungroup()

# ID conversion (if needed)
gid <- bitr(unique(markers$gene), 'SYMBOL', 'ENTREZID', OrgDb = 'org.Hs.eg.db')
markers <- full_join(markers, gid, by = c('gene' = 'SYMBOL'))

# Perform comparison using formula interface
x <- retry(compareCluster(ENTREZID ~ cluster, data = markers, fun = 'enrichKEGG'))

# Visualization
dotplot(x, label_format = 40) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.callout-note}
**Enrichment plots need enrichment result objects, not a table of pathway
statistics.** A frequent question is how to hand `enrichplot` a data frame of
per-pathway statistics (`PValue`, `FDR`, `NES`, `logFC`, ...) obtained elsewhere.
There is no way to do this directly: `dotplot()`, `gseaplot2()`, `ridgeplot()`
and friends dispatch on `enrichResult` / `gseaResult` / `compareClusterResult`
objects, and those carry the gene sets, the gene-to-set mapping and the
background that the plots need. A bare table has none of that.

The way in is to re-run the enrichment inside `clusterProfiler` on the same
gene list and gene sets, which gives you an object the plotting functions
understand:

- **Cut-off gene list per cluster** (what `FindAllMarkers()` gives you) →
  `compareCluster(ENTREZID ~ cluster, data = markers, fun = "enrichGO")`, then
  `dotplot()` / `cnetplot()` / `emapplot()`, as above.
- **A ranked list** (a statistic for every gene, not a cut-off list) →
  `gseGO()` / `GSEA()`, then `gseaplot2()` / `ridgeplot()` / `gseaplot()`.

If the statistics came from another tool (fgsea, GSEA desktop, a vendor
report), re-running the equivalent analysis in `clusterProfiler` with the same
gene list and gene sets is usually a couple of lines, and is the only way to get
the object the visualisation layer requires.
:::

### Using COSG results

Methods like `COSG` return marker genes as a list or data frame where columns represent clusters. Since a data frame is essentially a list of equal-length vectors, it can be passed directly to `compareCluster()`.

```{r eval=FALSE}
#| label: sc-cosg-example
library(COSG)

# Identify markers using COSG
marker_cosg <- cosg(
    sce,
    groups = 'all',
    assay = 'RNA',
    slot = 'data',
    mu = 1,
    n_genes_user = 100
)

# The first element is a data frame of gene symbols
# Columns correspond to clusters
head(marker_cosg[[1]])

# Directly use the data frame for enrichment
y <- compareCluster(marker_cosg[[1]], 
                    fun = 'enrichGO', 
                    OrgDb = 'org.Hs.eg.db', 
                    keyType = 'SYMBOL', 
                    ont = "MF")

# Visualization
dotplot(y, label_format = 60) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This demonstrates that `compareCluster()` can seamlessly handle various data structures commonly produced by single-cell analysis tools, simplifying the downstream functional interpretation.


## Next steps

- Visualize the resulting `compareClusterResult` with [comparison-aware dot and network plots](comparecluster-visualization.qmd#sec-comparecluster-visualization).
- For generic comparison figures beyond `compareCluster()`, see [compare functional profiles](enrichplot-comparison.qmd#sec-visualization-comparison).
- For term redundancy and organization, continue to [semantic similarity](semantic-similarity.qmd).
- For biological reporting, see [evidence-guided interpretation](interpretation.qmd#evidence-guided-interpretation).

## References {-}
