# Disease enrichment analysis {#dose-enrichment}


```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
library(clusterProfiler)
```

We developed `r Biocpkg("DOSE")` [@yu_dose_2015] package to promote the investigation of diseases. `r Biocpkg("DOSE")` provides five methods for [measuring semantic similarities among DO terms and gene products](do-semantic-similarity.qmd#DOSE-semantic-similarity), hypergeometric model and gene set enrichment analysis (GSEA) for associating disease with gene list and extracting disease association insight from genome wide expression profiles.


## Chapter overview

| Aspect | Disease and phenotype enrichment |
|---|---|
| **Questions** | Are selected genes associated with disease/phenotype terms, or are disease gene sets enriched across a ranked profile? |
| **Input** | Entrez IDs for ORA or a named ranked vector for GSEA; choose HDO, HPO, or MPO as appropriate. |
| **Methods** | Disease Ontology and phenotype ORA/GSEA, plus Network of Cancer Gene enrichment. |
| **Main functions** | `enrichDO()`, `gseDO()`, `enrichNCG()`, `gseNCG()`. |
| **Output** | `enrichResult` / `gseaResult` objects for filtering, comparison, and visualization. |
| **Main limitations** | Interpretation depends on ontology and annotation coverage; NCG is a separately curated resource. Keep organism, ID type, universe, and ontology explicit. |



## Disease over-representation analysis {#dose-ora}


`r Biocpkg("DOSE")` supports enrichment analysis of Disease Ontology (DO) [@schriml_disease_2011] and [Network of Cancer Gene](http://ncg.kcl.ac.uk/) [@omer_ncg]. In addition, several visualization methods were provided by `r Biocpkg("enrichplot")` to help interpreting semantic and enrichment results.


### Over-representation analysis for disease ontology {#dose-do-ora}


In the following example, we selected genes with fold change above 1.5 as the differential genes and analyzed their disease association.

```{r}
#| label: enrichdo-example
library(DOSE)
data(geneList)
gene <- names(geneList)[abs(geneList) > 1.5]
head(gene)
x <- enrichDO(gene          = gene,
              ont           = "HDO",
              pvalueCutoff  = 0.05,
              pAdjustMethod = "BH",
              universe      = names(geneList),
              minGSSize     = 5,
              maxGSSize     = 500,
              qvalueCutoff  = 0.05,
              readable      = FALSE)
head(x)
```

The `enrichDO()` function requires an entrezgene ID vector as input, which is mostly the differential gene list from gene expression profile studies. Please refer to @sec-id-convert if you need to convert other gene ID types to entrezgene ID. 


The `ont` parameter can be "HDO" (Human Disease Ontology), "HPO" (Human Phenotype Ontology) or "MPO" (Mouse Phenotype Ontology). `pvalueCutoff` setting the cutoff value of *p* value and adjusted *p* value; `pAdjustMethod` setting the *p* value correction methods, include the Bonferroni correction ("bonferroni"), Holm ("holm"), Hochberg ("hochberg"), Hommel ("hommel"), Benjamini \& Hochberg ("BH") and Benjamini \& Yekutieli ("BY") while `qvalueCutoff` is used to control *q*-values.

The `universe` sets the background gene universe for testing. If users do not explicitly set this parameter, `enrichDO()` will set the universe to all human genes that have DO annotation.

The `minGSSize` (and `maxGSSize`) indicates that only those DO terms that have more than `minGSSize` (and less than `maxGSSize`) annotated genes will be tested.

The `readable` is a logical parameter that indicates whether the entrezgene IDs will be mapped to gene symbols or not, see also @sec-setReadable.


### Over-representation analysis for the network of cancer gene {#dose-ncg-ora}

[Network of Cancer Gene](http://ncg.kcl.ac.uk/) (NCG) [@omer_ncg] is a manually curated repository of cancer genes. NCG release 5.0 (Aug. 2015) collects 1,571 cancer genes from 175 published studies. `r Biocpkg("DOSE")` supports analyzing gene list and determine whether they are enriched in genes known to be mutated in a given cancer type.

```{r}
#| label: enrichncg-example
gene2 <- names(geneList)[abs(geneList) > 3]
ncg <- enrichNCG(gene2) 
head(ncg)
```

### DisGeNET support has been removed {#dose-dgn-ora}


::: {.callout-warning}
**`enrichDGN()`, `enrichDGNv()` and `gseDGN()` no longer exist.** DOSE used to
provide over-representation and GSEA against
[DisGeNET](http://disgenet.org/) [@janet_disgenet], an integrative resource of
gene-disease associations. DisGeNET changed its licence from CC BY-NC-SA to a
proprietary one, so the data can no longer be redistributed with the package,
and the functions and their bundled datasets were removed in DOSE 4.7.0
(June 2026).

If you have code calling them, use the Disease Ontology route instead:

```{r eval=FALSE}
## over-representation
x <- enrichDO(gene, ont = "HDO")

## GSEA
y <- gseDO(geneList, ont = "HDO")
```

`ont = "HDO"` is the human Disease Ontology and is the default. This is a
genuine change of data source: DO annotates genes to *disease terms*, whereas
DisGeNET aggregated *gene-disease associations* from several sources, so the
results are not interchangeable. DisGeNET itself is still available for direct
download from its own website under its current terms.
:::

## Disease gene set enrichment analysis {#dose-gsea}

### `gseDO` function

In the following example, in order to speed up the compilation of this document, only gene sets with size above 120 were tested and only 100 permutations were performed.

```{r}
#| label: gsedo-example
library(DOSE)
data(geneList)
y <- gseDO(geneList,
           minGSSize     = 120,
           pvalueCutoff  = 0.2,
           pAdjustMethod = "BH",
           verbose       = FALSE)
head(y, 3)
```

### `gseNCG` function

```{r}
#| label: gsencg-example
ncg <- gseNCG(geneList,
              pvalueCutoff  = 0.5,
              pAdjustMethod = "BH",
              verbose       = FALSE)
ncg <- setReadable(ncg, 'org.Hs.eg.db')
head(ncg, 3) 
```

## Next steps

- For term relationships, see [DOSE semantic similarity](do-semantic-similarity.qmd).
- To plot result objects, use the [visualization map](enrichplot.qmd#sec-enrichplot).
- For custom collections, see [universal enrichment](universal-enrichment.qmd#universal-api).

## References {-}
