# Multi-omics enrichment integration {#sec-enrichment-multiomics}

This chapter covers the integration layer between molecular measurements and enrichment backends. It separates feature-level aggregation, identifier harmonization, directional conflict handling, pathway-level late fusion, and contribution tracing so that the provenance of a combined result remains inspectable.

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=FALSE)
source("_common.R")
```

## Multi-omics Early Integration

With the rapid development of sequencing technologies, it is very common to perform functional analysis on multi-omics data (e.g., Transcriptomics + Proteomics). A naive approach is to perform enrichment analysis on each omics separately and then intersect the enriched pathways. However, this late-integration strategy completely ignores the fact that different omics might provide weak but consistent signals for the same pathway, which would be filtered out by strict thresholds in single-omics analysis.

To solve this, `enrichit` introduces an **Early Integration** pipeline. The core philosophy is simple: **Integrate first, enrich later. Decouple the integration layer from the enrichment layer.**

### 1. Statistical Aggregation (`aggregate_omics`)

Instead of finding the intersection of pathways, we aggregate the evidence at the feature level. `aggregate_omics()` takes multi-source statistics (e.g., p-values or signed scores) and merges them into a unified evidence score. For p-values, it supports classic methods like `fisher` or `stouffer`, and internally converts them to a monotonically increasing score suitable for ranking. It also supports **Brown's method** for correlation-aware p-value aggregation and `weighted_mean` for signed statistics with explicit layer weights.

```{r}
#| eval: false
library(enrichit)

# Create mock p-values for RNA and Protein data across 1000 genes
set.seed(123)
omics_gene_ids <- paste0("Gene", 1:1000)

rna_pvals <- runif(1000, 0, 1)
names(rna_pvals) <- omics_gene_ids

prot_pvals <- runif(1000, 0, 1)
names(prot_pvals) <- omics_gene_ids

# Inject consistent but weak signals into the first 20 genes (pathway of interest)
rna_pvals[1:20] <- runif(20, 0.01, 0.08)
prot_pvals[1:20] <- runif(20, 0.01, 0.08)

# Aggregate p-values from RNA and Protein data using Fisher's method
agg_fisher_res <- aggregate_omics(
    x = list(RNA = rna_pvals, PROT = prot_pvals),
    method = "fisher",
    input = "pvalue",
    feature_type = "gene"
)

# Look at the top aggregated scores
head(sort(agg_fisher_res$score, decreasing = TRUE))
```

### 1a. Correlation-aware Aggregation with Brown's Method

Fisher's method is simple and powerful, but it silently assumes that the evidences are independent. Real multi-omics layers are often correlated: RNA and protein do not live on separate planets. If you pretend they are independent, you are inviting double counting.

Brown's method fixes this by adjusting the degrees of freedom of Fisher's statistic using an empirical or user-supplied covariance structure. In `enrichit`, this is exposed directly through `aggregate_omics(method = "brown")`.

```{r}
#| eval: false
set.seed(123)
omics_pmat <- cbind(RNA = rna_pvals[1:80], PROT = prot_pvals[1:80])
rownames(omics_pmat) <- omics_gene_ids[1:80]

cov_mat <- matrix(
  c(4, 1.2,
    1.2, 4),
  nrow = 2
)

agg_brown_res <- aggregate_omics(
  x = omics_pmat,
  method = "brown",
  input = "pvalue",
  cov_matrix = cov_mat
)

head(sort(agg_brown_res$score, decreasing = TRUE))
```

### 2. ID Harmonization (`harmonize_ids`)

Proteomics data often use UniProt IDs or protein groups, while your pathway annotations are usually gene-based. Directly feeding protein-level data to gene set enrichment will cause a huge ID mismatch.

`harmonize_ids()` acts as a dedicated ID mapping layer. It maps the protein-level aggregated results to the gene-level, and handles many-to-one mappings gracefully using folding strategies like `min_p` or `max_abs`.

```{r}
#| eval: false
# Suppose our proteomics data used UniProt IDs (e.g., Q9Y6N7)
# and we need to map them back to standard Gene Symbols
id_mapping_df <- data.frame(
    source_id = c("P12345", "P12346", "Q9Y6N7"),
    target_id = c("Gene1", "Gene1", "Gene2") # P12345 and P12346 both map to Gene1
)

# For demonstration, we create a mock protein-level aggregated result
protein_agg_res <- aggregate_omics(
    x = list(PROT = prot_pvals[1:3]),
    method = "fisher",
    input = "pvalue",
    feature_type = "protein"
)
names(protein_agg_res$score) <- c("P12345", "P12346", "Q9Y6N7")
protein_agg_res$feature_id <- c("P12345", "P12346", "Q9Y6N7")

# Map and collapse protein-level features to gene-level
harmonized_res <- harmonize_ids(
    x = protein_agg_res, # Assuming this is an aggregated result from proteomics
    mapping = id_mapping_df,
    from = "protein",
    to = "gene",
    collapse = "min_p" # If multiple proteins map to one gene, take the most significant one
)
```

### 3. Handling Directional Conflicts (`conflict_policy`)

When integrating signed statistics across omics (e.g., transcriptomics and proteomics), you often encounter conflicting directions—a gene might be significantly up-regulated in RNA but down-regulated in Protein.

Instead of naively averaging them out to zero, `aggregate_omics()` introduces a `conflict_policy` parameter when `input = "signed_score"`. You can choose `"strict"` to completely drop conflicting features, or `"penalty"` to halve their aggregated scores. This actively filters out biological noise and reduces false positives before enrichment.

```{r}
#| eval: false
# Example with conflicting directions between RNA and PROT
rna_signed <- -log10(rna_pvals) * rep(c(1, -1), length.out = length(rna_pvals))
prot_signed <- -log10(prot_pvals) * rep(c(1, 1, -1, -1), length.out = length(prot_pvals))
names(rna_signed) <- omics_gene_ids
names(prot_signed) <- omics_gene_ids

agg_strict <- aggregate_omics(
    x = list(RNA = rna_signed, PROT = prot_signed),
    input = "signed_score",
    method = "mean",
    conflict_policy = "strict" # Drops conflicting genes automatically
)
```

### 4. Distribution to Enrichment Backends

Once the multi-omics data is aggregated and harmonized, you obtain a unified object containing a ranking `score`.

For **GSEA** or **NSEA**, you can directly extract the sorted score vector and feed it to the engine:

```{r}
#| eval: false
omics_pathways <- list(TargetPathway = omics_gene_ids[1:20])

geneList <- sort(agg_fisher_res$score, decreasing = TRUE)

gsea_res <- gsea(
    geneList = geneList,
    gene_sets = omics_pathways
)
```

For **ORA**, which requires discrete gene lists, `enrichit` provides an adapter function `select_features_for_ora()`. It takes the aggregated object and a cutoff, and returns the selected `gene` list and background `universe`, perfectly bridging the gap between continuous multi-omics scores and traditional hypergeometric tests.

```{r}
#| eval: false
# Define a mock pathway containing the 20 genes we injected signal into
selected_ora_inputs <- select_features_for_ora(
    x = agg_fisher_res, 
    cutoff = 0.05, 
    by = "pvalue"
)

# Only genes with significant aggregated evidence will be selected
early_ora_res <- ora(
    gene = selected_ora_inputs$gene,
    universe = selected_ora_inputs$universe,
    gene_sets = omics_pathways
)
```

By decoupling ID mapping, statistical integration, and pathway enrichment, `enrichit` provides a clean, robust, and highly extensible pipeline for multi-omics functional analysis.

These workflow helpers are also available from `clusterProfiler`, which means the aggregated result can be sent directly to high-level functions such as `enrichGO()`, `gseGO()`, or `nseGO()` without changing the underlying multi-omics logic.

### 5. Late Fusion at the Pathway Level

Early fusion is elegant when your features align cleanly. But sometimes that assumption collapses immediately: transcriptomics might have a background of 20,000 genes, proteomics might only detect 4,000 proteins, and region-based omics may not even live in gene space before annotation. In these cases, forcing everything into one feature matrix is a great way to manufacture missing values and mapping noise.

This is where **Late Fusion** comes in. Each omics is enriched independently using its own background and preferred backend; then `aggregate_enrichment()` merges the pathway-level p-values across results. The mathematics is the same family as feature-level aggregation, but the biology is much cleaner because each layer is allowed to speak in its native space first.

```{r}
#| eval: false
late_fusion_pathways <- list(
  SharedPath = omics_gene_ids[1:8],
  RNAPath = omics_gene_ids[9:16],
  ProtPath = omics_gene_ids[17:24]
)

rna_sig <- c(omics_gene_ids[1:6], omics_gene_ids[9:13])
prot_sig <- c(omics_gene_ids[c(1:4, 7:8)], omics_gene_ids[17:21])

rna_ora_res <- ora(
  gene = rna_sig,
  universe = omics_gene_ids,
  gene_sets = late_fusion_pathways
)

prot_ora_res <- ora(
  gene = prot_sig,
  universe = omics_gene_ids[c(1:60)],
  gene_sets = late_fusion_pathways
)

late_fusion_res <- aggregate_enrichment(
  res_list = list(RNA = rna_ora_res, PROT = prot_ora_res),
  method = "brown"
)

as.data.frame(late_fusion_res)[, c("ID", "pvalue", "p.adjust", "Count")]
```

### 6. Contribution Tracing

After running a multi-omics integration, you may want to know whether a significant pathway was driven by the transcriptome, the proteome, or their combination. `enrichit` provides tools to trace the contribution back to the original omics data.

First, you can classify the pattern of each pathway by comparing the merged results with single-omics results:

```{r}
#| eval: false
# Run ORA for individual omics
rna_ora_res <- ora(
    gene = omics_gene_ids[rna_pvals < 0.05],
    universe = omics_gene_ids,
    gene_sets = omics_pathways
)
prot_ora_res <- ora(
    gene = omics_gene_ids[prot_pvals < 0.05],
    universe = omics_gene_ids,
    gene_sets = omics_pathways
)

# Classify contribution patterns
classified_ora_res <- classify_omics_pattern(
    merged_res = early_ora_res,
    single_res = list(RNA = rna_ora_res, PROT = prot_ora_res),
    p_cutoff = 0.05
)

head(as.data.frame(classified_ora_res)[, c("ID", "pvalue", "Omics_Pattern")])
```

The `Omics_Pattern` column will label the pathway as `Shared`, `RNA-Specific`, `PROT-Specific`, or `Enhanced (1+1>2)`.

Then, you can extract the gene-level contribution for a specific pathway to see the original statistics of the core enrichment genes:

```{r}
#| eval: false
# Get original multi-omics statistics for genes in the pathway
contribution_df <- get_omics_contribution(
    res = classified_ora_res, 
    agg = agg_fisher_res, 
    pathway_id = "TargetPathway"
)

head(contribution_df, 3)
```

This data frame is extremely useful for downstream visualization, such as drawing multi-omics heatmaps for the core enriched genes.


## Next steps

- For the statistical foundations used by each backend, see [ORA and GSEA](enrichment-foundations.qmd#sec-enrichment-foundations).
- For genomic regions that must first be mapped to genes, see [ChIPseeker](chipseeker-enrichment.qmd#chipseeker-enrichment).
- For result-object organization and interpretation, continue to [evidence integration](03-evidence-integration.qmd) and [evidence-guided interpretation](interpretation.qmd#evidence-guided-interpretation).
- For multi-omics contribution figures, see [specialized visualization](enrichplot-specialized.qmd#sec-visualization-specialized).

## References {-}
