# Visualization of functional enrichment result {#sec-enrichplot}


```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
library(clusterProfiler)
library(enrichplot)
```



The `r Biocpkg("enrichplot")` package implements several visualization methods to help interpret enrichment results. It supports visualizing enrichment results obtained from `r Biocpkg("DOSE")` [@yu_dose_2015], `r Biocpkg("clusterProfiler")` [@yu2012; @wu_clusterprofiler_2021],
`r Biocpkg("ReactomePA")` [@yu_reactomepa_2016] and `r Biocpkg("meshes")` [@yu_meshes_2018]. Both over representation analysis (ORA) and gene set enrichment analysis (GSEA) are supported.

Note: Several visualization methods were first implemented in `r Biocpkg("DOSE")` and rewrote from scratch using `r CRANpkg("ggplot2")`. If you want to use the [old methods](https://www.biostars.org/p/375555), you can use the [doseplot](https://github.com/GuangchuangYu/doseplot) package.


## Bar Plot


Bar plot is the most widely used method to visualize enriched terms. It depicts the enrichment scores (*e.g.* p values) and gene count or ratio as bar height
and color @fig-Barplot. Users can specify the number of terms (most significant) or selected terms (see also the [FAQ](#showing-specific-pathways)) to display via the `showCategory` parameter. 



```{r}
library(DOSE)
data(geneList)
de <- names(geneList)[abs(geneList) > 2]

edo <- enrichDGN(de)
```

```{r eval=F}
library(enrichplot)
barplot(edo, showCategory=20) 
```

Other variables that derived using `mutate` can also be used as bar height or color as demonstrated in @sec-mutate and @fig-Barplot.

```{r eval=F}
library(clusterProfiler)

mutate(edo, qscore = -log(p.adjust, base=10)) |> 
    barplot(x="qscore")
```



```{r}
#| label: fig-Barplot  
#| echo: false
#| fig-width: 11
#| fig-height: 6
#| fig-scap: |
#|   Bar plot of enriched terms.
#| fig-cap: |
#|   **Bar plot of enriched terms.**

p1 <- barplot(edo, showCategory=20) 
p2 <- mutate(edo, qscore = -log(p.adjust, base=10)) |>
    barplot(x="qscore")

library(aplot)
plot_list(p1, p2, ncol=2, tag_levels='A')    
```


## Dot plot


Dot plot is similar to bar plot with the capability to encode another score as dot size.

```r
library(ggplot2)
edo2 <- gseDO(geneList)
dotplot(edo, showCategory=30) + ggtitle("dotplot for ORA")
dotplot(edo2, showCategory=30) + ggtitle("dotplot for GSEA")
```


```{r}
#| label: fig-Dotplotcap  
#| echo: false
#| fig-width: 12
#| fig-height: 10
#| fig-scap: |
#|   Dot plot of enriched terms.
#| fig-cap: |
#|   **Dot plot of enriched terms.**
library(ggplot2)

edo2 <- gseDO(geneList)
p1 <- dotplot(edo, showCategory=30) + ggtitle("dotplot for ORA") 
p2 <- dotplot(edo2, showCategory=30) + ggtitle("dotplot for GSEA") 
plot_list(p1, p2, ncol=2, tag_levels = 'A')
```


Note: The `dotplot()` function also works with [`compareCluster()` output](##compare-dotplot).


## Gene-Concept Network {#cnetplot}

Both the `barplot()` and `dotplot()` only displayed most significant or selected enriched terms,
while users may want to know which genes are involved in these significant
terms. 
To consider the potential biological complexities in which a gene may belong to multiple annotation categories and provide information about numeric changes when available, we developed the `cnetplot()` function to extract the complex associations. 
The `cnetplot()` depicts the linkages of genes and biological concepts (*e.g.* GO terms or KEGG pathways) as a network. GSEA result is also supported
with only core enriched genes displayed.


Users can use the 'fc_threshold' parameter in the `cnetplot` function to filter genes to only include those with `|foldChange| > fc_threshold`. This allows users to plot only high fold-change genes (e.g., `fc_threshold = 1`).

```{r}
#| label: fig-Networkplot  
#| fig-width: 24
#| fig-height: 9
#| fig-scap: |
#|   Network plot of enriched terms.
#| fig-cap: |
#|   **Network plot of enriched terms.**
#| 
## convert gene ID to Symbol
edox <- setReadable(edo, 'org.Hs.eg.db', 'ENTREZID')
p1 <- cnetplot(edox, foldChange=geneList)
p2 <- cnetplot(edox, categorySizeBy=~ -log10(pvalue), foldChange=geneList)
## only plot high fold-change genes
p3 <- cnetplot(edox, foldChange=geneList, fc_threshold = 2)
plot_list(p1, p2, p3, ncol=3, tag_levels = 'A')
```


If you would like label subset of the nodes, you can use the `node_label` parameter, which supports 4 possible selections (i.e. "category", "gene", "all" and "none"), as demonstrated in @fig-cnetNodeLabel. 



```{r}
#| label: fig-cnetNodeLabel  
#| fig-width: 12
#| fig-height: 16
#| fig-scap: |
#|   Labelling nodes by selected subset.
#| fig-cap: |
#|   **Labelling nodes by selected subset.** 
#|   gene category (A), gene name (B), 
#|   both gene category and gene name (C, default) 
#|   and not to label at all (D).
p1 <- cnetplot(edox, node_label="category") 
p2 <- cnetplot(edox, node_label="gene") 
p3 <- cnetplot(edox, node_label="all") 
p4 <- cnetplot(edox, node_label="none", 
        color_category='firebrick', 
        color_item='steelblue') 
plot_list(p1, p2, p3, p4, ncol=2, tag_levels = 'A')
```

The `cnetplot` function can be used as a general method to visualize data relationships in a network diagram. Please refer to the vignette of `r CRANpkg("ggtangle")`. 


Note: The `cnetplot()` function also works with [`compareCluster()` output](##compare-cnetplot).


## Heatmap-like functional classification


The `heatplot` is similar to `cnetplot`, but displays the relationships as a
heatmap. The gene-concept network may become too complicated if users want to
show a large number of significant terms. The `heatplot` can simplify the result
and make it easier to identify expression patterns.

```{r}
#| label: fig-Heatplot  
#| fig-width: 12
#| fig-height: 5
#| fig-scap: |
#|   Heatmap plot of enriched terms.
#| fig-cap: |
#|   **Heatmap plot of enriched terms.** 
#|   default (A), `foldChange=geneList` (B)

p1 <- heatplot(edox, showCategory=5)
p2 <- heatplot(edox, foldChange=geneList, showCategory=5)
plot_list(p1, p2, ncol=1, tag_levels = 'A')
```


## Tree plot

The `treeplot()` function performs hierarchical clustering of enriched terms. It relies on the pairwise similarities of the enriched terms calculated by the `pairwise_termsim()` function, which by default uses Jaccard's similarity index (JC). Users can also use semantic similarity values when supported (*e.g.*, [GO](#GOSemSim), [DO](#DOSE-semantic-similarity), and [MeSH](#meshes-semantic-similarity)).

The default agglomeration method in `treeplot()` is `ward.D`, and users can specify other methods via the `cluster_method` parameter (*e.g.*, 'average', 'complete', 'median', 'centroid', *etc.*; see also the documentation of the `hclust()` function). The `treeplot()` function will cut the tree into several subtrees (specified by the `nCluster` parameter, default is 5) and label subtrees using high-frequency words. This reduces the complexity of the enriched result and improves user interpretation ability.



```{r}
#| label: fig-treeplot  
#| fig-width: 20
#| fig-height: 10
#| fig-scap: |
#|   Tree plot of enriched terms.
#| fig-cap: |
#|   **Tree plot of enriched terms.** 
#|   default (A), `hclust_method = "average"` (B)
library(ggtree)

edox2 <- pairwise_termsim(edox)
p1 <- treeplot(edox2, cladelab_offset=8, tiplab_offset=.3, fontsize_cladelab =5) + 
    hexpand(.2)
p2 <- treeplot(edox2, cluster_method = "average", 
            cladelab_offset=14, tiplab_offset=.3, fontsize_cladelab =5) + 
    hexpand(.3)
aplot::plot_list(p1, p2, tag_levels='A', ncol=2)
```


## Enrichment Map


Enrichment map organizes enriched terms into a network with edges connecting
overlapping gene sets. In this way, mutually overlapping gene sets are tend to
cluster together, making it easy to identify functional module.

The `emapplot` function supports results obtained from hypergeometric test and gene set enrichment analysis. 
The `size_category` parameter can be used to resize nodes and the `layout` parameter can adjust the layout, as demonstrated in @fig-Enrichment.



```{r}
#| label: fig-Enrichment  
#| fig-width: 15
#| fig-height: 12
#| out-width: 95%
#| fig-scap: |
#|   The `emapplot`.
#| fig-cap: |
#|   **The `emapplot`.**
#|   default (A), `node_label="none"` (B), `size_category=1.5` (C), and `layout="with_fr"` (D)
edo <- pairwise_termsim(edo)
p1 <- emapplot(edo) # node_label = "category" (default)
p2 <- emapplot(edo, node_label = "none") 
p3 <- emapplot(edo, node_label = "none", size_category=1.5)
p4 <- emapplot(edo, node_label = "none", layout="with_fr")

plot_list(p1, p2, p3, p4,
        ncol=2, tag_levels = 'A', 
        design="AAAAAA\nBBCCDD", 
        heights = c(1, .3))
```

The `node_label` parameter controls how the labels were dispalyed. The enriched terms will be displayed by default with `node_label="category"` and it can be disable by setting `node_label="none"`.

If `node_label="group"`, the `emapplot` function will clusters the enriched terms into different groups and only group names (determined by wordcloud) will be displayed. If `node_label="all"`, then the enriched terms and group names will be displayed simultaneously.


```{r}
#| label: fig-Enrichment-node-label  
#| fig-width: 16
#| fig-height: 21
#| fig-scap: |
#|   `emapplot` with enriched terms clustering.
#| fig-cap: |
#|   **`emapplot` with enriched terms clustering.**
#|   `node_label="group"` (A) and 
#|   `node_label="all"` (B).
p5 <- emapplot(edo, node_label = "group")
p6 <- emapplot(edo, node_label = "all")

plot_list(p5, p6, 
        ncol=1, tag_levels = 'A')
```

## Biological theme comparison 

The `emapplot` function also supports results obtained from `compareCluster` function of `clusterProfiler` package. In addition to `size_category` and `layout` parameters, the number of circles in the bottom left corner can be adjusted  using the `legend_n` parameteras, and proportion of clusters in the pie chart can be adjusted using the `pie` parameter, when `pie="count"`, the proportion of  clusters in the pie chart is determined by the number of genes, as demonstrated in @fig-Enrichment2.


```{r}
#| label: fig-Enrichment2  
#| fig-width: 16
#| fig-height: 18
#| fig-scap: |
#|   Plot for results obtained from `compareCluster` function of `clusterProfiler` package.
#| fig-cap: |
#|   **Plot for results obtained from `compareCluster` function of `clusterProfiler` package.** 
#|   default (A), `legend_n=2` (B), `pie="count"` (C) and `pie="count",
#|   size_category=1.5, layout="kk"` (D).
library(clusterProfiler)
data(gcSample)
xx <- compareCluster(gcSample, fun="enrichKEGG",
                     organism="hsa", pvalueCutoff=0.05)
xx <- pairwise_termsim(xx)                     
p1 <- emapplot(xx)
p2 <- emapplot(xx) 
p3 <- emapplot(xx, pie="count")
p4 <- emapplot(xx, pie="count", size_category=1.5, layout="kk")
plot_list(p1, p2, p3, p4, ncol=2, tag_levels = 'A')
```


## UpSet Plot


The `upsetplot` is an alternative to `cnetplot` for visualizing the complex
association between genes and gene sets. It emphasizes the gene overlapping
among different gene sets.


```{r}
#| label: fig-upsetORA  
#| fig-width: 12
#| fig-height: 5
#| fig-scap: |
#|   Upsetplot for over-representation analysis.
#| fig-cap: |
#|   **Upsetplot for over-representation analysis.** 
upsetplot(edo)
```

For over-representation analysis, `upsetplot` will calculate the overlaps among different gene sets as demonstrated in @fig-upsetORA. For GSEA result, it will plot the fold change distributions of different categories (e.g. unique to pathway, overlaps among different pathways).




```{r}
#| label: fig-upsetGSEA  
#| fig-width: 12
#| fig-height: 5
#| fig-scap: |
#|   Upsetplot for gene set enrichment analysis.
#| fig-cap: |
#|   **Upsetplot for gene set enrichment analysis.** 
kk2 <- gseKEGG(geneList     = geneList,
               organism     = 'hsa',
               minGSSize    = 120,
               pvalueCutoff = 0.05,
               verbose      = FALSE)
upsetplot(kk2) 
```


## ridgeline plot for expression distribution of GSEA result

The `ridgeplot` will visualize expression distributions of core enriched genes
for GSEA enriched categories. It helps users to interpret up/down-regulated pathways.


```{r}
#| label: fig-ridgeplot  
#| fig-width: 12
#| fig-height: 12
#| fig-scap: |
#|   Ridgeplot for gene set enrichment analysis.
#| fig-cap: |
#|   **Ridgeplot for gene set enrichment analysis.** 
ridgeplot(edo2)
```


## running score and preranked list of GSEA result

Running score and preranked list are traditional methods for visualizing GSEA
result. The `r Biocpkg("enrichplot")` package supports both of them to visualize
the distribution of the gene set and the enrichment score.


```{r}
#| label: fig-gseaplot  
#| fig-width: 12
#| fig-height: 16
#| fig-scap: |
#|   gseaplot for GSEA result(`by = "runningScore"`).
#| fig-cap: |
#|   **gseaplot for GSEA result(`by = "runningScore"`).** 
#|   `by = "runningScore"` (A), `by = "preranked"` (B), default (C)
p1 <- gseaplot(edo2, geneSetID = 1, by = "runningScore", title = edo2$Description[1])
p2 <- gseaplot(edo2, geneSetID = 1, by = "preranked", title = edo2$Description[1])
p3 <- gseaplot(edo2, geneSetID = 1, title = edo2$Description[1])
plot_list(p1, p2, p3, ncol=1, tag_levels='A')
```


Another method to plot GSEA result is the `gseaplot2` function:


```{r}
#| label: fig-gseaplot2  
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result.
#| fig-cap: |
#|   **Gseaplot2 for GSEA result.** 
gseaplot2(edo2, geneSetID = 1, title = edo2$Description[1])
```

The `gseaplot2` also supports multile gene sets to be displayed on the same figure:


```{r}
#| label: fig-gseaplot22  
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result of multile gene sets.
#| fig-cap: |
#|   **Gseaplot2 for GSEA result of multile gene sets.** 
gseaplot2(edo2, geneSetID = 1:3)
```

User can also displaying the pvalue table on the plot via `pvalue_table`
parameter:


```{r}
#| label: fig-gseaplot23  
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result of multile gene sets(add pvalue_table).
#| fig-cap: |
#|   **Gseaplot2 for GSEA result of multile gene sets(add pvalue_table).** 
gseaplot2(edo2, geneSetID = 1:3, pvalue_table = TRUE,
          color = c("#E495A5", "#86B875", "#7DB0DD"), ES_geom = "dot")
```


The `pvalue_table` can be customized using the following parameters:

- `pvalue_table_rownames` to specify row names of the table (if NULL, no row names will be displayed)
-  `pvalue_table_colnames` to specify column names of the table


```{r}
#| label: fig-gseaplot-pvalue  
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 with customized pvalue table.
#| fig-cap: |
#|   **Gseaplot2 with customized pvalue table.** 
#| 
gseaplot2(edo2, geneSetID = 1, pvalue_table = TRUE,
          pvalue_table_rownames = NULL,
          pvalue_table_columns = c("ID", "NES", "p.adjust"))
```


User can specify `subplots` to only display a subset of plots:


```{r}
#| label: fig-gseaplot24
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result of multile gene sets(add subplots).
#| fig-cap: |
#|   **Gseaplot2 for GSEA result of multile gene sets(add subplots).** 
#|   `subplots = 1` (A),`subplots = 1:2` (B)
#| 
p1 <- gseaplot2(edo2, geneSetID = 1:3, subplots = 1)
p2 <- gseaplot2(edo2, geneSetID = 1:3, subplots = 1:2)
plot_list(p1, p2, ncol=1, tag_levels = 'A')
```




The `gsearank` function plot the ranked list of genes belong to the specific
gene set.


```{r}
#| label: fig-gsearank
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Ranked list of genes belong to the specific gene set.
#| fig-cap: |
#|   **Ranked list of genes belong to the specific gene set.**
gsearank(edo2, 1, title = edo2[1, "Description"])
```

Multiple gene sets can be aligned using `cowplot`:

```{r}
#| label: fig-gsearank2
#| fig-width: 8
#| fig-height: 12
#| fig-scap: |
#|   Gsearank for multiple gene sets.
#| fig-cap: |
#|   **Gsearank for multiple gene sets.**
#| 
library(ggplot2)

pp <- lapply(1:3, function(i) {
    anno <- edo2[i, c("NES", "pvalue", "p.adjust")]
    lab <- paste0(names(anno), "=",  round(anno, 3), collapse="\n")

    gsearank(edo2, i, edo2[i, 2]) + xlab(NULL) +ylab(NULL) +
        annotate("text", 10000, edo2[i, "enrichmentScore"] * .75, label = lab, hjust=0, vjust=0)
})
plot_list(gglist=pp, ncol=1)
```



## pubmed trend of enriched terms

One of the problem of enrichment analysis is to find pathways for further
investigation. Here, we provide `pmcplot` function to plot the number/proportion
of publications trend based on the query result from PubMed Central. Of course,
users can use `pmcplot` in other scenarios. All text that can be queried on PMC
is valid as input of `pmcplot`.


```{r}
#| label: fig-pmcplot
#| fig-width: 14
#| fig-height: 6
#| fig-scap: |
#|   Pmcplot of enrichment analysis.
#| fig-cap: |
#|   **Pmcplot of enrichment analysis.**
#| 
terms <- edo$Description[1:5]
p <- pmcplot(terms, 2010:2020)
p2 <- pmcplot(terms, 2010:2020, proportion=FALSE)
plot_list(p, p2, ncol=2)
```



<!--

using ggplot2 api

`[` accessor


ggplot(egobp[1:20], aes(x=reorder(Description, -pvalue), y=Count, fill=-p.adjust)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_continuous(low="blue", high="red") +
    labs(x = "", y = "", fill = "p.adjust") +
    theme(axis.text=element_text(size=11))

-->

## References {-}
