# Common summary plots for enrichment results {#sec-visualization-summary}

| Aspect | Summary plots |
|---|---|
| **Question** | Which terms are enriched, how should they be ranked or filtered, and what information should be shown first? |
| **Input** | A checked `enrichResult`, `gseaResult`, or compatible result object with term statistics and labels. |
| **Functions** | `barplot()`, `dotplot()`, and summary-oriented extensions such as Manhattan or rich-factor views. |
| **Output** | Compact term-level figures for screening and reporting; the figure does not change the result object. |
| **Limitations** | Ordering and display choices can hide universe differences, effect/coverage, redundancy, and direction. Do not treat visual rank as a new statistical test. |

Use these plots after checking the result object and before moving to [gene–term networks](enrichplot-networks.qmd#sec-visualization-networks) or [GSEA views](enrichplot-gsea.qmd#sec-visualization-gsea).


```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
suppressPackageStartupMessages({
    library(clusterProfiler)
    library(enrichplot)
    library(dplyr)
})
```

## Bar Plot


Bar plot is the most widely used method to visualize enriched terms. It depicts the enrichment scores (*e.g.* p values) and gene count or ratio as bar height
and color @fig-Barplot. Users can specify the number of terms (most significant) or selected terms (see also the [FAQ](troubleshooting.qmd#showing-specific-pathways)) to display via the `showCategory` parameter. 



```{r}
#| label: dose-enrichdo-setup
library(DOSE)
data(geneList)
de <- names(geneList)[abs(geneList) > 2]

edo <- enrichDO(de)
```

```{r eval=F}
#| label: fig-barplot-eval-false
library(enrichplot)
barplot(edo, showCategory=20) 
```

Other variables that derived using `mutate` can also be used as bar height or color as demonstrated in @sec-mutate and @fig-Barplot.

```{r eval=F}
#| label: fig-mutate-barplot-eval-false
library(clusterProfiler)

mutate(edo, qscore = -log(p.adjust, base=10)) |> 
    barplot(x="qscore")
```

::: {.callout-note}
`library(clusterProfiler)` is not optional here. `mutate()` on an `enrichResult` works through an S3 method that `clusterProfiler` registers on load, so if only `DOSE` (or another downstream package) has been loaded, `mutate()` falls back to `dplyr::mutate` and fails with `no applicable method for 'mutate' applied to an object of class "enrichResult"`.
:::



```{r}
#| label: fig-Barplot  
#| echo: false
#| fig-width: 11
#| fig-height: 6
#| fig-scap: |
#|   Bar plot of enriched terms.
#| fig-cap: |
#|   **Bar plot of enriched terms.**

p1 <- barplot(edo, showCategory=20) 
p2 <- mutate(edo, qscore = -log(p.adjust, base=10)) |>
    barplot(x="qscore")

library(aplot)
plot_list(p1, p2, ncol=2, tag_levels='A')    
```

### Split top categories by ontology

When visualizing GO enrichment across all ontologies (`ont = "ALL"`), it is often more informative to show the top terms within each ontology (BP/CC/MF) rather than selecting the top terms globally. `barplot()` forwards additional arguments via `...`, and one useful option is `split`. Setting `split = "ONTOLOGY"` will perform the “top N per ontology” selection and makes it straightforward to color (and later facet) the results by the ontology.

```{r}
#| label: fig-enrichGO-ontology
#| echo: false
#| message: false
#| warning: false
library(DOSE)
data(geneList, package = "DOSE")
de <- names(geneList)[1:300]

library(clusterProfiler)
library(org.Hs.eg.db)
ego_all <- enrichGO(
  gene = de,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "ALL",
  readable = TRUE
)
```

```{r}
#| label: fig-barplot-ontology-nofacet
#| fig-width: 12
#| fig-height: 9
#| fig-scap: |
#|   Bar plot of GO terms using `split = "ONTOLOGY"` without faceting.
#| fig-cap: |
#|   **Bar plot of GO terms using `split = "ONTOLOGY"` (no faceting).**
library(enrichplot)
library(ggplot2)

p_nofacet <- barplot(
  ego_all,
  x = "Count",
  showCategory = 15,
  split = "ONTOLOGY"
) +
  aes(fill = ONTOLOGY) +
  scale_fill_brewer(palette = "Set2")

p_nofacet +
  geom_text(aes(label = Count), nudge_x = 2) +
  scale_y_discrete()
```

Note that `barplot()` wraps long term descriptions onto multiple lines by default (controlled by `label_format`, default is 30). If you prefer to keep the y-axis labels on a single line, you can override the default scale by adding `+ scale_y_discrete()` as shown above.

```{r}
#| label: fig-barplot-ontology-facet
#| fig-width: 12
#| fig-height: 9
#| fig-scap: |
#|   Bar plot of GO terms with `autofacet()` splitting by ontology.
#| fig-cap: |
#|   **Bar plot of GO terms with `autofacet()` splitting by ontology.**
p_facet <- barplot(
  ego_all,
  x = "Count",
  showCategory = 15,
  split = "ONTOLOGY"
) +
  aes(fill = ONTOLOGY) +
  scale_fill_brewer(palette = "Set2") +
  enrichplot::autofacet(by = "row", scales = "free")

p_facet + scale_y_discrete()
```

If the fortified data contain an `ONTOLOGY` column (as in the output generated by `split = "ONTOLOGY"`), you can facet the plot directly with `+ enrichplot::autofacet()`, and it will automatically split panels by ontology.


## Dot plot


Dot plot is similar to bar plot with the capability to encode another score as dot size.

```{r}
#| label: fig-dotplot-setup
library(ggplot2)
edo2 <- gseDO(geneList)
dotplot(edo, showCategory=30, label_format=NULL) + ggtitle("dotplot for ORA")
dotplot(edo2, showCategory=30, label_format=NULL) + ggtitle("dotplot for GSEA")
```


```{r}
#| label: fig-Dotplotcap  
#| echo: false
#| fig-width: 12
#| fig-height: 10
#| fig-scap: |
#|   Dot plot of enriched terms.
#| fig-cap: |
#|   **Dot plot of enriched terms.**
library(ggplot2)

edo2 <- gseDO(geneList)
p1 <- dotplot(edo, showCategory=30) + ggtitle("dotplot for ORA") 
p2 <- dotplot(edo2, showCategory=30) + ggtitle("dotplot for GSEA") 
plot_list(p1, p2, ncol=2, tag_levels = 'A')
```


Note: The `dotplot()` function also works with [`compareCluster()` output](enrichplot-comparison.qmd#compare-dotplot).

### Highlighting specific pathways

The `label_format` parameter in `dotplot()` accepts a function to format y-axis labels, which can be used to highlight specific pathways. This is useful for emphasizing pathways of interest in publication figures.

First, perform enrichment analysis:

```{r}
#| label: fig-enrichDO
library(DOSE)
data(geneList)
de <- names(geneList)[1:100]
x <- enrichDO(de)
```

Define a function to highlight pathways containing "cancer":

```{r}
#| label: fig-label-format-func
f <- function(ids) {
    i <- grepl('cancer', ids)
    ids[i] <- paste0("<i style='color:#009E73'>", ids[i], "</i>")
    return(ids)
}
```

To render the HTML formatting, use the `ggtext` package with `element_markdown()`:

```{r}
#| label: fig-dotplot-highlight
library(ggplot2)
library(ggtext)
library(enrichplot)
dotplot(x, label_format=f) + theme(axis.text.y = element_markdown())
```

You can also highlight specific pathways by index:

```{r}
#| label: fig-label-format-func2
f2 <- function(ids) {
    ids[1] <- paste0("<i style='color:#009E73'>", ids[1], "</i>")
    ids[2] <- paste0("<i style='color:#0072B2'>**", ids[2], "**</i>")
    return(ids)
}

dotplot(x, label_format=f2) + theme(axis.text.y = element_markdown())
```

Or apply different colors to all pathways:

```{r}
#| label: fig-label-format-func3
f3 <- function(ids) {
    # cols <- rcartocolor::carto_pal(length(ids), "Vivid")
    # cols <- colorspace::rainbow_hcl(length(ids))
    cols <- rainbow(length(ids))
    ids <- paste0("<i style='color:", cols, "'>**", ids, "**</i>")
    return(ids)
}

dotplot(x, label_format=f3) + theme(axis.text.y = element_markdown())
```

### Formula interface of dotplot

The `x` variable of `dotplot()` supports a formula interface, allowing users to use derived variables. For example, we can calculate `GeneRatio/BgRatio` or `-log(p.adjust)` and use them as the x-axis variable.

```{r}
#| label: fig-dotplot-formula
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Dotplot with formula interface.**
#|   (A) `x = ~GeneRatio/BgRatio`.
#|   (B) `x = ~ -log(p.adjust)`.
p1 <- dotplot(edo, x = ~GeneRatio/BgRatio)
p2 <- dotplot(edo, x = ~ -log(p.adjust))
plot_list(p1, p2, ncol=2, tag_levels='A')
```

This feature provides great flexibility in visualization. For instance, the **Rich Factor** is a common metric defined as the ratio of the number of differentially expressed genes annotated in a pathway to the total number of genes annotated in that pathway.

$$Rich Factor = \frac{Count}{BgRatio \times N}$$

where $N$ is the total number of genes in the background distribution.

We can easily visualize the **Rich Factor** using the formula interface.

```{r eval=FALSE}
#| label: fig-rich-factor
N <- 8007 # total number of genes in the background
dotplot(edo, x = ~Count/(BgRatio * N))
```

Alternatively, you can precompute derived variables with `mutate()` (e.g., add `neglog10p = -log10(p.adjust)` or `richFactor = Count / (BgRatio * N)`) and then pass the new column name to `x`. This approach is useful when you want to reuse the computed variables across multiple plots. See the dedicated section on using dplyr verbs with enrichment results in @sec-mutate.

### Adjusting dot sizes

To increase the size of the dots in the dot plot, we can use the `scale_size` function from the `ggplot2` package. This allows us to specify the range of the dot sizes.

```{r}
#| label: fig-dotplot-size
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Dotplot with adjusted dot sizes.**
#|   (A) Default dotplot.
#|   (B) Dotplot with `scale_size(range=c(2, 20))`.
p1 <- dotplot(edo, showCategory=10) + ggtitle("Default")
p2 <- dotplot(edo, showCategory=10) + scale_size(range=c(2, 20)) + ggtitle("Adjusted size")
plot_list(p1, p2, ncol=2, tag_levels='A')
```

## Manhattan plot

The `manhattanplot()` function provides a landscape-style overview of enrichment results. It arranges enriched terms along the x-axis by ontology (or by category in `compareCluster()` results), uses `-log10()` transformed significance values on the y-axis, and maps term size to a variable such as `Count`. The most significant terms can be labeled automatically with `showCategory`, making it useful for summarizing a large number of enriched terms in a single plot.

When GO enrichment is performed with `ont = "ALL"`, `manhattanplot()` can display the top terms across BP, CC, and MF simultaneously by combining it with `split = "ONTOLOGY"`.

```{r}
#| label: fig-manhattanplot
#| fig-width: 12
#| fig-height: 7
#| fig-cap: |
#|   **Manhattan plot of GO enrichment results.**
#|   Terms are grouped by ontology and plotted by `-log10(p.adjust)`.
manhattanplot(
  ego_all,
  color = "p.adjust",
  showCategory = 12,
  size = "Count",
  split = "ONTOLOGY",
  title = "GO enrichment landscape"
)
```

The `manhattanplot()` function also supports `gseaResult`, `enrichResultList`, `gseaResultList`, and `compareClusterResult` objects, which makes it convenient for comparing the distribution of significant terms across multiple ontologies or experimental groups.

## Next steps

- For gene–term relationships, continue to [network plots](enrichplot-networks.qmd#sec-visualization-networks).
- For result comparison, see [compare functional profiles](enrichplot-comparison.qmd#sec-visualization-comparison).

