# Enrichment foundations: ORA and GSEA {#sec-enrichment-foundations}

This chapter answers the first statistical decision: should a thresholded gene list be tested with over-representation analysis (ORA), or should the complete ranking be analyzed with GSEA? The examples retain the original algorithm labels so historical links remain meaningful.

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=FALSE)
source("_common.R")
```

## Over-Representation Analysis (ORA) {#ora-algorithm}

Over Representation Analysis (ORA) [@boyle2004] is a widely used approach to determine whether known biological functions or processes are over-represented (= enriched) in an experimentally-derived gene list, *e.g.* a list of differentially expressed genes (DEGs).

`enrichit` implements ORA using the **hypergeometric distribution** (one-sided Fisher's exact test). The p-value is calculated as the probability of observing at least $k$ genes from the specific gene set in the selected list of $n$ genes, given a background population (universe) of $N$ genes containing $M$ genes from that set.

$$
p = 1 - \sum_{i=0}^{k-1} \frac{\binom{M}{i} \binom{N-M}{n-i}}{\binom{N}{n}}
$$

**Example:** Suppose we have 17,980 genes detected in a Microarray study and 57 genes were differentially expressed. From these, 2641 are annotated to gene set of interest. Among the differentially expressed genes, 28 belong to the gene set^[example adopted from <https://guangchuangyu.github.io/2012/04/enrichment-analysis/>]. 

```{r}
#| eval: false
#| label: enrichment-data
d <- data.frame(DE_gene=c(28, 29), non_DE_gene=c(2613, 15310))
row.names(d) <- c("In_category", "not_in_category")
d
```                        

The orientation of the contingency table matters here. `fisher.test()` with `alternative = "greater"` tests whether the **top-left** cell is larger than expected under independence, so the top-left cell must hold the observed overlap — the 28 differentially expressed genes that belong to the gene set. Written this way, the alternative hypothesis reads exactly like our question: *are the DE genes over-represented in the gene set?*

Whether the gene set of interest is significantly over-represented in the differentially expressed genes can then be assessed using a hypergeometric distribution. This corresponds to a one-sided version of Fisher's exact test.

```{r}
#| eval: false
#| label: fisher-test
fisher.test(d, alternative = "greater")
```

::: {.callout-note}
**`universe` must be a character vector.** It is matched against gene IDs as strings, so a numeric background (for example `universe = as.numeric(ids)`) matches nothing and is discarded — the analysis then silently falls back to the default background, which shifts every p-value with no sign that the argument was ignored.

Since clusterProfiler 4.21.2 you are told: the enrichment functions warn that they will ignore a non-character `universe`, and `compareCluster()` raises that warning itself — it wraps each per-cluster call in `suppressMessages()`, which previously swallowed the message completely. Pass the IDs as character:

```{r}
#| eval: false
universe = as.character(background_ids)
```
:::

While mathematically sound, this classic model makes a naive assumption: it assumes that every gene has an *equal probability* of being detected. We will see how to fix this systematic bias later in the [Weighted Enrichment Analysis](enrichment-network.qmd#weighted-enrichment) section.

## Gene Set Enrichment Analysis (GSEA) {#gsea-algorithm}

Unlike ORA, which requires an arbitrary threshold to define "significant" genes and ignores the rest, GSEA uses the entire ranked list of genes. It calculates an **Enrichment Score (ES)** that reflects the degree to which a gene set is over-represented at the top or bottom of a ranked list.

There are three key elements of the GSEA method:

* Calculation of an Enrichment Score.
  + The enrichment score (_ES_) represents the degree to which a set _S_ is over-represented at the top or bottom of the ranked list _L_. The score is calculated by walking down the list _L_, increasing a running-sum statistic when we encounter a gene in _S_ and decreasing when it is not encountered. The magnitude of the increment depends on the gene statistics (e.g., correlation of the gene with phenotype). The _ES_ is the maximum deviation from zero encountered in the random walk; it corresponds to a weighted Kolmogorov-Smirnov(KS)-like statistic [@subramanian_gene_2005].
* Estimation of Significance Level of _ES_.
  + The _p_-value of the _ES_ is calculated using a permutation test. Specifically, we permute the gene labels of the gene list _L_ and recompute the _ES_ of the gene set for the permutated data, which generates a null distribution for the _ES_. The _p_-value of the observed _ES_ is then calculated relative to this null distribution.
* Adjustment for Multiple Hypothesis Testing.
  + When the entire gene sets are evaluated, the estimated significance level is adjusted to account for multiple hypothesis testing and also _q_-values are calculated for FDR control.

We implemented the GSEA algorithm proposed by Subramanian [@subramanian_gene_2005]. `enrichit` offers a fast C++ implementation of GSEA. The default method (`method = "multilevel"`) uses an adaptive multi-level splitting Monte Carlo approach (derived from `fgsea` [@korotkevich_fast_2019]) to estimate low p-values efficiently with high accuracy. 

```{r}
#| eval: false
library(enrichit)

# Generate synthetic ranked gene list
set.seed(42)
geneList <- sort(rnorm(1000), decreasing = TRUE)
names(geneList) <- paste0("Gene", 1:1000)

gene_sets <- list(
  PathwayTop = names(geneList)[1:50],
  PathwayBottom = names(geneList)[951:1000]
)

# Run GSEA using the multilevel method
gsea_result <- gsea(
  geneList = geneList,
  gene_sets = gene_sets,
  method = "multilevel"
)
```

### Leading edge analysis and core enriched genes

Leading edge analysis reports `Tags` to indicate the percentage of genes contributing to the enrichment score, `List` to indicate where in the list the enrichment score is attained and `Signal` for enrichment signal strength.

It would also be very interesting to get the core enriched genes that contribute to the enrichment. Our packages (`clusterProfiler`, `DOSE`, `meshes` and `ReactomePA`) support leading edge analysis and report core enriched genes in GSEA analysis.

### One-sided or two-sided: the `scoreType` argument {#scoretype}

`scoreType` controls which end of the ranked list the enrichment score is allowed to look at. It takes three values, and the choice is not cosmetic — it decides which gene sets are even *visible* to the test:

| `scoreType` | ES is | Detects |
|---|---|---|
| `"std"` (default) | the maximum deviation from zero in **either** direction | gene sets enriched at the top **and** at the bottom |
| `"pos"` | the maximum deviation **above** zero only | gene sets enriched at the top only |
| `"neg"` | the maximum deviation **below** zero only | gene sets enriched at the bottom only |

With the usual signed metric (a log fold change or a signed correlation), the default `"std"` is what you want: it reports both up- and down-regulated sets. Running the three settings on the same mixed-sign ranking makes the difference obvious — `"pos"` misses the bottom-enriched set entirely (its ES is 0 and its p-value is 1), and `"neg"` misses the top-enriched one.

```{r}
#| label: gsea-scoretype
#| eval: false
## gene sets enriched at the top and at the bottom of the same ranking
gsea(geneList, gene_sets, scoreType = "std")   # finds both
gsea(geneList, gene_sets, scoreType = "pos")   # only the top set
gsea(geneList, gene_sets, scoreType = "neg")   # only the bottom set
```

The case that trips people up is a metric that has **no sign at all** — a score where higher simply means "more strongly associated with the phenotype", such as an evidence score, a correlation magnitude, or a rank-based statistic that runs from 1 to $n$. Here every value is positive, so the "bottom" of the list is not a meaningful direction and `"std"` is the wrong model. Use `scoreType = "pos"`:

```{r}
#| label: gsea-scoretype-pos
#| eval: false
## all-positive evidence scores: higher = more associated
gsea(evidence_score, gene_sets, scoreType = "pos")
```

The package warns you when it can detect this situation (`All values in the stats vector are greater than zero and scoreType is "std", maybe you should switch to scoreType = "pos"`), but the warning is easy to ignore — and it only fires when *every* value is positive, so it will not catch a metric that is positive except for a handful of zeros. Note also that the normalized enrichment score (NES) is computed against a null that depends on `scoreType`, so comparing NES values obtained with different settings is not meaningful.

### Choosing between ORA and GSEA {#ora-vs-gsea}

A question that comes up constantly in teaching: *given my experiment, which one should I run?* The deciding factor is mostly **how many genes pass your significance threshold** — not personal preference, and not the analysis software you happen to have installed.

| Situation | Recommended | Why |
|---|---|---|
| Many DEGs (rule of thumb: dozens and up) | ORA | The gene list is long enough for the hypergeometric test to have real power, and the result is easy to read term by term. |
| Few DEGs (say, under ~20–30) | GSEA | ORA has almost no power on a short list: with $n$ small, only a very strong overlap can reach significance. GSEA ranks the *whole* list, so a modest but coordinated shift across a gene set is still detectable. |
| No DEG passes the cutoff at all | GSEA | ORA has literally nothing to test. GSEA only needs a ranking, so it still works when the threshold yields an empty list. |
| You do not want to commit to an arbitrary cutoff | GSEA | No threshold is involved anywhere in the method. |
| You want a per-term statement of the form "this set is over-represented in my DE genes" | ORA | That is exactly the question the hypergeometric test answers. |

The failure mode to avoid is running ORA on a handful of genes. With, say, ten DEGs the test can only detect very large overlaps, so it typically returns nothing — and it is easy to misread that null result as "no biology here" rather than "not enough data for this test". In that regime GSEA (or the network-based and weighted methods introduced below) is the appropriate tool.

The two methods are not mutually exclusive, and reporting both is common. They answer slightly different questions: ORA asks whether your *thresholded* genes hit a set more often than chance, while GSEA asks whether a set is *shifted* toward one end of the *full* ranking.


## Next steps

- Connect the selected method to [GO, KEGG, Reactome, or another knowledge source](02-knowledge-sources.qmd).
- For network topology or detection bias, continue to [network-aware and weighted enrichment](enrichment-network.qmd#sec-enrichment-network).
- For several molecular layers, continue to [multi-omics integration](enrichment-multiomics.qmd#sec-enrichment-multiomics).
- Preserve the result object before [evidence integration](03-evidence-integration.qmd) and [visualization](enrichplot.qmd#sec-enrichplot).

## References {-}
