# Gene–term and term–term networks {#sec-visualization-networks}

| Aspect | Network views |
|---|---|
| **Question** | Which genes, terms, and functional modules are connected, shared, or redundant? |
| **Input** | A checked enrichment result, gene–term membership, and optionally a term-similarity matrix or PPI evidence layer. |
| **Functions** | `cnetplot()`, `heatplot()`, `treeplot()`, `emapplot()`, `upsetplot()`, and related network helpers. |
| **Output** | Gene–term, term–term, hierarchy, overlap, or evidence-network figures. |
| **Limitations** | Dense graphs are difficult to audit; edge definitions differ between semantic similarity, shared genes, and PPI. Keep the object, filtering rule, and evidence source with the figure. |

Use network views after result filtering and term organization; compare them with the simpler [summary plots](enrichplot-summary.qmd#sec-visualization-summary) before making a biological claim.


```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
suppressPackageStartupMessages({
    library(clusterProfiler)
    library(enrichplot)
    library(ggplot2)
    library(dplyr)
})
```

```{r}
# Recreate all objects used by the network examples in this chapter.
library(DOSE)
library(org.Hs.eg.db)
data(geneList, package = "DOSE")
de <- names(geneList)[abs(geneList) > 2]
edo <- enrichDO(de)
edo2 <- gseDO(geneList)
```

## Gene-Concept Network {#cnetplot}

```{r}
#| label: fig-cnet-setup
#| echo: false
#| message: false
#| warning: false
library(clusterProfiler)
library(DOSE)
library(enrichplot)
library(org.Hs.eg.db)
data(geneList, package="DOSE")
de <- names(geneList)[abs(geneList) > 2]
edo <- enrichDO(de)
```

Both the `barplot()` and `dotplot()` only displayed most significant or selected enriched terms,
while users may want to know which genes are involved in these significant
terms. 
To consider the potential biological complexities in which a gene may belong to multiple annotation categories and provide information about numeric changes when available, we developed the `cnetplot()` function to extract the complex associations. 
The `cnetplot()` depicts the linkages of genes and biological concepts (*e.g.* GO terms or KEGG pathways) as a network. GSEA result is also supported
with only core enriched genes displayed. For GSEA results, `cnetplot` specifically visualizes the core enriched genes identified through leading edge analysis, which represent the subset of genes most responsible for driving the enrichment signal.


Users can use the `fc_threshold` parameter in the `cnetplot` function to filter genes to only include those with `|foldChange| > fc_threshold`. This allows users to plot only high fold-change genes (e.g., `fc_threshold = 1`). In addition, users can use the `node_label` parameter to label genes that are shared between groups of terms (i.e. `node_label = "share"`).

::: {.callout-tip}
**Too many labels overlapping?** `cnetplot()` draws gene and category labels with `ggrepel`, which hides labels that would collide. If you want every label shown, raise the limit — `options(ggrepel.max.overlaps = Inf)`. If the network is simply too dense, it is usually better to reduce what gets drawn rather than to force all labels out: label only the categories with `node_label = "category"`, or keep only the genes you care about with `fc_threshold` (e.g. `fc_threshold = 1` keeps `|foldChange| > 1`) or by passing a `foldChange` vector that contains just those genes.
:::

```{r}
#| label: fig-Networkplot  
#| fig-width: 16
#| fig-height: 18
#| fig-scap: |
#|   Network plot of enriched terms.
#| fig-cap: |
#|   **Network plot of enriched terms.**
#| 
## convert gene ID to Symbol
edox <- setReadable(edo, 'org.Hs.eg.db', 'ENTREZID')
p1 <- cnetplot(edox, foldChange=geneList)
p2 <- cnetplot(edox, categorySizeBy=~ -log10(pvalue), foldChange=geneList)
## only plot high fold-change genes
p3 <- cnetplot(edox, foldChange=geneList, fc_threshold = 2)
p4 <- cnetplot(edox, foldChange=geneList, node_label = "share")
plot_list(p1, p2, p3, p4, ncol=2, tag_levels = 'A')
```


If you would like label subset of the nodes, you can use the `node_label` parameter, which supports 4 possible selections (i.e. "category", "gene", "all" and "none"), as demonstrated in @fig-cnetNodeLabel.

The `node_label` parameter also supports enhanced functionality:

1. **Vector selection**: Specify specific genes to label using a character vector, e.g., `node_label = c("gene1", "gene2")`
2. **'exclusive' labeling**: Label only genes that belong exclusively to one category using `node_label = "exclusive"`
3. **'share' labeling**: Label genes that are shared between multiple categories using `node_label = "share"` (as shown in the example above)
4. **Conditional filtering**: Use comparison operators to filter genes based on fold change, e.g., `node_label = "> 1"` or `node_label = "< -1"` to label genes with absolute fold change greater than 1 



```{r}
#| label: fig-cnetNodeLabel  
#| fig-width: 12
#| fig-height: 16
#| fig-scap: |
#|   Labelling nodes by selected subset.
#| fig-cap: |
#|   **Labelling nodes by selected subset.** 
#|   gene category (A), gene name (B), 
#|   both gene category and gene name (C, default) 
#|   and not to label at all (D).
p1 <- cnetplot(edox, node_label="category") 
p2 <- cnetplot(edox, node_label="gene") 
p3 <- cnetplot(edox, node_label="all") 
p4 <- cnetplot(edox, node_label="none", 
        color_category='firebrick', 
        color_item='steelblue') 
plot_list(p1, p2, p3, p4, ncol=2, tag_levels = 'A')
```

The `cnetplot` function can be used as a general method to visualize data relationships in a network diagram. Please refer to the vignette of `r CRANpkg("ggtangle")`. 


### Customizing gene colors in cnetplot

Users can color specific genes by providing a named vector of fold changes containing only those genes.

```{r}
#| label: fig-cnetplot-custom-color
foldChange <- c(rep(1, 6), rep(-1, 4))
names(foldChange) <- c("MARCO", "GZMB", "CXCL11", "CXCL10",
                       "LAG3", "CCL8", "PDK1", "GABRP", "MELK", "CENPE")
p <- cnetplot(edox, foldChange=foldChange)
p
```

If you want to remove the color legend:

```{r}
#| label: fig-cnetplot-no-legend
p <- p + guides(color='none')
p
```

To create a custom legend, you can use a dummy data frame and `geom_point`:

```{r}
#| label: fig-cnetplot-custom-legend
d <- data.frame(type=c('upregulated', 'downregulated'), x=0, y=0)
g <- p + geom_point(aes(alpha=type, x=x, y=y), data=d, shape=16, size=0)

g + guides(alpha=guide_legend(
    override.aes=list(color=c("blue", "red"),
                      alpha=1,
                      size=3),
    title = "VIP genes",
    reverse = TRUE
))
```


### Tuning cnetplot labels and edge colors

Besides the size of the category / item nodes (`size_category` / `size_item`)
and the fold-change coloring above, two dedicated knobs are commonly needed:

- `color_edge = "category"` colors each edge by the category (term) it belongs to,
  instead of mapping every edge to a single color. `color_edge` also accepts a
  single color, e.g. `color_edge = "grey"`.
- `geom_cnet_label()` is the label layer behind `cnetplot()` (from `ggtangle`).
  Because `cnetplot()` returns a `ggplot` object, you can add another
  `geom_cnet_label()` layer on top with `size`, `color` and `fontface` to
  fine-tune the label text of a specific node type (e.g. only the category
  labels), which is useful when the last few labels are clipped or you want a
  distinct label style.

```{r}
#| label: fig-cnet-label-tune
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Tuning cnetplot edges and labels.
#| fig-cap: |
#|   **Tuning cnetplot edges and labels.**
#|   Default single-color edges (A), edges colored by category with category
#|   labels re-styled via `geom_cnet_label()` (B).
p1 <- cnetplot(edox, node_label = "none", showCategory = 4)
p2 <- cnetplot(edox, node_label = "none", showCategory = 4, color_edge = "category") +
    geom_cnet_label(node_label = "category", size = 5, color = "firebrick", fontface = "bold")
plot_list(p1, p2, ncol = 2, tag_levels = "A")
```

#### Node and edge colors

The default edge color (a red/green pair on gene-ID plots in some releases) can be
replaced either with a single color (`color_edge = "grey55"`) or, when you want
the edges colored separately for each term and shown in the legend, with
`color_edge = "category"`. Category node colors are set with
`color_category`, and — unlike some older versions — setting `foldChange` for the
gene nodes no longer overrides `color_category`: the term nodes keep their
`color_category` fill while the gene nodes get the fold-change palette.

```{r}
#| label: fig-cnet-node-edge-color
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Node and edge color customization in cnetplot.
#| fig-cap: |
#|   **Node and edge color customization in cnetplot.**
#|   Edges in a single color with custom category nodes (A), edges colored by
#|   category (B), and gene nodes re-colored via `scale_color_gradient()` (C).
p1 <- cnetplot(edox, node_label = "none", color_edge = "grey55",
               color_category = "steelblue")
p2 <- cnetplot(edox, node_label = "none", color_edge = "category")
p3 <- cnetplot(edox, node_label = "category", foldChange = geneList) +
    scale_color_gradient(low = "blue", high = "red")
plot_list(p1, p2, p3, ncol = 3, tag_levels = "A")
```

Because these plots are ordinary `ggplot` objects, any `ggplot2` scale can be
plugged on top to re-map the aesthetics — `scale_color_*()` for gene / label
colors and `scale_fill_*()` for node fills. Match the scale to the aesthetic:
`foldChange` puts a **continuous** value (the fold change) on the colour
aesthetic, so re-colour it with a continuous scale such as
`scale_color_gradient()` / `scale_color_gradient2()`; a discrete
`scale_color_manual()` would fail with *"Continuous value supplied to a discrete
scale"*. For `emapplot()`, node colors are
mapped to the fill aesthetic, so use `scale_fill_manual(values = ...)` (or
`scale_fill_gradient*()`) there to customize the node colors.

Note: The `cnetplot()` function also works with [`compareCluster()` output](comparecluster-visualization.qmd#compare-cnetplot).

## Heatmap-like functional classification


The `heatplot` is similar to `cnetplot`, but displays the relationships as a
heatmap. The gene-concept network may become too complicated if users want to
show a large number of significant terms. The `heatplot` can simplify the result
and make it easier to identify expression patterns.

```{r}
#| label: fig-Heatplot  
#| fig-width: 12
#| fig-height: 5
#| fig-scap: |
#|   Heatmap plot of enriched terms.
#| fig-cap: |
#|   **Heatmap plot of enriched terms.** 
#|   default (A), `foldChange=geneList` (B)

p1 <- heatplot(edox, showCategory=5)
p2 <- heatplot(edox, foldChange=geneList, showCategory=5)
plot_list(p1, p2, ncol=1, tag_levels = 'A')
```

The `showTop` parameter can be used to limit the number of genes displayed in the heatmap. This is particularly useful when dealing with large gene sets where only the top genes based on fold change or significance need to be visualized. For example, `showTop = 20` will display only the top 20 genes in the heatmap.

## Tree plot

The `treeplot()` function performs hierarchical clustering of enriched terms. It relies on the pairwise similarities of the enriched terms calculated by the `pairwise_termsim()` function, which by default uses Jaccard's similarity index (JC). Users can also use semantic similarity values when supported (*e.g.*, [GO](go-semantic-similarity.qmd#GOSemSim), [DO](do-semantic-similarity.qmd#DOSE-semantic-similarity), and [MeSH](mesh-semantic-similarity.qmd#sec-meshes-semantic-similarity)).

The default agglomeration method in `treeplot()` is `ward.D`, and users can specify other methods via the `cluster_method` parameter (*e.g.*, 'average', 'complete', 'median', 'centroid', *etc.*; see also the documentation of the `hclust()` function). The `treeplot()` function will cut the tree into several subtrees (specified by the `nCluster` parameter, default is 5) and label subtrees using high-frequency words. This reduces the complexity of the enriched result and improves user interpretation ability.

For fine-grained control over text appearance, the `fontsize_tiplab` and `fontsize_cladelab` parameters allow users to adjust the font size of tip (leaf) labels and clade labels respectively. For example, `fontsize_tiplab = 8, fontsize_cladelab = 10` will set tip labels to 8pt and clade labels to 10pt.

### Colouring the clades and wrapping labels {#treeplot-labels}

Two things that are easy to get wrong here:

**Colouring the clades.** Pass `group_color` a vector of colours, and it is used for both the clade labels and the highlight bars, so the two stay in sync:

```{r eval=FALSE}
treeplot(edox2, showCategory = 20,
         group_color = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442"))
```

**Wrapping the tip labels.** `label_format` controls the wrap length of the **clade** labels (the word-cloud summaries of each cluster), not the tip labels — passing it a number or a function will not change the terms along the tips:

```{r eval=FALSE}
## wraps the clade labels only
treeplot(edox2, label_format = 15)
```

The tip labels are taken straight from the `Description` of the enrichment result, so to wrap them you have to shorten the descriptions *before* building the tree. `pairwise_termsim()` names the similarity matrix from `Description`, and `treeplot()` labels the tips from that matrix, so wrapping at this stage propagates all the way through:

```{r eval=FALSE}
edox_wrapped <- edox
edox_wrapped@result$Description <- stringr::str_wrap(edox_wrapped@result$Description, width = 30)

treeplot(pairwise_termsim(edox_wrapped), showCategory = 20)
```

`str_wrap()` inserts newlines at word boundaries, which the labels render as line breaks. Note that this rewrites the term descriptions of the object, so do it on a copy if you still need the original wording elsewhere.


```{r}
#| label: fig-treeplot  
#| fig-width: 20
#| fig-height: 10
#| fig-scap: |
#|   Tree plot of enriched terms.
#| fig-cap: |
#|   **Tree plot of enriched terms.** 
#|   default (A), `hclust_method = "average"` (B)
library(ggtree)

edox2 <- pairwise_termsim(edox)
p1 <- treeplot(edox2, cladelab_offset=8, tiplab_offset=.3, fontsize_cladelab =5) + 
    hexpand(.2)
p2 <- treeplot(edox2, cluster_method = "average", 
            cladelab_offset=14, tiplab_offset=.3, fontsize_cladelab =5) + 
    hexpand(.3)
aplot::plot_list(p1, p2, tag_levels='A', ncol=2)
```

## Semantic Space Plot {#ssplot}

While `treeplot()` visualizes semantic similarity using a hierarchical structure, the `ssplot()` (Semantic Space Plot) projects enriched terms into a low-dimensional space (e.g., using Multidimensional Scaling, MDS). This provides a complementary spatial view where terms with high semantic similarity are clustered together.

Like `treeplot()`, `ssplot()` requires `pairwise_termsim()` to be run first. It automatically groups terms into clusters (default `nCluster` is determined automatically) and labels them with representative words.

```{r}
#| label: fig-ssplot
#| fig-width: 15
#| fig-height: 10
#| fig-scap: |
#|   Semantic space plot of enriched terms.
#| fig-cap: |
#|   **Semantic space plot of enriched terms.**
#|   ssplot with nCluster=5.

ssplot(edox2, nCluster=5) + ggtitle("ssplot (nCluster=5)")
```

`ssplot()` is particularly useful for visualizing the overall semantic landscape of enrichment results and identifying distinct functional modules in a continuous space.


## Enrichment Map


Enrichment map organizes enriched terms into a network with edges connecting
overlapping gene sets. In this way, mutually overlapping gene sets are tend to
cluster together, making it easy to identify functional module.

### Handling GO term redundancy

GO annotations often contain redundant terms that can dominate enrichment results, potentially obscuring other biological stories. The `simplify()` function from the `clusterProfiler` package uses semantic similarity (via the GOSemSim package) to remove redundant GO terms, providing a clearer view of distinct functional modules.

```{r}
#| label: fig-simplify-go
#| fig-format: png
#| fig-dpi: 150
# Load required packages
library(clusterProfiler)
library(enrichplot)
library(DOSE)

# Prepare example data
data(geneList, package="DOSE")
de <- names(geneList)[abs(geneList) > 2]

# Perform GO enrichment analysis
ego <- enrichGO(de, OrgDb = "org.Hs.eg.db", ont="BP", readable=TRUE)

# Remove redundant GO terms using simplify()
ego_simplified <- simplify(ego, cutoff=0.7, by="p.adjust", select_fun=min)

# Visualize both original and simplified results
ego <- pairwise_termsim(ego)
ego_simplified <- pairwise_termsim(ego_simplified)

p1 <- emapplot(ego, node_label_size=.8, size_edge=.5) + 
    scale_fill_continuous(low = "#e06663", high = "#327eba", name = "p.adjust",
          guide = guide_colorbar(reverse = TRUE, order=1), trans='log10') +
    ggtitle("Original GO terms")

p2 <- emapplot(ego_simplified, node_label_size=.8, size_edge=.5) + 
    scale_fill_continuous(low = "#e06663", high = "#327eba", name = "p.adjust",
                guide = guide_colorbar(reverse = TRUE, order=1), trans='log10') +
    ggtitle("After removing redundant terms")

# Combine plots
library(patchwork)
p1 + p2 + plot_layout(ncol = 2)
```

The `simplify()` function removes redundant GO terms based on semantic similarity (default cutoff = 0.7). This reveals distinct functional modules that might be obscured by redundant terms in the original enrichment results. The `pairwise_termsim()` function calculates pairwise similarities between terms, which is required for `emapplot()` visualization.

The `emapplot` function supports results obtained from hypergeometric test and gene set enrichment analysis. 
The `size_category` parameter can be used to resize nodes and the `layout` parameter can adjust the layout, as demonstrated in @fig-Enrichment.



```{r}
#| label: fig-Enrichment  
#| fig-width: 15
#| fig-height: 12
#| out-width: 95%
#| fig-scap: |
#|   The `emapplot`.
#| fig-cap: |
#|   **The `emapplot`.**
#|   default (A), `node_label="none"` (B), `size_category=1.5` (C), and `layout="with_fr"` (D)
edo <- pairwise_termsim(edo)
p1 <- emapplot(edo) # node_label = "category" (default)
p2 <- emapplot(edo, node_label = "none") 
p3 <- emapplot(edo, node_label = "none", size_category=1.5)
p4 <- emapplot(edo, node_label = "none", layout="with_fr")

plot_list(p1, p2, p3, p4,
        ncol=2, tag_levels = 'A', 
        design="AAAAAA\nBBCCDD", 
        heights = c(1, .3))
```

The `node_label` parameter controls how the labels were displayed. The enriched terms will be displayed by default with `node_label="category"` and it can be disabled by setting `node_label="none"`.

The `node_label_size` parameter allows users to adjust the font size of node labels in the enrichment map. This is particularly useful when dealing with many overlapping terms or when labels need to be more readable. For example, `node_label_size = 3` will increase the label size compared to the default.

If `node_label="group"`, the `emapplot` function will cluster the enriched terms into different groups and only group names (determined by wordcloud) will be displayed. If `node_label="all"`, then the enriched terms and group names will be displayed simultaneously.


```{r}
#| label: fig-Enrichment-node-label  
#| fig-width: 16
#| fig-height: 21
#| fig-scap: |
#|   `emapplot` with enriched terms clustering.
#| fig-cap: |
#|   **`emapplot` with enriched terms clustering.**
#|   `node_label="group"` (A) and 
#|   `node_label="all"` (B).
p5 <- emapplot(edo, node_label = "group")
p6 <- emapplot(edo, node_label = "all")

plot_list(p5, p6, 
        ncol=1, tag_levels = 'A')
```



## UpSet Plot


The `upsetplot` is an alternative to `cnetplot` for visualizing the complex
association between genes and gene sets. It emphasizes the gene overlapping
among different gene sets.


```{r}
#| label: fig-upsetORA  
#| fig-width: 12
#| fig-height: 5
#| fig-scap: |
#|   Upsetplot for over-representation analysis.
#| fig-cap: |
#|   **Upsetplot for over-representation analysis.** 
upsetplot(edo)
```

For over-representation analysis, `upsetplot` will calculate the overlaps among different gene sets as demonstrated in @fig-upsetORA. For GSEA result, it will plot the fold change distributions of different categories (e.g. unique to pathway, overlaps among different pathways).




```{r}
#| label: fig-upsetGSEA  
#| fig-width: 12
#| fig-height: 5
#| fig-scap: |
#|   Upsetplot for gene set enrichment analysis.
#| fig-cap: |
#|   **Upsetplot for gene set enrichment analysis.** 
# Reuse the local DOSE GSEA result so this visualization does not require
# the KEGG REST API during the book build.
kk2 <- edo2
upsetplot(kk2) 
```

## Next steps

- For GSEA-specific distributions, see [GSEA views](enrichplot-gsea.qmd#sec-visualization-gsea).
- For biological reporting, continue to [evidence-guided interpretation](interpretation.qmd#evidence-guided-interpretation).

