# GSEA and distribution views {#sec-visualization-gsea}

| Aspect | GSEA views |
|---|---|
| **Question** | Where is the enrichment signal located, how is it distributed, and which genes drive it? |
| **Input** | A `gseaResult` or compatible ranked result with the original named statistic vector and core-enrichment information. |
| **Functions** | `gseaplot()`, `gseaplot2()`, `ridgeplot()`, `gsearank()`, and distribution/label helpers. |
| **Output** | Enrichment curves, rank views, ridgelines, and gene-level distribution figures. |
| **Limitations** | Direction depends on ranking sign; curve shape is not an effect-size estimate by itself. Record ranking construction, gene-set size filters, and database release. |

Use these figures after checking the ranking and leading-edge genes; continue to [result reading](interpretation-reading.qmd#how-to-read-enrichment) before writing a pathway narrative.


```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
suppressPackageStartupMessages({
    library(clusterProfiler)
    library(enrichplot)
    library(dplyr)
})
```

```{r}
library(DOSE)
data(geneList, package = "DOSE")
de <- names(geneList)[abs(geneList) > 2]
edo <- enrichDO(de)
edo2 <- gseDO(geneList)
```

## ridgeline plot for expression distribution of GSEA result

The `ridgeplot` will visualize expression distributions of core enriched genes
for GSEA enriched categories. It helps users to interpret up/down-regulated pathways.

The `stat` parameter controls the geometry used to summarize the distributions. By default, `ridgeplot()` uses `stat = "density_ridges"`, and users can also set `stat = "binline"` to display binned profiles instead.


```{r}
#| label: fig-ridgeplot  
#| fig-width: 12
#| fig-height: 12
#| fig-scap: |
#|   Ridgeplot for gene set enrichment analysis.
#| fig-cap: |
#|   **Ridgeplot for gene set enrichment analysis.** 
ridgeplot(edo2)
```

```{r eval=FALSE}
ridgeplot(edo2, stat = "binline")
```


## running score and preranked list of GSEA result

Running score and preranked list are traditional methods for visualizing GSEA
result. The `r Biocpkg("enrichplot")` package supports both of them to visualize
the distribution of the gene set and the enrichment score.


```{r}
#| label: fig-gseaplot  
#| fig-format: png
#| fig-dpi: 150
#| fig-width: 12
#| fig-height: 16
#| fig-scap: |
#|   gseaplot for GSEA result(`by = "runningScore"`).
#| fig-cap: |
#|   **gseaplot for GSEA result(`by = "runningScore"`).** 
#|   `by = "runningScore"` (A), `by = "preranked"` (B), default (C)
p1 <- gseaplot(edo2, geneSetID = 1, by = "runningScore", title = edo2$Description[1])
p2 <- gseaplot(edo2, geneSetID = 1, by = "preranked", title = edo2$Description[1])
p3 <- gseaplot(edo2, geneSetID = 1, title = edo2$Description[1])
plot_list(p1, p2, p3, ncol=1, tag_levels='A')
```


The `gseaplot` function also allows users to customize the colors of the running score line and the rank line.

```{r}
#| label: fig-gseaplot-custom-color
#| fig-format: png
#| fig-dpi: 150
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **gseaplot with custom colors.**
#|   (A) Default color scheme.
#|   (B) Customized colors using `color`, `color.line`, and `color.vline`.
p_default <- gseaplot(edo2, geneSetID = 1, title = edo2$Description[1])
p_custom <- gseaplot(edo2, geneSetID = 1, color="#DAB546", color.line='firebrick', 
                    color.vline="steelblue", title = edo2$Description[1])
plot_list(p_default, p_custom, ncol=2, tag_levels='A')
```



## Running-score and ranked-list views

Another method to plot GSEA result is the `gseaplot2` function:

::: {.callout-note}
**Reading the three panels.** `gseaplot2()` stacks three panels that share the same x-axis, which is the *position in the ranked gene list* (not a gene identifier):

* **Running Enrichment Score** — the running-sum statistic. It rises when a gene in the set is encountered and falls otherwise; the enrichment score is its maximum deviation from zero (marked by the dashed vertical line).
* **The middle band** — the black vertical ticks are the positions of the genes in the set. The coloured rectangles behind them divide the *hits* (not the ranked list) into equal-count bins, running from the top of the list (red) to the bottom (blue); their widths therefore encode local hit density, with narrow rectangles indicating a region where the set's genes are packed tightly, and the final rectangle extending to the end of the list. A gene set enriched at the top shows narrow red rectangles crowded on the left.
* **Ranked List Metric** — the ranking statistic itself (e.g. the log fold change or the signed correlation), with the gene set members drawn as vertical segments.

::: {.callout-warning}
**The coloured rectangles used to be wrong — the bug is fixed in this build.** Earlier versions computed the hit bins from a cumulative count taken from the *bottom* of the ranked list and never reversed it, so the gradient came out backwards: for a strongly enriched gene set nearly all hits collapsed into a single rectangle and the rest were empty. That is the bug behind [enrichplot#20](https://github.com/YuLab-SMU/enrichplot/issues/20) and [enrichplot#221](https://github.com/YuLab-SMU/enrichplot/issues/221). It is fixed in the development version of `enrichplot`, which is the version this book builds against: measured on a gene set with 40 hits in a 1000-gene list, all 9 rectangles are populated and the largest holds 15% of the hits, instead of one rectangle holding all of them. **On an older release the `master` branch still carries the old order**, so there the red/blue split is not a faithful picture of where the hits are; if the rectangles look collapsed, that is why. The coloured rectangles are in any case a property of `gseaplot2()`, not of the GSEA method — the original Broad GSEA plots colour the ranked-list-metric panel by the *sign* of the metric instead, so the two conventions do not agree either.
:::
:::


```{r}
#| label: fig-gseaplot2  
#| fig-format: png
#| fig-dpi: 150
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result.
#| fig-cap: |
#|   **Gseaplot2 for GSEA result.** 
gseaplot2(edo2, geneSetID = 1, title = edo2$Description[1])
```

The `gseaplot2` also supports multile gene sets to be displayed on the same figure:


```{r}
#| label: fig-gseaplot22  
#| fig-format: png
#| fig-dpi: 150
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result of multile gene sets.
#| fig-cap: |
#|   **Gseaplot2 for GSEA result of multile gene sets.** 
gseaplot2(edo2, geneSetID = 1:3)
```

User can also displaying the pvalue table on the plot via `pvalue_table`
parameter:


```{r}
#| label: fig-gseaplot23  
#| fig-format: png
#| fig-dpi: 150
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result of multile gene sets(add pvalue_table).
#| fig-cap: |
#|   **Gseaplot2 for GSEA result of multile gene sets(add pvalue_table).** 
gseaplot2(edo2, geneSetID = 1:3, pvalue_table = TRUE,
          color = c("#E495A5", "#86B875", "#7DB0DD"), ES_geom = "dot")
```


The `pvalue_table` can be customized using the following parameters:

- `pvalue_table_rownames` to specify row names of the table (if NULL, no row names will be displayed)
- `pvalue_table_columns` to specify column names of the table


```{r}
#| label: fig-gseaplot-pvalue  
#| fig-format: png
#| fig-dpi: 150
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 with customized pvalue table.
#| fig-cap: |
#|   **Gseaplot2 with customized pvalue table.** 
#| 
gseaplot2(edo2, geneSetID = 1, pvalue_table = TRUE,
          pvalue_table_rownames = NULL,
          pvalue_table_columns = c("ID", "NES", "p.adjust"))
```


User can specify `subplots` to only display a subset of plots:


```{r}
#| label: fig-gseaplot24
#| fig-format: png
#| fig-dpi: 150
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result of multile gene sets(add subplots).
#| fig-cap: |
#|   **Gseaplot2 for GSEA result of multile gene sets(add subplots).** 
#|   `subplots = 1` (A),`subplots = 1:2` (B)
#| 
p1 <- gseaplot2(edo2, geneSetID = 1:3, subplots = 1)
p2 <- gseaplot2(edo2, geneSetID = 1:3, subplots = 1:2)
plot_list(p1, p2, ncol=1, tag_levels = 'A')
```


### Labeling genes in GSEA plot

Users can use `geom_gsea_gene()` to label specific genes in the GSEA plot.

```{r}
#| label: fig-geom-gsea-gene
#| fig-format: png
#| fig-dpi: 150
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Labeling genes in GSEA plot.**
library(ggplot2)
library(ggrepel)

# Get gene set ID
id <- edo2$ID[1]

# Randomly select genes to label
set.seed(123)
genes <- sample(edo2[[id]], 5)

# Label genes on gseaplot2
p <- gseaplot2(edo2, geneSetID = 1, title = edo2$Description[1])

# Add geom_gsea_gene layer to the first subplot (running score)
p[[1]] <- p[[1]] + geom_gsea_gene(genes, geom=geom_label)
p
```

If users prefer to label with gene symbols, they can convert the gene IDs to symbols (e.g., using `setReadable`) before plotting.

```{r}
#| label: fig-geom-gsea-gene-symbol
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Labeling gene symbols in GSEA plot.**
#| eval: false

library(clusterProfiler)
# Assuming org.Hs.eg.db is available
if (require("org.Hs.eg.db")) {
    edo2_symbol <- setReadable(edo2, 'org.Hs.eg.db', 'ENTREZID')
    id <- edo2_symbol$ID[1]
    genes_symbol <- sample(edo2_symbol[[id]], 5)
    
    p_symbol <- gseaplot2(edo2_symbol, geneSetID = 1, title = edo2_symbol$Description[1])
    p_symbol[[1]] <- p_symbol[[1]] + geom_gsea_gene(genes_symbol, geom=geom_text_repel)
    p_symbol
}
```




The `gsearank` function plot the ranked list of genes belong to the specific
gene set.


```{r}
#| label: fig-gsearank
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Ranked list of genes belong to the specific gene set.
#| fig-cap: |
#|   **Ranked list of genes belong to the specific gene set.**
gsearank(edo2, 1, title = edo2[1, "Description"])
```

Multiple gene sets can be aligned using `cowplot`:

```{r}
#| label: fig-gsearank2
#| fig-width: 8
#| fig-height: 12
#| fig-scap: |
#|   Gsearank for multiple gene sets.
#| fig-cap: |
#|   **Gsearank for multiple gene sets.**
#| 
library(ggplot2)

pp <- lapply(1:3, function(i) {
    anno <- edo2[i, c("NES", "pvalue", "p.adjust")]
    lab <- paste0(names(anno), "=",  round(anno, 3), collapse="\n")

    gsearank(edo2, i, edo2[i, 2]) + xlab(NULL) +ylab(NULL) +
        annotate("text", 10000, edo2[i, "enrichmentScore"] * .75, label = lab, hjust=0, vjust=0)
})
plot_list(gglist=pp, ncol=1)
```

### Extracting data from gsearank plot

The `gsearank` function can also output the data used for plotting by setting `output = "table"`. This allows users to inspect the running score and other metrics for genes in the gene set.

```{r}
#| label: tbl-gsearank
#| tbl-cap: |
#|   **Data extracted from gsearank plot.**
gsearank(edo2, 1, output = "table") |> head()
```

Users can also merge this table with gene information (e.g. Symbol) using `bitr` or other methods.

```{r}
#| label: tbl-gsearank-symbol
#| tbl-cap: |
#|   **Data extracted from gsearank plot with gene symbols.**
#| eval: false

rank_table <- gsearank(edo2, 1, output = "table")
# Assuming 'gene' column contains Entrez IDs
if (require("org.Hs.eg.db") && require("clusterProfiler")) {
    gene_info <- bitr(rank_table$gene, fromType="ENTREZID", toType="SYMBOL", OrgDb="org.Hs.eg.db")
    rank_table_symbol <- merge(rank_table, gene_info, by.x="gene", by.y="ENTREZID")
    head(rank_table_symbol)
}
```

## Next steps

- For common term summaries, see [summary plots](enrichplot-summary.qmd#sec-visualization-summary).
- For leading-edge interpretation, see [the recipes](leading-edge-nonmodel.qmd#recipes).

