# Visualization of functional enrichment result {#sec-enrichplot}


```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
library(clusterProfiler)
library(enrichplot)
library(dplyr)
```



The `r Biocpkg("enrichplot")` package implements several visualization methods to help interpret enrichment results. It supports visualizing enrichment results obtained from `r Biocpkg("DOSE")` [@yu_dose_2015], `r Biocpkg("clusterProfiler")` [@yu2012; @wu_clusterprofiler_2021],
`r Biocpkg("ReactomePA")` [@yu_reactomepa_2016] and `r Biocpkg("meshes")` [@yu_meshes_2018]. Both over representation analysis (ORA) and gene set enrichment analysis (GSEA) are supported.

Note: Several visualization methods were first implemented in `r Biocpkg("DOSE")` and rewrote from scratch using `r CRANpkg("ggplot2")`. If you want to use the [old methods](https://www.biostars.org/p/375555), you can use the [doseplot](https://github.com/GuangchuangYu/doseplot) package.


## Bar Plot


Bar plot is the most widely used method to visualize enriched terms. It depicts the enrichment scores (*e.g.* p values) and gene count or ratio as bar height
and color @fig-Barplot. Users can specify the number of terms (most significant) or selected terms (see also the [FAQ](#showing-specific-pathways)) to display via the `showCategory` parameter. 



```{r}
#| label: fig-enrichDGN
library(DOSE)
data(geneList)
de <- names(geneList)[abs(geneList) > 2]

edo <- enrichDGN(de)
```

```{r eval=F}
#| label: fig-barplot-eval-false
library(enrichplot)
barplot(edo, showCategory=20) 
```

Other variables that derived using `mutate` can also be used as bar height or color as demonstrated in @sec-mutate and @fig-Barplot.

```{r eval=F}
#| label: fig-mutate-barplot-eval-false
library(clusterProfiler)

mutate(edo, qscore = -log(p.adjust, base=10)) |> 
    barplot(x="qscore")
```



```{r}
#| label: fig-Barplot  
#| echo: false
#| fig-width: 11
#| fig-height: 6
#| fig-scap: |
#|   Bar plot of enriched terms.
#| fig-cap: |
#|   **Bar plot of enriched terms.**

p1 <- barplot(edo, showCategory=20) 
p2 <- mutate(edo, qscore = -log(p.adjust, base=10)) |>
    barplot(x="qscore")

library(aplot)
plot_list(p1, p2, ncol=2, tag_levels='A')    
```

### Split top categories by ontology

When visualizing GO enrichment across all ontologies (`ont = "ALL"`), it is often more informative to show the top terms within each ontology (BP/CC/MF) rather than selecting the top terms globally. `barplot()` forwards additional arguments via `...`, and one useful option is `split`. Setting `split = "ONTOLOGY"` will perform the “top N per ontology” selection and makes it straightforward to color (and later facet) the results by the ontology.

```{r}
#| label: fig-enrichGO-ontology
#| echo: false
#| message: false
#| warning: false
library(DOSE)
data(geneList, package = "DOSE")
de <- names(geneList)[1:300]

library(clusterProfiler)
library(org.Hs.eg.db)
ego_all <- enrichGO(
  gene = de,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "ALL",
  readable = TRUE
)
```

```{r}
#| label: fig-barplot-ontology-nofacet
#| fig-width: 12
#| fig-height: 9
#| fig-scap: |
#|   Bar plot of GO terms using `split = "ONTOLOGY"` without faceting.
#| fig-cap: |
#|   **Bar plot of GO terms using `split = "ONTOLOGY"` (no faceting).**
library(enrichplot)
library(ggplot2)

p_nofacet <- barplot(
  ego_all,
  x = "Count",
  showCategory = 15,
  split = "ONTOLOGY"
) +
  aes(fill = ONTOLOGY) +
  scale_fill_brewer(palette = "Set2")

p_nofacet +
  geom_text(aes(label = Count), nudge_x = 2) +
  scale_y_discrete()
```

Note that `barplot()` wraps long term descriptions onto multiple lines by default (controlled by `label_format`, default is 30). If you prefer to keep the y-axis labels on a single line, you can override the default scale by adding `+ scale_y_discrete()` as shown above.

```{r}
#| label: fig-barplot-ontology-facet
#| fig-width: 12
#| fig-height: 9
#| fig-scap: |
#|   Bar plot of GO terms with `autofacet()` splitting by ontology.
#| fig-cap: |
#|   **Bar plot of GO terms with `autofacet()` splitting by ontology.**
p_facet <- barplot(
  ego_all,
  x = "Count",
  showCategory = 15,
  split = "ONTOLOGY"
) +
  aes(fill = ONTOLOGY) +
  scale_fill_brewer(palette = "Set2") +
  enrichplot::autofacet(by = "row", scales = "free")

p_facet + scale_y_discrete()
```

If the fortified data contain an `ONTOLOGY` column (as in the output generated by `split = "ONTOLOGY"`), you can facet the plot directly with `+ enrichplot::autofacet()`, and it will automatically split panels by ontology.


## Dot plot


Dot plot is similar to bar plot with the capability to encode another score as dot size.

```{r}
#| label: fig-dotplot-setup
library(ggplot2)
edo2 <- gseDO(geneList)
dotplot(edo, showCategory=30, label_format=NULL) + ggtitle("dotplot for ORA")
dotplot(edo2, showCategory=30, label_format=NULL) + ggtitle("dotplot for GSEA")
```


```{r}
#| label: fig-Dotplotcap  
#| echo: false
#| fig-width: 12
#| fig-height: 10
#| fig-scap: |
#|   Dot plot of enriched terms.
#| fig-cap: |
#|   **Dot plot of enriched terms.**
library(ggplot2)

edo2 <- gseDO(geneList)
p1 <- dotplot(edo, showCategory=30) + ggtitle("dotplot for ORA") 
p2 <- dotplot(edo2, showCategory=30) + ggtitle("dotplot for GSEA") 
plot_list(p1, p2, ncol=2, tag_levels = 'A')
```


Note: The `dotplot()` function also works with [`compareCluster()` output](#compare-dotplot).

### Highlighting specific pathways

The `label_format` parameter in `dotplot()` accepts a function to format y-axis labels, which can be used to highlight specific pathways. This is useful for emphasizing pathways of interest in publication figures.

First, perform enrichment analysis:

```{r}
#| label: fig-enrichDO
library(DOSE)
data(geneList)
de <- names(geneList)[1:100]
x <- enrichDO(de)
```

Define a function to highlight pathways containing "cancer":

```{r}
#| label: fig-label-format-func
f <- function(ids) {
    i <- grepl('cancer', ids)
    ids[i] <- paste0("<i style='color:#009E73'>", ids[i], "</i>")
    return(ids)
}
```

To render the HTML formatting, use the `ggtext` package with `element_markdown()`:

```{r}
#| label: fig-dotplot-highlight
library(ggplot2)
library(ggtext)
library(enrichplot)
dotplot(x, label_format=f) + theme(axis.text.y = element_markdown())
```

You can also highlight specific pathways by index:

```{r}
#| label: fig-label-format-func2
f2 <- function(ids) {
    ids[1] <- paste0("<i style='color:#009E73'>", ids[1], "</i>")
    ids[2] <- paste0("<i style='color:#0072B2'>**", ids[2], "**</i>")
    return(ids)
}

dotplot(x, label_format=f2) + theme(axis.text.y = element_markdown())
```

Or apply different colors to all pathways:

```{r}
#| label: fig-label-format-func3
f3 <- function(ids) {
    # cols <- rcartocolor::carto_pal(length(ids), "Vivid")
    # cols <- colorspace::rainbow_hcl(length(ids))
    cols <- rainbow(length(ids))
    ids <- paste0("<i style='color:", cols, "'>**", ids, "**</i>")
    return(ids)
}

dotplot(x, label_format=f3) + theme(axis.text.y = element_markdown())
```

### Formula interface of dotplot

The `x` variable of `dotplot()` supports a formula interface, allowing users to use derived variables. For example, we can calculate `GeneRatio/BgRatio` or `-log(p.adjust)` and use them as the x-axis variable.

```{r}
#| label: fig-dotplot-formula
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Dotplot with formula interface.**
#|   (A) `x = ~GeneRatio/BgRatio`.
#|   (B) `x = ~ -log(p.adjust)`.
p1 <- dotplot(edo, x = ~GeneRatio/BgRatio)
p2 <- dotplot(edo, x = ~ -log(p.adjust))
plot_list(p1, p2, ncol=2, tag_levels='A')
```

This feature provides great flexibility in visualization. For instance, the **Rich Factor** is a common metric defined as the ratio of the number of differentially expressed genes annotated in a pathway to the total number of genes annotated in that pathway.

$$Rich Factor = \frac{Count}{BgRatio \times N}$$

where $N$ is the total number of genes in the background distribution.

We can easily visualize the **Rich Factor** using the formula interface.

```{r eval=FALSE}
#| label: fig-rich-factor
N <- 8007 # total number of genes in the background
dotplot(edo, x = ~Count/(BgRatio * N))
```

Alternatively, you can precompute derived variables with `mutate()` (e.g., add `neglog10p = -log10(p.adjust)` or `richFactor = Count / (BgRatio * N)`) and then pass the new column name to `x`. This approach is useful when you want to reuse the computed variables across multiple plots. See the dedicated section on using dplyr verbs with enrichment results in @sec-mutate.

### Adjusting dot sizes

To increase the size of the dots in the dot plot, we can use the `scale_size` function from the `ggplot2` package. This allows us to specify the range of the dot sizes.

```{r}
#| label: fig-dotplot-size
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Dotplot with adjusted dot sizes.**
#|   (A) Default dotplot.
#|   (B) Dotplot with `scale_size(range=c(2, 20))`.
p1 <- dotplot(edo, showCategory=10) + ggtitle("Default")
p2 <- dotplot(edo, showCategory=10) + scale_size(range=c(2, 20)) + ggtitle("Adjusted size")
plot_list(p1, p2, ncol=2, tag_levels='A')
```

## Manhattan plot

The `manhattanplot()` function provides a landscape-style overview of enrichment results. It arranges enriched terms along the x-axis by ontology (or by category in `compareCluster()` results), uses `-log10()` transformed significance values on the y-axis, and maps term size to a variable such as `Count`. The most significant terms can be labeled automatically with `showCategory`, making it useful for summarizing a large number of enriched terms in a single plot.

When GO enrichment is performed with `ont = "ALL"`, `manhattanplot()` can display the top terms across BP, CC, and MF simultaneously by combining it with `split = "ONTOLOGY"`.

```{r}
#| label: fig-manhattanplot
#| fig-width: 12
#| fig-height: 7
#| fig-cap: |
#|   **Manhattan plot of GO enrichment results.**
#|   Terms are grouped by ontology and plotted by `-log10(p.adjust)`.
manhattanplot(
  ego_all,
  color = "p.adjust",
  showCategory = 12,
  size = "Count",
  split = "ONTOLOGY",
  title = "GO enrichment landscape"
)
```

The `manhattanplot()` function also supports `gseaResult`, `enrichResultList`, `gseaResultList`, and `compareClusterResult` objects, which makes it convenient for comparing the distribution of significant terms across multiple ontologies or experimental groups.

## Dotplot2: Comparing two clusters

The `dotplot2()` function is a specialized version of dotplot designed for comparing two selected clusters from `compareCluster` results. This function provides focused visualization when users want to directly contrast specific biological conditions or experimental groups.

```{r}
#| label: fig-dotplot2
#| fig-width: 12
#| fig-height: 8
#| eval: false
#| fig-cap: |
#|   **Dotplot2 for comparing two clusters.**
#|   Direct comparison of cluster A and cluster B.

# Example using compareCluster result
library(clusterProfiler)
data(gcSample)
xx <- compareCluster(gcSample, fun="enrichKEGG",
                     organism="hsa", pvalueCutoff=0.05)

# Compare cluster A and cluster B
dotplot2(xx, vars = c("A", "B"))
```

The `dotplot2()` function accepts the same parameters as `dotplot()` but requires `cluster1` and `cluster2` arguments to specify which clusters to compare. This focused visualization helps in identifying differential enrichment patterns between specific experimental conditions.

## Gene-Concept Network {#cnetplot}

```{r}
#| label: fig-cnet-setup
#| echo: false
#| message: false
#| warning: false
library(clusterProfiler)
library(DOSE)
library(enrichplot)
library(org.Hs.eg.db)
data(geneList, package="DOSE")
de <- names(geneList)[abs(geneList) > 2]
edo <- enrichDGN(de)
```

Both the `barplot()` and `dotplot()` only displayed most significant or selected enriched terms,
while users may want to know which genes are involved in these significant
terms. 
To consider the potential biological complexities in which a gene may belong to multiple annotation categories and provide information about numeric changes when available, we developed the `cnetplot()` function to extract the complex associations. 
The `cnetplot()` depicts the linkages of genes and biological concepts (*e.g.* GO terms or KEGG pathways) as a network. GSEA result is also supported
with only core enriched genes displayed. For GSEA results, `cnetplot` specifically visualizes the core enriched genes identified through leading edge analysis, which represent the subset of genes most responsible for driving the enrichment signal.


Users can use the 'fc_threshold' parameter in the `cnetplot` function to filter genes to only include those with `|foldChange| > fc_threshold`. This allows users to plot only high fold-change genes (e.g., `fc_threshold = 1`). In addition, users can use the `node_label_gene` parameter to label genes that are shared between groups of terms (i.e. `node_label_gene = "share"`).

```{r}
#| label: fig-Networkplot  
#| fig-width: 16
#| fig-height: 18
#| fig-scap: |
#|   Network plot of enriched terms.
#| fig-cap: |
#|   **Network plot of enriched terms.**
#| 
## convert gene ID to Symbol
edox <- setReadable(edo, 'org.Hs.eg.db', 'ENTREZID')
p1 <- cnetplot(edox, foldChange=geneList)
p2 <- cnetplot(edox, categorySizeBy=~ -log10(pvalue), foldChange=geneList)
## only plot high fold-change genes
p3 <- cnetplot(edox, foldChange=geneList, fc_threshold = 2)
p4 <- cnetplot(edox, foldChange=geneList, node_label = "share")
plot_list(p1, p2, p3, p4, ncol=2, tag_levels = 'A')
```


If you would like label subset of the nodes, you can use the `node_label` parameter, which supports 4 possible selections (i.e. "category", "gene", "all" and "none"), as demonstrated in @fig-cnetNodeLabel.

The `node_label` parameter also supports enhanced functionality:

1. **Vector selection**: Specify specific genes to label using a character vector, e.g., `node_label = c("gene1", "gene2")`
2. **'exclusive' labeling**: Label only genes that belong exclusively to one category using `node_label = "exclusive"`
3. **'share' labeling**: Label genes that are shared between multiple categories using `node_label = "share"` (as shown in the example above)
4. **Conditional filtering**: Use comparison operators to filter genes based on fold change, e.g., `node_label = "> 1"` or `node_label = "< -1"` to label genes with absolute fold change greater than 1 



```{r}
#| label: fig-cnetNodeLabel  
#| fig-width: 12
#| fig-height: 16
#| fig-scap: |
#|   Labelling nodes by selected subset.
#| fig-cap: |
#|   **Labelling nodes by selected subset.** 
#|   gene category (A), gene name (B), 
#|   both gene category and gene name (C, default) 
#|   and not to label at all (D).
p1 <- cnetplot(edox, node_label="category") 
p2 <- cnetplot(edox, node_label="gene") 
p3 <- cnetplot(edox, node_label="all") 
p4 <- cnetplot(edox, node_label="none", 
        color_category='firebrick', 
        color_item='steelblue') 
plot_list(p1, p2, p3, p4, ncol=2, tag_levels = 'A')
```

The `cnetplot` function can be used as a general method to visualize data relationships in a network diagram. Please refer to the vignette of `r CRANpkg("ggtangle")`. 


### Customizing gene colors in cnetplot

Users can color specific genes by providing a named vector of fold changes containing only those genes.

```{r}
#| label: fig-cnetplot-custom-color
foldChange <- c(rep(1, 6), rep(-1, 4))
names(foldChange) <- c("MARCO", "GZMB", "CXCL11", "CXCL10",
                       "LAG3", "CCL8", "PDK1", "GABRP", "MELK", "CENPE")
p <- cnetplot(edox, foldChange=foldChange)
p
```

If you want to remove the color legend:

```{r}
#| label: fig-cnetplot-no-legend
p <- p + guides(color='none')
p
```

To create a custom legend, you can use a dummy data frame and `geom_point`:

```{r}
#| label: fig-cnetplot-custom-legend
d <- data.frame(type=c('upregulated', 'downregulated'), x=0, y=0)
g <- p + geom_point(aes(alpha=type, x=x, y=y), data=d, shape=16, size=0)

g + guides(alpha=guide_legend(
    override.aes=list(color=c("blue", "red"),
                      alpha=1,
                      size=3),
    title = "VIP genes",
    reverse = TRUE
))
```


Note: The `cnetplot()` function also works with [`compareCluster()` output](#compare-cnetplot).


## Heatmap-like functional classification


The `heatplot` is similar to `cnetplot`, but displays the relationships as a
heatmap. The gene-concept network may become too complicated if users want to
show a large number of significant terms. The `heatplot` can simplify the result
and make it easier to identify expression patterns.

```{r}
#| label: fig-Heatplot  
#| fig-width: 12
#| fig-height: 5
#| fig-scap: |
#|   Heatmap plot of enriched terms.
#| fig-cap: |
#|   **Heatmap plot of enriched terms.** 
#|   default (A), `foldChange=geneList` (B)

p1 <- heatplot(edox, showCategory=5)
p2 <- heatplot(edox, foldChange=geneList, showCategory=5)
plot_list(p1, p2, ncol=1, tag_levels = 'A')
```

The `showTop` parameter can be used to limit the number of genes displayed in the heatmap. This is particularly useful when dealing with large gene sets where only the top genes based on fold change or significance need to be visualized. For example, `showTop = 20` will display only the top 20 genes in the heatmap.

## Tree plot

The `treeplot()` function performs hierarchical clustering of enriched terms. It relies on the pairwise similarities of the enriched terms calculated by the `pairwise_termsim()` function, which by default uses Jaccard's similarity index (JC). Users can also use semantic similarity values when supported (*e.g.*, [GO](#GOSemSim), [DO](#DOSE-semantic-similarity), and [MeSH](#meshes-semantic-similarity)).

The default agglomeration method in `treeplot()` is `ward.D`, and users can specify other methods via the `cluster_method` parameter (*e.g.*, 'average', 'complete', 'median', 'centroid', *etc.*; see also the documentation of the `hclust()` function). The `treeplot()` function will cut the tree into several subtrees (specified by the `nCluster` parameter, default is 5) and label subtrees using high-frequency words. This reduces the complexity of the enriched result and improves user interpretation ability.

For fine-grained control over text appearance, the `leave_fontsize` and `clade_fontsize` parameters allow users to adjust the font size of leaf labels and clade labels respectively. For example, `leave_fontsize = 8, clade_fontsize = 10` will set leaf labels to 8pt and clade labels to 10pt.


```{r}
#| label: fig-treeplot  
#| fig-width: 20
#| fig-height: 10
#| fig-scap: |
#|   Tree plot of enriched terms.
#| fig-cap: |
#|   **Tree plot of enriched terms.** 
#|   default (A), `hclust_method = "average"` (B)
library(ggtree)

edox2 <- pairwise_termsim(edox)
p1 <- treeplot(edox2, cladelab_offset=8, tiplab_offset=.3, fontsize_cladelab =5) + 
    hexpand(.2)
p2 <- treeplot(edox2, cluster_method = "average", 
            cladelab_offset=14, tiplab_offset=.3, fontsize_cladelab =5) + 
    hexpand(.3)
aplot::plot_list(p1, p2, tag_levels='A', ncol=2)
```

## Semantic Space Plot {#ssplot}

While `treeplot()` visualizes semantic similarity using a hierarchical structure, the `ssplot()` (Semantic Space Plot) projects enriched terms into a low-dimensional space (e.g., using Multidimensional Scaling, MDS). This provides a complementary spatial view where terms with high semantic similarity are clustered together.

Like `treeplot()`, `ssplot()` requires `pairwise_termsim()` to be run first. It automatically groups terms into clusters (default `nCluster` is determined automatically) and labels them with representative words.

```{r}
#| label: fig-ssplot
#| fig-width: 15
#| fig-height: 10
#| fig-scap: |
#|   Semantic space plot of enriched terms.
#| fig-cap: |
#|   **Semantic space plot of enriched terms.**
#|   ssplot with nCluster=5.

ssplot(edox2, nCluster=5) + ggtitle("ssplot (nCluster=5)")
```

`ssplot()` is particularly useful for visualizing the overall semantic landscape of enrichment results and identifying distinct functional modules in a continuous space.


## Enrichment Map


Enrichment map organizes enriched terms into a network with edges connecting
overlapping gene sets. In this way, mutually overlapping gene sets are tend to
cluster together, making it easy to identify functional module.

### Handling GO term redundancy

GO annotations often contain redundant terms that can dominate enrichment results, potentially obscuring other biological stories. The `simplify()` function from the `clusterProfiler` package uses semantic similarity (via the GOSemSim package) to remove redundant GO terms, providing a clearer view of distinct functional modules.

```{r}
#| label: fig-simplify-go
# Load required packages
library(clusterProfiler)
library(enrichplot)
library(DOSE)

# Prepare example data
data(geneList, package="DOSE")
de <- names(geneList)[abs(geneList) > 2]

# Perform GO enrichment analysis
ego <- enrichGO(de, OrgDb = "org.Hs.eg.db", ont="BP", readable=TRUE)

# Remove redundant GO terms using simplify()
ego_simplified <- simplify(ego, cutoff=0.7, by="p.adjust", select_fun=min)

# Visualize both original and simplified results
ego <- pairwise_termsim(ego)
ego_simplified <- pairwise_termsim(ego_simplified)

p1 <- emapplot(ego, node_label_size=.8, size_edge=.5) + 
    scale_fill_continuous(low = "#e06663", high = "#327eba", name = "p.adjust",
          guide = guide_colorbar(reverse = TRUE, order=1), trans='log10') +
    ggtitle("Original GO terms")

p2 <- emapplot(ego_simplified, node_label_size=.8, size_edge=.5) + 
    scale_fill_continuous(low = "#e06663", high = "#327eba", name = "p.adjust",
                guide = guide_colorbar(reverse = TRUE, order=1), trans='log10') +
    ggtitle("After removing redundant terms")

# Combine plots
library(patchwork)
p1 + p2 + plot_layout(ncol = 2)
```

The `simplify()` function removes redundant GO terms based on semantic similarity (default cutoff = 0.7). This reveals distinct functional modules that might be obscured by redundant terms in the original enrichment results. The `pairwise_termsim()` function calculates pairwise similarities between terms, which is required for `emapplot()` visualization.

The `emapplot` function supports results obtained from hypergeometric test and gene set enrichment analysis. 
The `size_category` parameter can be used to resize nodes and the `layout` parameter can adjust the layout, as demonstrated in @fig-Enrichment.



```{r}
#| label: fig-Enrichment  
#| fig-width: 15
#| fig-height: 12
#| out-width: 95%
#| fig-scap: |
#|   The `emapplot`.
#| fig-cap: |
#|   **The `emapplot`.**
#|   default (A), `node_label="none"` (B), `size_category=1.5` (C), and `layout="with_fr"` (D)
edo <- pairwise_termsim(edo)
p1 <- emapplot(edo) # node_label = "category" (default)
p2 <- emapplot(edo, node_label = "none") 
p3 <- emapplot(edo, node_label = "none", size_category=1.5)
p4 <- emapplot(edo, node_label = "none", layout="with_fr")

plot_list(p1, p2, p3, p4,
        ncol=2, tag_levels = 'A', 
        design="AAAAAA\nBBCCDD", 
        heights = c(1, .3))
```

The `node_label` parameter controls how the labels were displayed. The enriched terms will be displayed by default with `node_label="category"` and it can be disabled by setting `node_label="none"`.

The `node_label_size` parameter allows users to adjust the font size of node labels in the enrichment map. This is particularly useful when dealing with many overlapping terms or when labels need to be more readable. For example, `node_label_size = 3` will increase the label size compared to the default.

If `node_label="group"`, the `emapplot` function will cluster the enriched terms into different groups and only group names (determined by wordcloud) will be displayed. If `node_label="all"`, then the enriched terms and group names will be displayed simultaneously.


```{r}
#| label: fig-Enrichment-node-label  
#| fig-width: 16
#| fig-height: 21
#| fig-scap: |
#|   `emapplot` with enriched terms clustering.
#| fig-cap: |
#|   **`emapplot` with enriched terms clustering.**
#|   `node_label="group"` (A) and 
#|   `node_label="all"` (B).
p5 <- emapplot(edo, node_label = "group")
p6 <- emapplot(edo, node_label = "all")

plot_list(p5, p6, 
        ncol=1, tag_levels = 'A')
```

## Biological theme comparison 

The `emapplot` function also supports results obtained from `compareCluster` function of `clusterProfiler` package. In addition to `size_category` and `layout` parameters, the number of circles in the bottom left corner can be adjusted  using the `legend_n` parameteras, and proportion of clusters in the pie chart can be adjusted using the `pie` parameter, when `pie="count"`, the proportion of  clusters in the pie chart is determined by the number of genes, as demonstrated in @fig-Enrichment2.


```{r}
#| label: fig-Enrichment2  
#| fig-width: 16
#| fig-height: 18
#| fig-scap: |
#|   Plot for results obtained from `compareCluster` function of `clusterProfiler` package.
#| fig-cap: |
#|   **Plot for results obtained from `compareCluster` function of `clusterProfiler` package.** 
#|   default (A), `legend_n=2` (B), `pie="count"` (C) and `pie="count",
#|   size_category=1.5, layout="kk"` (D).
library(clusterProfiler)
data(gcSample)
xx <- compareCluster(gcSample, fun="enrichKEGG",
                     organism="hsa", pvalueCutoff=0.05)
xx <- pairwise_termsim(xx)                     
p1 <- emapplot(xx)
p2 <- emapplot(xx) 
p3 <- emapplot(xx, pie="count")
p4 <- emapplot(xx, pie="count", size_category=1.5, layout="kk")
plot_list(p1, p2, p3, p4, ncol=2, tag_levels = 'A')
```


## UpSet Plot


The `upsetplot` is an alternative to `cnetplot` for visualizing the complex
association between genes and gene sets. It emphasizes the gene overlapping
among different gene sets.


```{r}
#| label: fig-upsetORA  
#| fig-width: 12
#| fig-height: 5
#| fig-scap: |
#|   Upsetplot for over-representation analysis.
#| fig-cap: |
#|   **Upsetplot for over-representation analysis.** 
upsetplot(edo)
```

For over-representation analysis, `upsetplot` will calculate the overlaps among different gene sets as demonstrated in @fig-upsetORA. For GSEA result, it will plot the fold change distributions of different categories (e.g. unique to pathway, overlaps among different pathways).




```{r}
#| label: fig-upsetGSEA  
#| fig-width: 12
#| fig-height: 5
#| fig-scap: |
#|   Upsetplot for gene set enrichment analysis.
#| fig-cap: |
#|   **Upsetplot for gene set enrichment analysis.** 
kk2 <- gseKEGG(geneList     = geneList,
               organism     = 'hsa',
               minGSSize    = 120,
               pvalueCutoff = 0.05,
               verbose      = FALSE)
upsetplot(kk2) 
```


## ridgeline plot for expression distribution of GSEA result

The `ridgeplot` will visualize expression distributions of core enriched genes
for GSEA enriched categories. It helps users to interpret up/down-regulated pathways.

The `stat` parameter controls the geometry used to summarize the distributions. By default, `ridgeplot()` uses `stat = "density_ridges"`, and users can also set `stat = "binline"` to display binned profiles instead.


```{r}
#| label: fig-ridgeplot  
#| fig-width: 12
#| fig-height: 12
#| fig-scap: |
#|   Ridgeplot for gene set enrichment analysis.
#| fig-cap: |
#|   **Ridgeplot for gene set enrichment analysis.** 
ridgeplot(edo2)
```

```{r eval=FALSE}
ridgeplot(edo2, stat = "binline")
```


## running score and preranked list of GSEA result

Running score and preranked list are traditional methods for visualizing GSEA
result. The `r Biocpkg("enrichplot")` package supports both of them to visualize
the distribution of the gene set and the enrichment score.


```{r}
#| label: fig-gseaplot  
#| fig-width: 12
#| fig-height: 16
#| fig-scap: |
#|   gseaplot for GSEA result(`by = "runningScore"`).
#| fig-cap: |
#|   **gseaplot for GSEA result(`by = "runningScore"`).** 
#|   `by = "runningScore"` (A), `by = "preranked"` (B), default (C)
p1 <- gseaplot(edo2, geneSetID = 1, by = "runningScore", title = edo2$Description[1])
p2 <- gseaplot(edo2, geneSetID = 1, by = "preranked", title = edo2$Description[1])
p3 <- gseaplot(edo2, geneSetID = 1, title = edo2$Description[1])
plot_list(p1, p2, p3, ncol=1, tag_levels='A')
```


The `gseaplot` function also allows users to customize the colors of the running score line and the rank line.

```{r}
#| label: fig-gseaplot-custom-color
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **gseaplot with custom colors.**
#|   (A) Default color scheme.
#|   (B) Customized colors using `color`, `color.line`, and `color.vline`.
p_default <- gseaplot(edo2, geneSetID = 1, title = edo2$Description[1])
p_custom <- gseaplot(edo2, geneSetID = 1, color="#DAB546", color.line='firebrick', 
                    color.vline="steelblue", title = edo2$Description[1])
plot_list(p_default, p_custom, ncol=2, tag_levels='A')
```

### Unified color setting with `set_enrichplot_color()`

For consistent color settings across different plot types in `enrichplot`, the `set_enrichplot_color()` function provides a unified approach to customize color mappings. This avoids repetitive color code specifications in different plotting functions.

#### Function introduction

The `set_enrichplot_color()` function can be added to any `enrichplot` visualization using the `+` operator (ggplot2 style). It offers flexible control over color transformations and palettes.

#### Parameters

- `type`: Type of color aesthetic to modify (default: `"fill"`, can also be `"color"` for line colors)
- `transform`: Transformation to apply to the values used for color mapping
  - `"log10"`: Apply log10 transformation (default for recent versions)
  - `"identity"`: Use original values without transformation
- `colors`: Custom color palette as a vector of colors (e.g., `c("red", "blue")`)

#### Usage examples

```{r}
#| label: fig-set-enrichplot-color
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Color customization using `set_enrichplot_color()`.** 
#|   (A) Default log10 transformation.
#|   (B) Custom red-blue palette with log10 transformation.
#|   (C) Identity transformation (no transformation).

# Example data
library(DOSE)
data(geneList)
de <- names(geneList)[abs(geneList) > 2]
edo <- enrichDGN(de)

# A: Default log10 transformation
p1 <- dotplot(edo, showCategory=15) + 
  set_enrichplot_color(type='fill', transform='log10') +
  ggtitle("Default log10 transform")

# B: Custom colors with log10 transformation  
p2 <- dotplot(edo, showCategory=15) + 
  set_enrichplot_color(type='fill', transform='log10', colors=c("red", "blue")) +
  ggtitle("Red-blue palette")

# C: Identity transformation (no transformation)
p3 <- dotplot(edo, showCategory=15) + 
  set_enrichplot_color(type='fill', transform='identity') +
  ggtitle("Identity transform")

library(aplot)
plot_list(p1, p2, p3, ncol=3, tag_levels='A')
```

**Note**: Starting from enrichplot v1.29.2, log10 transformation of p-values is applied by default in functions like `dotplot()`. The `set_enrichplot_color()` function allows users to override this default behavior or customize color palettes consistently across all visualization types.

Another method to plot GSEA result is the `gseaplot2` function:


```{r}
#| label: fig-gseaplot2  
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result.
#| fig-cap: |
#|   **Gseaplot2 for GSEA result.** 
gseaplot2(edo2, geneSetID = 1, title = edo2$Description[1])
```

The `gseaplot2` also supports multile gene sets to be displayed on the same figure:


```{r}
#| label: fig-gseaplot22  
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result of multile gene sets.
#| fig-cap: |
#|   **Gseaplot2 for GSEA result of multile gene sets.** 
gseaplot2(edo2, geneSetID = 1:3)
```

User can also displaying the pvalue table on the plot via `pvalue_table`
parameter:


```{r}
#| label: fig-gseaplot23  
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result of multile gene sets(add pvalue_table).
#| fig-cap: |
#|   **Gseaplot2 for GSEA result of multile gene sets(add pvalue_table).** 
gseaplot2(edo2, geneSetID = 1:3, pvalue_table = TRUE,
          color = c("#E495A5", "#86B875", "#7DB0DD"), ES_geom = "dot")
```


The `pvalue_table` can be customized using the following parameters:

- `pvalue_table_rownames` to specify row names of the table (if NULL, no row names will be displayed)
- `pvalue_table_columns` to specify column names of the table


```{r}
#| label: fig-gseaplot-pvalue  
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 with customized pvalue table.
#| fig-cap: |
#|   **Gseaplot2 with customized pvalue table.** 
#| 
gseaplot2(edo2, geneSetID = 1, pvalue_table = TRUE,
          pvalue_table_rownames = NULL,
          pvalue_table_columns = c("ID", "NES", "p.adjust"))
```


User can specify `subplots` to only display a subset of plots:


```{r}
#| label: fig-gseaplot24
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Gseaplot2 for GSEA result of multile gene sets(add subplots).
#| fig-cap: |
#|   **Gseaplot2 for GSEA result of multile gene sets(add subplots).** 
#|   `subplots = 1` (A),`subplots = 1:2` (B)
#| 
p1 <- gseaplot2(edo2, geneSetID = 1:3, subplots = 1)
p2 <- gseaplot2(edo2, geneSetID = 1:3, subplots = 1:2)
plot_list(p1, p2, ncol=1, tag_levels = 'A')
```


### Labeling genes in GSEA plot

Users can use `geom_gsea_gene()` to label specific genes in the GSEA plot.

```{r}
#| label: fig-geom-gsea-gene
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Labeling genes in GSEA plot.**
library(ggplot2)
library(ggrepel)

# Get gene set ID
id <- edo2$ID[1]

# Randomly select genes to label
set.seed(123)
genes <- sample(edo2[[id]], 5)

# Label genes on gseaplot2
p <- gseaplot2(edo2, geneSetID = 1, title = edo2$Description[1])

# Add geom_gsea_gene layer to the first subplot (running score)
p[[1]] <- p[[1]] + geom_gsea_gene(genes, geom=geom_label)
p
```

If users prefer to label with gene symbols, they can convert the gene IDs to symbols (e.g., using `setReadable`) before plotting.

```{r}
#| label: fig-geom-gsea-gene-symbol
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Labeling gene symbols in GSEA plot.**
#| eval: false

library(clusterProfiler)
# Assuming org.Hs.eg.db is available
if (require("org.Hs.eg.db")) {
    edo2_symbol <- setReadable(edo2, 'org.Hs.eg.db', 'ENTREZID')
    id <- edo2_symbol$ID[1]
    genes_symbol <- sample(edo2_symbol[[id]], 5)
    
    p_symbol <- gseaplot2(edo2_symbol, geneSetID = 1, title = edo2_symbol$Description[1])
    p_symbol[[1]] <- p_symbol[[1]] + geom_gsea_gene(genes_symbol, geom=geom_text_repel)
    p_symbol
}
```




The `gsearank` function plot the ranked list of genes belong to the specific
gene set.


```{r}
#| label: fig-gsearank
#| fig-width: 12
#| fig-height: 8
#| fig-scap: |
#|   Ranked list of genes belong to the specific gene set.
#| fig-cap: |
#|   **Ranked list of genes belong to the specific gene set.**
gsearank(edo2, 1, title = edo2[1, "Description"])
```

Multiple gene sets can be aligned using `cowplot`:

```{r}
#| label: fig-gsearank2
#| fig-width: 8
#| fig-height: 12
#| fig-scap: |
#|   Gsearank for multiple gene sets.
#| fig-cap: |
#|   **Gsearank for multiple gene sets.**
#| 
library(ggplot2)

pp <- lapply(1:3, function(i) {
    anno <- edo2[i, c("NES", "pvalue", "p.adjust")]
    lab <- paste0(names(anno), "=",  round(anno, 3), collapse="\n")

    gsearank(edo2, i, edo2[i, 2]) + xlab(NULL) +ylab(NULL) +
        annotate("text", 10000, edo2[i, "enrichmentScore"] * .75, label = lab, hjust=0, vjust=0)
})
plot_list(gglist=pp, ncol=1)
```

### Extracting data from gsearank plot

The `gsearank` function can also output the data used for plotting by setting `output = "table"`. This allows users to inspect the running score and other metrics for genes in the gene set.

```{r}
#| label: tbl-gsearank
#| tbl-cap: |
#|   **Data extracted from gsearank plot.**
gsearank(edo2, 1, output = "table") |> head()
```

Users can also merge this table with gene information (e.g. Symbol) using `bitr` or other methods.

```{r}
#| label: tbl-gsearank-symbol
#| tbl-cap: |
#|   **Data extracted from gsearank plot with gene symbols.**
#| eval: false

rank_table <- gsearank(edo2, 1, output = "table")
# Assuming 'gene' column contains Entrez IDs
if (require("org.Hs.eg.db") && require("clusterProfiler")) {
    gene_info <- bitr(rank_table$gene, fromType="ENTREZID", toType="SYMBOL", OrgDb="org.Hs.eg.db")
    rank_table_symbol <- merge(rank_table, gene_info, by.x="gene", by.y="ENTREZID")
    head(rank_table_symbol)
}
```


## pubmed trend of enriched terms

One of the problem of enrichment analysis is to find pathways for further
investigation. Here, we provide `pmcplot` function to plot the number/proportion
of publications trend based on the query result from PubMed Central. Of course,
users can use `pmcplot` in other scenarios. All text that can be queried on PMC
is valid as input of `pmcplot`.


```{r}
#| label: fig-pmcplot
#| fig-width: 14
#| fig-height: 6
#| fig-scap: |
#|   Pmcplot of enrichment analysis.
#| fig-cap: |
#|   **Pmcplot of enrichment analysis.**
#| 
terms <- edo$Description[1:5]
p <- pmcplot(terms, 2010:2020)
p2 <- pmcplot(terms, 2010:2020, proportion=FALSE)
plot_list(p, p2, ncol=2)
```



<!--

using ggplot2 api

`[` accessor


ggplot(egobp[1:20], aes(x=reorder(Description, -pvalue), y=Count, fill=-p.adjust)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_continuous(low="blue", high="red") +
    labs(x = "", y = "", fill = "p.adjust") +
    theme(axis.text=element_text(size=11))

-->

## Volcano plot for ORA results

The `volplot()` function provides volcano plot visualization for over-representation analysis (ORA) results, allowing users to visualize the relationship between fold change (or other effect size metrics) and statistical significance.

```{r}
#| label: fig-volplot
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Volcano plot for ORA results.**
#|   Visualizing -log10(p-value) vs z-score.

# Example using enrichment result with fold change information
# Note: volplot requires results with fold change values
volplot(edo)
```

The `volplot()` function accepts standard ggplot2 aesthetics and can be customized with additional parameters for point size, color, and labeling thresholds.

## Horizontal GSEA plot

The `hplot()` function creates horizontal versions of GSEA plots, which can be useful for comparing multiple gene sets side-by-side or for presentations where horizontal layout is preferred.

```{r}
#| label: fig-hplot
#| fig-width: 12
#| fig-height: 8
#| fig-cap: |
#|   **Horizontal GSEA plot.**
#|   GSEA results displayed in horizontal orientation.

# Example using GSEA result
edo2 <- gseDO(geneList)
hplot(edo2, geneSetID = 1:3)
```

The `hplot()` function supports the same parameters as `gseaplot2()` but arranges the output horizontally. This is particularly useful for comparing multiple pathways or when vertical space is limited.

## GO Graph

The `goplot()` function can accept the output of `enrichGO` and visualize the enriched GO induced graph. See @fig-goplot for an example.

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
data(geneList, package = "DOSE")
de <- names(geneList)[abs(geneList) > 2]
ego <- enrichGO(de, OrgDb = org.Hs.eg.db, ont = "BP", readable = TRUE)
```

 

```{r}
#| label: fig-goplot
#| fig-height: 12
#| fig-width: 8
#| fig-cap: "**Goplot of enrichment analysis.**"
#| fig-scap: "Goplot of enrichment analysis."
goplot(ego)
```

The `plotGOgraph()` function is another option to visualize the GO topology.

```{r eval=FALSE}
plotGOgraph(ego)
```


## References {-}
