# Compare functional profiles {#sec-visualization-comparison}

| Aspect | Comparison views |
|---|---|
| **Question** | How do functional profiles differ across clusters, conditions, or experimental groups? |
| **Input** | A checked `compareClusterResult` with comparable input rules, universe definitions, and database identity across groups. |
| **Functions** | `dotplot()` and comparison-specific profile, network, and facet helpers. |
| **Output** | Comparative term tables and figures that expose shared and group-specific enrichment patterns. |
| **Limitations** | Visual differences can reflect different input sizes, backgrounds, filtering, or annotation coverage. The plot does not replace a formal between-group model. |

Keep the statistical comparison design in the analysis-engine chapters; use these plots to communicate the resulting profiles.


```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
suppressPackageStartupMessages({
    library(clusterProfiler)
    library(enrichplot)
    library(dplyr)
})
```

```{r}
library(clusterProfiler)
library(enrichplot)
library(DOSE)
library(org.Hs.eg.db)
data(gcSample)
data(geneList, package = "DOSE")

# Use a local GO mapping for deterministic compareCluster examples rather
# than requiring the KEGG REST service during every book build.
go_map <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = keys(org.Hs.eg.db, keytype = "ENTREZID"),
    columns = c("GO", "ONTOLOGY"),
    keytype = "ENTREZID"
) |>
    dplyr::filter(!is.na(GO), ONTOLOGY == "BP") |>
    dplyr::transmute(term = GO, gene = ENTREZID) |>
    dplyr::distinct()
go_names <- AnnotationDbi::select(
    GO.db::GO.db,
    keys = unique(go_map$term),
    columns = "TERM",
    keytype = "GOID"
) |>
    dplyr::filter(!is.na(TERM)) |>
    dplyr::transmute(term = GOID, name = TERM) |>
    dplyr::distinct()
xx <- compareCluster(
    gcSample,
    fun = "enricher",
    TERM2GENE = go_map,
    TERM2NAME = go_names,
    universe = names(geneList),
    pvalueCutoff = 1,
    qvalueCutoff = 1,
    minGSSize = 5
)
xx <- pairwise_termsim(xx)
```

## Dotplot2: Comparing two clusters {#compare-dotplot}

The focused `dotplot2()` example belongs here because it compares selected groups rather than introducing a new statistical test.

```{r}
#| label: fig-dotplot2
#| fig-width: 14
#| fig-height: 7
#| fig-cap: |
#|   **Dotplot2 for comparing two clusters.**
#|   Enrichment is shown on opposite sides for clusters X1 and X2;
#|   points encode GeneRatio and adjusted significance.
dotplot2(
    xx,
    vars = c("X1", "X2"),
    x = "FoldEnrichment",
    showCategory = 5,
    label_format = 30,
    font.size = 10
) +
    ggplot2::labs(
        title = "X1 versus X2",
        subtitle = "Terms selected separately within each cluster"
    )
```

`dotplot2()` accepts the same plotting controls as `dotplot()` but requires exactly two cluster labels in `vars`. The left and right sides show the selected clusters on a shared fold-enrichment scale; the dashed center line marks no enrichment difference. This is a visual comparison, not a replacement for a formal between-group test.


## Enrichment maps for compareCluster results {#compare-emapplot}

`compareCluster()` produces a standard comparison object. The map below shows how its functional profiles can be compared visually; the statistical design and multiple-testing questions remain in the analysis-engine chapter.

```{r}
#| label: fig-Enrichment2
#| fig-width: 16
#| fig-height: 14
#| fig-scap: |
#|   Enrichment maps for a compareCluster result with a small, explicit term set.
#| fig-cap: |
#|   **Enrichment maps for the compareCluster result.**
#|   Each cluster contributes at most two terms. Panels show equal-size pies
#|   (A), count-weighted pies (B), larger count-weighted nodes (C), and the
#|   count-weighted Kamada–Kawai layout (D).
map_guides <- ggplot2::guides(size = "none")
p1 <- emapplot(
    xx,
    showCategory = 2,
    pie = "equal",
    min_edge = 0.3,
    label_format = 18,
    node_label_size = 3.2,
    show_category_size_legend = FALSE
) + map_guides
p2 <- emapplot(
    xx,
    showCategory = 2,
    pie = "count",
    min_edge = 0.3,
    label_format = 18,
    node_label_size = 3.2,
    show_category_size_legend = FALSE
) + map_guides
p3 <- emapplot(
    xx,
    showCategory = 2,
    pie = "count",
    size_category = 1.2,
    min_edge = 0.3,
    label_format = 18,
    node_label_size = 3.2,
    show_category_size_legend = FALSE
) + map_guides
p4 <- emapplot(
    xx,
    showCategory = 2,
    pie = "count",
    size_category = 1.2,
    layout = "kk",
    min_edge = 0.3,
    label_format = 18,
    node_label_size = 3.2,
    show_category_size_legend = FALSE
) + map_guides
aplot::plot_list(p1, p2, p3, p4, ncol = 2, tag_levels = "A")
```

## Next steps

- For the underlying comparison design, see [compareCluster](comparecluster-analysis.qmd#clusterprofiler-comparecluster).
- For term organization and reporting, continue to [evidence integration](03-evidence-integration.qmd).

