

# Frequently asked questions {#faq}

```{r}
#| echo: false
#| message: false
#| cache: false

source("_common.R")
```

## How to prepare your own geneList {#sec-genelist}

GSEA analysis requires a ranked gene list, which contains three features:


+ numeric vector: fold change or other type of numerical variable
+ named vector: every number has a name, the corresponding gene ID
+ sorted vector: number should be sorted in decreasing order


If you import your data from a `csv` file, the file should contains two columns, one for gene ID (no duplicated ID allowed) and another one for fold change. You can prepare your own `geneList` via the following command:


```r
d = read.csv(your_csv_file)
## assume 1st column is ID
## 2nd column is FC

## feature 1: numeric vector
geneList = d[,2]

## feature 2: named vector
names(geneList) = as.character(d[,1])

## feature 3: decreasing order
geneList = sort(geneList, decreasing = TRUE)
```





## No gene can be mapped

+ <https://www.biostars.org/p/431270/>
+ <https://github.com/YuLab-SMU/clusterProfiler/issues/280>

### First, check that your identifiers are still identifiers

Before blaming the annotation database, look at the IDs themselves. A very common cause of `No gene can be mapped` is that the ID column passed through a spreadsheet on its way into R. Excel quietly turns long numbers into scientific notation, and when the column is too narrow or the cell is formatted as a number it writes out the *displayed* text rather than the value. A gene list then arrives looking like this:

```
25081
84604
1.01E+08     <- should be an Entrez ID
1E+08        <- should be an Entrez ID
24614
```

None of these can be mapped, and the failure is easy to misread as "clusterProfiler does not support my organism" — particularly when pasting the same list into DAVID seems to work, because DAVID re-parses the text differently.

Note that `1E+08` is not merely mis-formatted, it is **lossy**: the original identifier is gone and cannot be recovered. The fix is therefore not to repair the strings in R, but to re-export the IDs from their source.

A quick check in R — any ID that is not a plain integer is suspect:

```{r}
#| label: faq-check-ids
ids <- c("25081", "84604", "1.01E+08", "1E+08", "24614")
ids[!grepl("^[0-9]+$", ids)]
```

**How to avoid it.** Export the ID column as *text*: in Excel, format the column as "Text" *before* pasting the IDs in, or import via "Data → From Text/CSV" and set the column type to text. Then read it into R as character, e.g. `readr::read_tsv(file, col_types = readr::cols(ENTREZID = "c"))`. Never round-trip identifiers through a spreadsheet's default number formatting.

### Other causes

If the IDs are clean integers and the problem persists, the usual suspects are a mismatched `keyType`, an `organism` code that does not correspond to your IDs, or an outdated annotation source — see the two links above.


## Showing specific pathways {#showing-specific-pathways}

By default, all the visualization methods provided by `r Biocpkg("enrichplot")` display most significant pathways. If users are interested to show some specific pathways (e.g. excluding some unimportant pathways among the top categories), users can pass a vector of selected pathways to the `showCategory` parameter in `dotplot()`, `barplot()`, `treeplot()`, `cnetplot()` and `emapplot()` etc.


See @fig-selectedPaths for a side-by-side comparison of default top categories versus user-selected pathways.

```{r}
#| label: fig-selectedPaths
#| fig-height: 6
#| fig-width: 15
#| fig-cap: "**Showing specific pathways.** Top ten most significant pathways (A), selected ten pathways (B)."
#| fig-scap: "Showing specific pathways."
library(clusterProfiler)
library(enrichplot)
data(geneList, package='DOSE')
de <- names(geneList)[abs(geneList)>1]

x <- enrichKEGG(de)

## show top 10 most significant pathways and want to exclude the second one
## dotplot(x, showCategory = x$Description[1:10][-2])

set.seed(2020-10-27)
selected_pathways <- sample(x$Description, 5)
selected_pathways

p1 <- dotplot(x, showCategory = 10, font.size=14)
p2 <- dotplot(x, showCategory = selected_pathways, font.size=14)


aplot::plot_list(p1, p2, tag_levels = "A")
```

Note: Another solution is using the `filter` verb to extract a subset of the result as described in [Chapter 16](#clusterProfiler-dplyr).


## How to extract genes of a specific term/pathway 

```{r}
id <- x$ID[1:3]
id
x[[id[1]]]
geneInCategory(x)[id]
```


## Wrap long axis labels {#label-format}

Most of the functions in `r Biocpkg("enrichplot")` can automatically split long labels across multiple lines. Users can passed a line width to the `label_format` parameter (default is 30). It also supports user defined function to format label strings. 

See @fig-formatLabel for examples of wrapping long axis labels using numeric width and a custom labeller.

```{r}
#| label: fig-formatLabel
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "**Wrap long axis labels.** Passing a numeric value to specify string width (A), a user specifiable labeller function (B)."
#| fig-scap: "Wrap long axis labels."
library(ReactomePA)
y <- enrichPathway(de)

p1 <- dotplot(y, label_format = 20) 
p2 <- dotplot(y, label_format = function(x) stringr::str_wrap(x, width=20))
cowplot::plot_grid(p1, p2, ncol=2, labels=c("A", "B")) 
```


The `label_format` option works with `barplot()`, `dotplot()`, `heatplot()`, `treeplot` and `ridgeplot()`.

## Why many genes are not annotated in KEGG?

Users often find that a significant portion of their gene list is dropped in the KEGG enrichment analysis. For example, a user reported that only 281 genes were retained from a list of over 700 genes.

This is not a software issue but rather a reflection of the database coverage. KEGG pathways mainly focus on metabolic and signaling pathways, and many genes are not annotated in any KEGG pathway.

We can verify the annotation coverage using `bitr_kegg()`:

```{r eval=FALSE}
# gene is a vector of gene IDs (e.g., Entrez IDs)
kk <- bitr_kegg(geneID = gene, fromType='ncbi-geneid', toType='Path', organism='hsa')
```

The warning message will indicate the percentage of genes that failed to map.

To manually verify a specific gene, you can visit the KEGG website using a URL constructed with the organism code and gene ID, for example: `http://www.genome.jp/dbget-bin/www_bget?hsa:100506775`. If the gene page does not list any pathway, it means the gene has no KEGG pathway annotation.

