# The first complete workflow: ORA, GSEA, and plots {#first-workflow}

This chapter starts from the committed `datasets/de_table.tsv` file derived from the Bioconductor `airway` example with DESeq2. The table contains Ensembl IDs, differential-expression statistics, and adjusted p-values for dexamethasone-treated versus untreated airway smooth-muscle samples. The table is local at render time; the regeneration script and package provenance are documented in `datasets/readme.md`.

The example answers two related questions about the same contrast:

* **ORA:** among genes selected by an explicit adjusted-p-value and fold-change rule, which biological-process terms are over-represented?
* **GSEA:** across the complete DESeq2 Wald-statistic ranking, which terms are shifted toward the top or bottom?

The thresholds are teaching choices. In a real analysis, replace them with rules defined before inspecting enrichment results.

## 1. Load local packages and inspect the input

```{r}
#| label: first-workflow-setup
#| message: false
#| warning: false

library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)

de_table <- read.delim(
    "datasets/de_table.tsv",
    stringsAsFactors = FALSE,
    check.names = FALSE
)

stopifnot(
    identical(
        names(de_table),
        c("gene_id", "baseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj")
    ),
    !anyDuplicated(de_table$gene_id)
)

head(de_table)
summary(de_table[, c("baseMean", "log2FoldChange", "stat", "padj")])
```

The table uses Ensembl gene IDs and the `stat` column supplies a signed ranking. Keep both facts visible in an analysis script. An identifier can be syntactically valid and still be incompatible with the selected annotation database.

## 2. Define the universe, selected genes, and ranking

The ORA universe starts with genes having a finite DESeq2 statistic and is then restricted to Ensembl IDs present in `org.Hs.eg.db`. The selected list applies a fixed `padj` and absolute log2 fold-change rule. GSEA uses the complete finite, mapped statistic ranking, including genes that do not pass the ORA threshold.

```{r}
#| label: first-workflow-inputs

eligible <- is.finite(de_table$stat)
candidate_universe <- de_table$gene_id[eligible]
annotation_keys <- AnnotationDbi::keys(org.Hs.eg.db, keytype = "ENSEMBL")
eligible <- eligible & de_table$gene_id %in% annotation_keys

universe <- de_table$gene_id[eligible]
selected <- de_table$gene_id[
    eligible &
        !is.na(de_table$padj) &
        de_table$padj < 0.05 &
        abs(de_table$log2FoldChange) >= 1
]

ranking <- de_table$stat[eligible]
names(ranking) <- de_table$gene_id[eligible]
ranking <- sort(ranking, decreasing = TRUE)

cat("Rows in input table:", nrow(de_table), "\n")
cat("Genes eligible for testing:", length(candidate_universe), "\n")
cat("Genes mapped to org.Hs.eg.db:", length(universe), "\n")
cat("Unmapped eligible genes:", length(setdiff(candidate_universe, universe)), "\n")
cat("Selected genes for ORA:", length(selected), "\n")
cat("Selected genes in universe:", all(selected %in% universe), "\n")
```

`universe`, `selected`, and `ranking` use the same Ensembl namespace after filtering against the keys available in `org.Hs.eg.db`. The unmapped count is reported rather than silently treated as biological absence. In a real experiment, construct the candidate universe from genes that were measurable and eligible for selection, then record the annotation coverage.

## 3. Run over-representation analysis

`enrichGO()` tests Gene Ontology terms using the selected vector. `ont = "BP"` restricts this example to biological process terms; use `"MF"`, `"CC"`, or `"ALL"` when that choice matches the question. The permissive `pvalueCutoff = 1` and `qvalueCutoff = 1` settings keep the complete result object for teaching and plotting; apply a pre-specified reporting threshold when interpreting a real study.

```{r}
#| label: first-workflow-ora
#| message: false
#| warning: false

ora <- enrichGO(
  gene          = selected,
  universe      = universe,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  qvalueCutoff  = 1,
  minGSSize     = 10,
  maxGSSize     = 500,
  readable      = FALSE
)

ora
head(as.data.frame(ora))
```

The output is an `enrichResult`. Its `@result` slot can be converted to a data frame, but retain `ora` itself: the object carries the gene-to-term information needed by downstream plots.

`enrichGO()` accepts the Ensembl input but uses the OrgDb mapping internally; the returned object records the internal key type as `ENTREZID`. Readable symbols can be added for inspection without changing the test:

```{r}
#| label: first-workflow-readable-ora
#| message: false

ora_readable <- setReadable(ora, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
head(as.data.frame(ora_readable)[, c("ID", "Description", "GeneRatio", "p.adjust", "geneID")])
```

## 4. Visualize the ORA result

A dot plot is a compact first view. Dot size is the number of selected genes in a term, and colour represents the adjusted significance used by the plotting method. A network plot is useful for seeing shared genes among a small number of terms.

```{r}
#| label: first-workflow-ora-plots
#| fig-width: 8
#| fig-height: 5

p_ora_dot <- dotplot(ora, showCategory = 10) +
  ggtitle("ORA: selected genes and biological processes")
p_ora_dot
```

```{r}
#| label: first-workflow-ora-network
#| fig-width: 8
#| fig-height: 7

p_ora_network <- cnetplot(
  ora,
  showCategory = 3,
  foldChange = ranking,
  categorySize = "pvalue"
)
p_ora_network
```

`cnetplot()` is intentionally limited to three terms here. Showing every term usually produces a dense graph that is harder to read than the dot plot.

## 5. Run GSEA on the complete ranking

GSEA does not use `selected` as its input. It uses the complete named `ranking`, so genes below the ORA threshold still contribute. The `minGSSize` and `maxGSSize` settings keep this first local example reasonably quick while excluding extremely small or very broad terms. Setting cutoffs to `1` retains the tested terms so that the plotting example remains informative even when few terms meet a conventional significance threshold.

```{r}
#| label: first-workflow-gsea
#| message: false
#| warning: false
#| cache: true

set.seed(1)
gsea <- gseGO(
  geneList      = ranking,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  minGSSize     = 100,
  maxGSSize     = 300,
  pAdjustMethod = "BH",
  pvalueCutoff  = 1,
  eps           = 0,
  verbose       = FALSE
)

gsea
head(as.data.frame(gsea))
```

The result is a `gseaResult`. Its `NES` and `enrichmentScore` describe direction and magnitude relative to the ranked list; the adjusted p-value controls for testing many terms. Inspect both effect direction and statistical evidence rather than sorting only by a single column.

## 6. Visualize the GSEA result

The dot plot gives a profile-level overview. `gseaplot2()` shows the running enrichment score, the locations of genes in the selected set, and the ranked statistic for one term.

```{r}
#| label: first-workflow-gsea-dotplot
#| fig-width: 8
#| fig-height: 5

p_gsea_dot <- dotplot(gsea, showCategory = 10) +
  ggtitle("GSEA: biological processes across the full ranking")
p_gsea_dot
```

```{r}
#| label: first-workflow-gsea-curve
#| fig-width: 8
#| fig-height: 6

if (nrow(as.data.frame(gsea)) > 0) {
  gseaplot2(gsea, geneSetID = 1, title = gsea$Description[1])
}
```

The x-axis of a GSEA curve is the position in the ranked list, not a gene identifier. A term enriched at the top has a positive direction for this signed ranking; a term enriched at the bottom has a negative direction.

## 7. Compare the two questions and save results

ORA and GSEA can both be useful, but they are not duplicate tests. ORA asks whether a thresholded set has more members of a term than expected under the universe. GSEA asks whether term members accumulate toward one end of the complete ranking. Report which input and background were used for each result.

```{r}
#| label: first-workflow-export
#| eval: false

# These files are optional outputs; the result objects remain the analysis record.
write.csv(de_table, "de-table-used.csv", row.names = FALSE)
write.csv(as.data.frame(ora_readable), "ora-go-bp-results.csv", row.names = FALSE)
write.csv(as.data.frame(gsea), "gsea-go-bp-results.csv", row.names = FALSE)
saveRDS(
    list(
        input = de_table,
        universe = universe,
        selected = selected,
        ranking = ranking,
        ora = ora,
        gsea = gsea
    ),
    "enrichment-objects.rds"
)
```

For a short report, start with the two dot plots and one GSEA curve. For a detailed report, include the ID type, universe construction, threshold, ontology, package versions, and the complete result tables.

::: {.callout-note}
**Offline boundary.** This chapter uses data shipped with R/Bioconductor packages. `org.Hs.eg.db` is a local annotation package, and the example does not query KEGG, Enrichr, g:Profiler, or another remote service. A real study should record the annotation package version because local annotations change over time.
:::

## Next steps

- For input and universe details, see [the data contract](data-contract.qmd).
- For alternative designs, use [Choose a route](choose-a-route.qmd).
- For deeper plotting and interpretation, continue to [visualization](enrichplot.qmd#sec-enrichplot) and [evidence-guided interpretation](interpretation.qmd#evidence-guided-interpretation).

