# GO enrichment analysis {#clusterprofiler-go}

```{r}
#| label: load-packages
#| echo: false
#| message: false
#| cache: false

source("_common.R")
loadNamespace("Rcpp")
```


GO comprises three orthogonal ontologies, i.e. molecular function (MF), biological
process (BP), and cellular component (CC).


## Chapter overview

| Aspect | GO enrichment workflow |
|---|---|
| **Questions** | Which GO terms are over-represented in a selected gene set or enriched across a ranked profile? How can genes be classified by ontology level? |
| **Input** | Gene IDs compatible with an `OrgDb` (commonly Entrez), a named ranked vector for GSEA, or custom GO `TERM2GENE` annotations. |
| **Methods** | ORA, GSEA, GO-level classification, custom annotation enrichment, and semantic redundancy reduction. |
| **Main functions** | `groupGO()`, `enrichGO()`, `gseGO()`, `buildGOmap()`, `enricher()`, `GSEA()`, `simplify()`. |
| **Output** | Classification tables and `enrichResult` / `gseaResult` objects ready for `enrichplot`. |
| **Main limitations** | Coverage depends on the `OrgDb` release; GO is hierarchical, so ancestor propagation and ontology choice affect interpretation. Use a defensible universe and include ancestor annotations in custom mappings. |

::: {.callout-note}
Confirm identifier and background-set choices first. For custom GO maps, include ancestor terms where required before testing.
:::




```{r include=FALSE}
#| label: setup-options
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, cache=TRUE)
library(clusterProfiler)
library(org.Hs.eg.db)
```


## Supported organisms {#clusterProfiler-go-supported-organisms}

GO analyses (`groupGO()`, `enrichGO()` and `gseGO()`) support organisms that have an `OrgDb` object available (see also [session 2.2](go-semantic-similarity.qmd#gosemsim-supported-organisms)).


If a user has GO annotation data (in a `data.frame` format with the first column as gene ID and the second column as GO ID), they can use the `enricher()` and `GSEA()` functions to perform an over-representation test and gene set enrichment analysis. Users can also read GO annotation from GMT files using the `read.gmt()` function, which parses the file into a `data.frame` suitable for these functions.

If the genes are annotated by direct annotation, they should also be annotated by their ancestor GO nodes (indirect annotation). If a user only has direct annotation, they can pass their annotation to the `buildGOmap` function, which will infer indirect annotation and generate a `data.frame` that is suitable for both `enricher()` and `GSEA()`.

## Direct vs indirect GO annotation

One common issue users encounter is the discrepancy between direct gene-to-GO mapping and the results from enrichment analysis. For instance, a user might find that a specific GO term is enriched, but when they check the input gene list against the direct annotations, the term doesn't appear to be associated with many genes.

This happens because GO structure is a Directed Acyclic Graph (DAG). If a gene is annotated to a specific term (e.g., "glucose metabolic process"), it is implicitly annotated to all its ancestor terms (e.g., "metabolic process"). `enrichGO` and `gseGO` account for this **indirect annotation** by propagating annotations up the graph.

If you are providing your own annotation file (e.g., using `enricher` or `GSEA`), you must ensure that it includes these indirect annotations. If your annotation file only contains direct annotations, you can use the `buildGOmap()` function to generate the full annotation set (direct + indirect) before performing enrichment analysis.

```{r eval=FALSE}
#| label: buildGOmap-example
# Assuming 'gomap' is a data.frame with direct annotations (GeneID, GOID)
# Build indirect annotations
gomap_with_ancestors <- buildGOmap(gomap)

# Use the result for enrichment
enricher(gene, TERM2GENE = gomap_with_ancestors, ...)
```

This step is crucial for accurate GO enrichment analysis when not using `OrgDb` objects.

## GO classification

In `r Biocpkg("clusterProfiler")`, the `groupGO()` function is designed for gene classification based on GO distribution at a specific level. Here we use the dataset `geneList` provided by `r Biocpkg("DOSE")`. 

```{r warning=FALSE}
#| label: groupGO-example
library(clusterProfiler)
library(org.Hs.eg.db)

data(geneList, package="DOSE")
gene <- names(geneList)[abs(geneList) > 2]

# Entrez gene ID
head(gene)

ggo <- groupGO(gene     = gene,
               OrgDb    = org.Hs.eg.db,
               ont      = "CC",
               level    = 3,
               readable = TRUE)

head(ggo)
```

The `gene` parameter is a vector of gene IDs (can be any ID type that is supported by the corresponding `OrgDb`, see also @sec-id-convert). If `readable` is set to `TRUE`, the input gene IDs will be converted to gene symbols.


## GO over-representation analysis {#clusterprofiler-go-ora}


The `r Biocpkg("clusterProfiler")` package implements `enrichGO()` for gene ontology over-representation test.


```{r}
#| label: enrichGO-example
ego <- enrichGO(gene          = gene,
                universe      = names(geneList),
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
		readable      = TRUE)
head(ego)
```

Any gene ID type that is supported in `OrgDb` can be directly used in GO analyses. Users need to specify the `keyType` parameter to specify the input gene ID type.

For `enrichGO()`, non-`ENTREZID` inputs use an Entrez-backed representation as a memory-saving internal path when possible. If source IDs cannot be mapped to Entrez, their native `keyType` GO annotations are added to a hybrid GSON instead of being silently dropped. When no `universe` is supplied, the effective annotated hybrid background is constructed and reported by the function.

The result is mapped back to the original input ID type, and the result object's `@keytype` records that source type. This keeps the user-facing result aligned with the input namespace while retaining the memory protection needed for one-to-many identifiers such as `ACCNUM`.

For example, if we use ENSEMBL IDs as input:

```{r}
#| label: enrichGO-ensembl
gene.df <- bitr(gene, fromType = "ENTREZID",
	   	toType = c("ENSEMBL", "SYMBOL"),
		OrgDb = org.Hs.eg.db)

ego2 <- enrichGO(gene         = gene.df$ENSEMBL,
                OrgDb         = org.Hs.eg.db,
                keyType       = 'ENSEMBL',
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05)
head(ego2, 3)                
```

Gene IDs can be mapped to gene Symbols by using the parameter `readable=TRUE` or [`setReadable()` function](identifier-utilities.qmd#sec-setReadable).


## GO Gene Set Enrichment Analysis

The `r Biocpkg("clusterProfiler")` package provides the `gseGO()` function for [gene set enrichment analysis](enrichment-foundations.qmd#gsea-algorithm) using gene ontology.

```{r eval=FALSE}
#| label: gseGO-example
ego3 <- gseGO(geneList     = geneList,
              OrgDb        = org.Hs.eg.db,
              ont          = "CC",
              minGSSize    = 100,
              maxGSSize    = 500,
              pvalueCutoff = 0.05,
              verbose      = FALSE)
```

The format of input data, `geneList`, is documented in the @sec-id-convert. Beware that only gene sets with size in `[minGSSize, maxGSSize]` will be tested.

::: {.callout-note title="Size filtering is overlap-based, not annotation-based" icon="lightbulb"}
`minGSSize` and `maxGSSize` operate on the **overlap** of each gene set with the genes you actually provide (the candidate vector for ORA, or the named `geneList` for GSEA), not on the raw, full annotation size stored in the ontology database. The code therefore does roughly:

```r
gene_sets <- lapply(gene_sets, intersect, names(geneList)) # or universe for ORA
idx <- get_geneSet_index(gene_sets, minGSSize, maxGSSize)
```

This is important for broad libraries such as MSigDB Hallmark collections, Reactome, or the upper HDO Slim. A database gene set annotated with 3 000 genes is **not** rejected just because 3 000 > `maxGSSize = 500`; it is kept if its overlap with your 12 000 measured genes has, say, 480 members. The only terms genuinely dropped are (a) very small overlaps (below `minGSSize`, which usually means the term has too few observed members for Fisher's exact test or the weighted Kolmogorov-Smirnov statistic to be reliable) and (b) extremely broad overlaps (above `maxGSSize`, which usually correspond to root-level, biologically uninformative nodes that inflate the multiple-testing burden without adding insight).

See also [Tuning `minGSSize` and `maxGSSize`](#sec-tune-gene-set-size).
:::

## Topology-aware GO enrichment

In classical GO enrichment, genes are treated as independent members of a gene set. When a biological network is available, `clusterProfiler` can now call the `enrichit` engine through the high-level wrappers `nseGO()` and `mnseGO()`. The former performs single-network enrichment, while the latter extends the same idea to a multi-layer setting.

To keep the example reproducible and consistent with the KEGG chapter, we reuse one shared demonstration object.

```{r}
#| label: topology-demo-go
demo <- clusterprofiler_enrichit_demo()
```

```{r}
#| label: nseGO-example
ego_nse <- nseGO(
  geneList = demo$geneList_evidence,
  network = demo$network,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  minGSSize = 5,
  maxGSSize = 500,
  threshold = 1e-6,
  maxIter = 50,
  verbose = FALSE,
  pvalueCutoff = 1,
  method = "sample",
  nPerm = 30
)
head(as.data.frame(ego_nse)[, c("ID", "Description", "NES", "p.adjust")], 3)
```

The `mnseGO()` function keeps the same GO-aware interface, but accepts a list of layer-specific networks together with explicit inter-layer couplings.

```{r}
#| label: mnseGO-example
ego_mnse <- mnseGO(
  seed_list = demo$seed_list,
  networks = list(RNA = demo$network, PROT = demo$network_2),
  couplings = demo$couplings,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  collapse = "weighted_mean",
  layer_weights = c(RNA = 1, PROT = 1.2),
  minGSSize = 5,
  maxGSSize = 500,
  threshold = 1e-6,
  maxIter = 50,
  verbose = FALSE,
  pvalueCutoff = 1,
  method = "sample",
  nPerm = 30
)
head(as.data.frame(ego_mnse)[, c("ID", "Description", "NES", "p.adjust")], 3)
```

These wrappers preserve the familiar GO semantics (`OrgDb`, `ont`, and `keyType`) while delegating the network propagation and enrichment engine to `enrichit`.




Both the `enrichGO()` and `gseGO()` functions require an `OrgDb` object as the background annotation. For organisms that don't have `OrgDb` provided by `Bioconductor`, users can query one (if available) online via `r Biocpkg("AnnotationHub")`. If there is no `OrgDb` available, users can obtain GO annotation from other sources, e.g. from `r Biocpkg("biomaRt")`, or annotate the genes using [Blast2GO](https://www.blast2go.com/) or the [Trinotate](https://rnabio.org/module-07-trinotate/0007/02/01/Trinotate/) pipeline. Then the `enricher()` or `GSEA()` functions can be used to perform GO analysis for these organisms, similar to the examples using WikiPathways and MSigDB. Another solution is to create an `OrgDb` on your own using the `r Biocpkg("AnnotationForge")` package.


Here is an example of querying GO annotation from Ensembl using `r Biocpkg("biomaRt")`. 


```{r eval=FALSE}
#| label: biomaRt-example
library(biomaRt)
ensembl <- useEnsemblGenomes(biomart = "plants_mart", dataset = "nattenuata_eg_gene")
gene2go <- getBM(attributes =c("ensembl_gene_id", "go_id"), mart=ensembl)
```

Alternatively, you can use `AnnotationHub` to query and retrieve `OrgDb` objects for a wide range of organisms, including non-model species.

**Example: GO enrichment analysis for Maize (Zea mays)**

```{r eval=FALSE}
#| label: annotationhub-maize
library(AnnotationHub)
hub <- AnnotationHub()

# Query for Zea mays OrgDb
query(hub, "zea")
# Retrieve the OrgDb (e.g., AH55736, assume it's the correct record)
maize <- hub[['AH55736']]

# Check keys
length(keys(maize))
columns(maize)

# Perform enrichment analysis
# Assuming 'sample_genes' is a vector of Entrez IDs
res <- enrichGO(sample_genes, OrgDb=maize, pvalueCutoff=0.05, qvalueCutoff=0.05)
```

This approach allows you to use the standard `enrichGO` workflow with organisms not included in the default Bioconductor `OrgDb` packages.

## Visualization of GO enrichment results

The `clusterProfiler` package provides several visualization methods to help interpret enrichment results, including `barplot`, `dotplot`, `cnetplot`, `emapplot`, `goplot` and `plotGOgraph`.

Please refer to the @sec-enrichplot chapter for details.


## Filtering and simplifying GO terms

GO enrichment analysis often results in many redundant terms due to the hierarchical structure of GO. Parent terms and child terms often share a large proportion of genes, leading to multiple significant results that represent similar biological processes.

### Removing specific terms or levels

The `dropGO()` function can be used to remove specific GO terms or GO levels from the results. The `enrichGO()` function tests the whole GO corpus and enriched result may contains very general terms. If users want to restrict the result to a specific GO level (e.g., level 3 or 4), they can use the `gofilter()` function. Both functions work with results obtained from `enrichGO()`, `gseGO()` and `compareCluster()`.

To see the level of each term rather than filter by a level you already know, `add_go_level()` appends it as a `level` column, where level 1 is the ontology root (`GO:0008150` for BP, `GO:0005575` for CC, `GO:0003674` for MF). This makes an arbitrary band easy to keep:

```{r}
#| label: go-add-level
#| eval: false
x <- enrichGO(gene, OrgDb = "org.Hs.eg.db", ont = "BP")
x <- add_go_level(x)
subset(x, level >= 3 & level <= 6)
```

Note that GO is a directed acyclic graph, so a term can be reachable at more than one depth; the shallowest level is reported. The level is resolved from the ontology the term itself belongs to, so an `ont = "ALL"` result (or a `compareCluster()` result) is labelled term by term.

### Simplifying enriched GO terms

To reduce redundancy, `clusterProfiler` provides a `simplify` method. It uses semantic similarity (via `GOSemSim`) to calculate the similarity between enriched terms and removes highly similar terms, keeping the most significant one.

The `simplify` method applies `select_fun` (which can be a user-defined function) to the feature specified by `by` (e.g., p.adjust) to select one representative term from redundant terms (which have similarity higher than `cutoff`). This results in a cleaner and more interpretable visualization.

A complementary strategy is Bayesian term selection, which uses term-gene coverage to rank terms by posterior explanatory support instead of semantic similarity alone (see @sec-bayes-enrich).

In @fig-simplify-emapplot(A), we can found that there are many redundant terms form a highly condense network. After removing redundant terms using the `simplify()` method, the result is more clear to view the whole story.

```{r}
#| label: fig-simplify-emapplot
#| fig-width: 16
#| fig-height: 8
#| fig-scap: |
#|   Visualize enriched terms by EnrichmentMap using the `emapplot()` function.
#| fig-cap: |
#|   **Visualize enriched terms by EnrichmentMap using the `emapplot()` function.**
#|   (A) original result. (B) simplify result.
#|
library(enrichplot)
data(geneList, package="DOSE")
de <- names(geneList)[abs(geneList) > 2]
bp <- enrichGO(de, ont="BP", OrgDb = 'org.Hs.eg.db')
bp <- pairwise_termsim(bp)
bp2 <- simplify(bp, cutoff=0.7, by="p.adjust", select_fun=min)
p1 <- emapplot(bp)
p2 <- emapplot(bp2)
plot_list(p1, p2, ncol=2, tag_levels = 'A')
```

Alternatively, users can use slim version of GO and use the `enricher()` or `gseGO()` functions to analyze.

**Note on `enricher` results**: The `simplify()` method only needs to know which ontology the terms belong to. For `enrichGO()` and `gseGO()` results that ontology is recorded in the object, but a result built with `enricher()` (e.g. from a custom annotation or an MSigDB GO collection) does not carry one. In that case `simplify()` now derives it from the GO IDs themselves, so no manual step is required:

```{r eval=FALSE}
#| label: simplify-enricher-example
# 'res' is the result from enricher() using a GO gene-set collection
res_simplified <- simplify(res)
```

The inference requires that the gene-set IDs really are GO terms (`GO:0008150`, ...): if a single ontology is represented it is used directly, and if the terms span several ontologies they are simplified as `GOALL`. A result whose gene sets are *not* GO terms is still rejected, because `simplify()` reduces redundancy through GO semantic similarity. Setting the slot explicitly (`res@ontology <- "BP"`) still works, and remains necessary only when the IDs are not GO terms.


## Troubleshooting

### No results found

It is common to receive a message saying "No gene set have size > 10 ... return NULL" or finding no significant terms.

1.  **Gene Set Size**: The default `minGSSize` is 10 and `maxGSSize` is 500. Remember that filtering is based on the overlap with your provided input, not the raw database annotation size (see [Size filtering is overlap-based, not annotation-based](#size-filtering-is-overlap-based-not-annotation-based)). A term whose full annotation has 3 000 genes may still pass if only 200 of those genes were measured in your experiment. If no term passes both cutoffs, enlarge the window with, for example, `minGSSize = 5, maxGSSize = 5000`, or set `maxGSSize = Inf` to disable the upper cap entirely.

### Tuning `minGSSize` and `maxGSSize` {#sec-tune-gene-set-size}

The default `maxGSSize = 500` is a conservative first-pass safety net, not a hard recommendation. It was chosen to avoid root-level GO/HDO nodes and extremely broad Hallmark-style gene sets overwhelming the multiple-testing correction in a typical short list of a few hundred differentially expressed genes. It also used to cap raw annotation sizes in older code paths, but that restriction no longer applies: size filtering now happens **after** intersecting with your measured genes.

The table below is a pragmatic starting point. Always verify the term list after the first run and adjust the window to the characteristic scale of your annotation library and input list.

| Scenario                                                | Suggested `minGSSize` | Suggested `maxGSSize` |
|---------------------------------------------------------|----------------------:|----------------------:|
| Standard GO BP or HDO term-level analysis on ~200 DEGs  | 5–10                  | 500 (default)         |
| GO Slim, DO Slim or broad candidates / Reactome         | 10–20                 | 2000–5000             |
| Custom GMT, MSigDB Hallmark / C2, or user-defined sets  | 10                    | `Inf` (disable cap)   |
| You want to see everything and filter the result later  | 1                     | `Inf`                 |

Setting `maxGSSize = Inf` is safe and supported: `get_geneSet_index()` treats `NA` or `NULL` as `Inf` as well. If after running with wide bounds you see obviously uninformative root terms such as `biological_process` or broad "positive regulation of …" nodes, filter them out afterwards with `dplyr::filter()` or re-run with a narrower window rather than using a tight global default that may hide relevant biology.

For reproducibility, document the values you chose together with the annotation release and package versions; the [Appendix reproducibility record](appendix.qmd#reproducibility-record) lists this explicitly.
2.  **Universe**: If you provide a custom `universe` (background gene list), ensure it is large enough. A small universe can make it statistically difficult to find enrichment.
3.  **P-value Adjustment**: `clusterProfiler` uses multiple testing correction (e.g., BH) by default. Some web tools might use raw p-values or different thresholds, leading to more (but potentially false positive) results. `clusterProfiler` tends to be more conservative and reliable.

### Annotation quality

Sometimes, the `OrgDb` provided by Bioconductor might be outdated compared to the latest online databases (e.g., TAIR for Arabidopsis).

*   **Check Dates**: Always check the metadata of the `OrgDb` object to see the source and date of the annotation.
*   **Custom Annotation**: If the `OrgDb` is too old, consider downloading the latest annotation (e.g., GAF or GOSLIM files) from the organism's primary database and using `enricher()`/`GSEA()` with `buildGOmap()` as described in the [Direct vs indirect GO annotation] section.

## Summary {#clusterprofiler-go-summary}

GO semantic similarity can be calculated by `r Biocpkg("GOSemSim")` [@yu2010]. We can use it to cluster genes/proteins into different clusters based on their functional similarity and can also use it to measure the similarities among GO terms to reduce the redundancy of GO enrichment results.

## Next steps

- For online annotation refreshes, downloads, and release metadata, see [external-service troubleshooting](troubleshooting.qmd#external-services).
- For custom annotations or unsupported organisms, see [universal enrichment](universal-enrichment.qmd#universal-api).
- To reduce redundancy, continue to [GOSemSim](go-semantic-similarity.qmd) and [Bayesian term selection](bayesian-term-selection.qmd#sec-bayes-enrich).
- For plots, use the [visualization map](enrichplot.qmd#sec-enrichplot).

## References {-}
