# GSON: the knowledge representation and exchange layer {#sec-gson}

```{r}
#| label: gson-load-packages
#| echo: false
#| message: false
#| warning: false
#| cache: false

source("_common.R")
suppressMessages(library(gson))
suppressMessages(library(clusterProfiler))
```

A functional enrichment workflow has two different jobs:

1. **the analysis engine** tests whether gene sets are enriched; and
2. **the knowledge layer** defines which terms, pathways, diseases, or other
   biological concepts belong to which genes.

These jobs should not be confused. `enricher()` and `GSEA()` provide enrichment
interfaces, whereas a gene-set resource provides the biological knowledge that
those interfaces test. The `r Biocpkg("gson")` package supplies a standard way
to represent that knowledge and to move it between data sources, analysis
functions, and machines.

## Chapter overview

| Aspect | GSON knowledge layer |
|---|---|
| **Questions** | How can gene-set knowledge be represented, versioned, shared, subset, validated, combined, and passed to enrichment interfaces? |
| **Input** | Term-to-gene mappings plus metadata: gene set names, species, gene identifier type, gene set ID domain, source, release, and access date. |
| **Methods** | Knowledge packaging, inspection, subsetting, serialization to `.gson` and GMT, merging or paralleling several resources, and integration with ORA/GSEA. GSON is not a statistical test. |
| **Main functions** | `gson()`, `gson_GO()`, `gson_KEGG()`, `gson_WP()`, `read.gmt()`, `read.gmt.wp()`, `write.gson()`, `read.gson()`, `write.gmt()`, `x[i]`, `gsonList()`, `gson_union()`, `validate_gson()`, `gson_url()`, `browseGS()`. |
| **Output** | A `GSON` object, a `.gson` or GMT file, a `GSONList` of parallel resources, or one enrichment result per source. |
| **Main limitations** | GSON stores and moves knowledge. It does not convert identifiers, does not compute ontology structure, and cannot resolve two sources that use the same gene set ID for different gene sets. |

## What GSON represents

GSON means **Gene Set Object Notation**. The `r Biocpkg("gson")` package
defines the `GSON` class for storing gene-set information together with the
metadata needed to interpret and reuse it. A GSON object is therefore more
than a two-column `TERM2GENE` table: it is a portable knowledge resource that
can retain information about the source, species, identifiers, and the version
or release from which the gene sets were obtained.

This separation is useful in practice. The same enrichment engine can be used
with GO, KEGG, WikiPathways, a disease collection, a cell-marker catalogue, or
a project-specific annotation table. Only the knowledge object changes; the
statistical question and the downstream result workflow can remain the same.

::: {.callout-note}
GSON is a **knowledge representation and exchange layer**, not a new
statistical enrichment test. ORA, GSEA, network-aware enrichment, and other
methods still determine how evidence is evaluated. GSON determines the
knowledge that is evaluated.
:::

An object of the class holds three mapping tables and nine metadata slots. The
tables are the knowledge; the slots say what the tables mean.

| Component | Columns | What it records |
|---|---|---|
| `gsid2gene` | `gsid`, `gene` | Gene set membership. The only required table. |
| `gsid2name` | `gsid`, `name` | Human-readable name of each gene set. |
| `gene2name` | `gene`, `name` | Symbol or display name of each gene. |

| Slot | Records |
|---|---|
| `gsname` | Name of the collection, for example `GO`, `KEGG`, `WikiPathways`. |
| `species` | Organism the genes belong to. |
| `keytype` | Identifier domain of the **genes**, for example `ENTREZID`, `SYMBOL`. |
| `gsidtype` | Identifier domain of the **gene set IDs**, for example `GO`, `KEGG`, `REAC`, `WP`. |
| `version` | Release of the resource. |
| `accessed_date` | Date the resource was obtained. |
| `urlpattern` | Template that turns a gene set ID into the web address of that gene set. |
| `schema_version` | Version of the `.gson` file format, as declared by the producer. |
| `info` | Free-text note for the producer's own use. |

`keytype` and `gsidtype` describe two different namespaces. A collection of KEGG
pathways whose members are Entrez gene IDs has `gsidtype = "KEGG"` and
`keytype = "ENTREZID"`. Keeping them apart means a consumer does not have to
infer which namespace the gene set IDs come from.

## Obtaining a GSON object

`clusterProfiler` and related packages expose helper functions whose names start
with `gson_`. The helpers do not share one argument convention, so a script that
passes arguments by name needs the right ones. `gson_GO()` reads a local organism
package; the other two download their source, so their result should be frozen
once, as described under [Writing and reading a collection](#gson-writing).

| Helper | Example call | Argument conventions |
|---|---|---|
| `gson_GO()` | `gson_GO(OrgDb = "org.Hs.eg.db", keytype = "ENTREZID", ont = "BP")` | `OrgDb` is an object or the name of an organism package; `keytype` and `ont` default to `"ENTREZID"` and `"BP"` |
| `gson_KEGG()` | `gson_KEGG(species = "hsa")` | The gene argument is `keyType`, with a capital `T`; `KEGG_Type = "KEGG"` selects pathway membership |
| `gson_WP()` | `gson_WP(organism = "Homo sapiens")` | The argument is `organism`, not `species`, and takes the Latin name |

```{r}
#| label: gson-download
#| eval: false

library(gson)
library(clusterProfiler)

kk_gson <- gson_KEGG(species = "hsa")
wp_gson <- gson_WP(organism = "Homo sapiens")
go_gson <- gson_GO(OrgDb = "org.Hs.eg.db", keytype = "ENTREZID", ont = "BP")
```

A collection that is already on disk is the offline route, and the `gson`
installation ships one so that this chapter can run without a network. The
WikiPathways GMT parser `read.gmt.wp()` returns a `GSON` when asked for one.

```{r}
#| label: gson-read-wp
wpfile <- system.file(
  "extdata",
  "wikipathways-20220310-gmt-Homo_sapiens.gmt",
  package = "gson"
)
wp <- read.gmt.wp(wpfile, output = "GSON")
wp
```

The parser filled in what the file stated and what the format declares: the
source, the organism, the release, the gene set ID domain `WP`, and a
`urlpattern` that resolves a `WP` identifier to its WikiPathways page.

The same file parsed by the generic reader is the long `term`/`gene` table that
`enricher()` accepts as `TERM2GENE`, which is the difference between the two
representations: a data frame carries the membership and nothing else.

```{r}
#| label: gson-read-plain
head(read.gmt(wpfile))
```

## Constructing a collection from your own tables

`gson()` builds an object from user-supplied data frames. The arguments are the
three tables named above, not `TERM2GENE` and `TERM2NAME`. Each table is read by
position: the first two columns become `gsid` and `gene`, `gsid` and `name`, or
`gene` and `name`, so the names you use in your own data frame do not matter and
any further columns are dropped.

```{r}
#| label: custom-gson

gsid2gene <- data.frame(
  term = c("GS1", "GS1", "GS1", "GS2", "GS2"),
  gene = c("10196", "7157", "3689", "7157", "5728")
)
gsid2name <- data.frame(
  term = c("GS1", "GS2"),
  name = c("apoptosis signalling", "ribosomal proteins")
)
gene2name <- data.frame(
  gene = c("10196", "7157", "3689", "5728"),
  symbol = c("BAX", "TP53", "MDM2", "NPM1")
)

custom <- gson(
  gsid2gene = gsid2gene,
  gsid2name = gsid2name,
  gene2name = gene2name,
  species = "Homo sapiens",
  gsname = "Project-curated",
  version = "2026-10",
  accessed_date = "2026-10-08",
  keytype = "ENTREZID",
  gsidtype = "GS"
)
custom
```

Identifiers are stored as character. A producer that passes a factor gets
character columns in the result, which matters because a factor carries the
levels it no longer uses: a grouping step over such a column yields one empty
gene set per unused level, and the phantom sets then show up in `length()`,
`names()`, and every file the object is written to. A double column is stored by
its digits, formatted without scientific notation, so one million is
`"1000000"` rather than `"1e+06"`.

::: {.callout-tip}
Use a plain annotation table for a one-off exploratory test; promote it to a
versioned GSON object when it will be shared, reused, combined with other
resources, or cited in a report.
:::

## Inspecting and subsetting a collection

A `GSON` behaves like a container of gene sets. `length()` counts gene sets,
`names()` lists their IDs in the order the collection holds them, `x[[i]]`
returns the genes of one gene set, and `as.data.frame()` gives the full
membership table with the names joined in.

```{r}
#| label: gson-inspect
length(wp)
head(names(wp))
wp[["WP4685"]][1:5]
head(as.data.frame(custom))
```

`x[i]` returns a smaller `GSON`. Taking a subset used to mean filtering
`gsid2gene` by hand and keeping `gsid2name` and `gene2name` in step with it,
which is easy to get wrong in exactly the direction a validator reports: a
`gene2name` row for a gene that is no longer a member of any kept gene set. The
subset method prunes all three tables together and carries the metadata over.

```{r}
#| label: gson-subset
by_name <- wp[c("WP4685", "WP5043", "WP179")]
by_position <- wp[1:3]
by_logical <- wp[names(wp) %in% c("WP4685", "WP2840")]
length(by_name)
names(by_position)
length(by_logical)
by_name@urlpattern
```

Gene set IDs, positions, and logical vectors all select, and `x[]` returns `x`.
An ID the object does not hold, or an index out of bounds, is an error that names
what is missing. `x[, j]` is also an error, since a `GSON` has no second
dimension.

## Checking an object

Two checks answer two different questions.

The class enforces its structural contract when an object is created, so a table
that is not a data frame, lacks the documented columns, or holds identifiers that
are not character is refused rather than stored.

```{r}
#| label: gson-validity
tryCatch(
  methods::new("GSON", gsid2gene = data.frame(gsid = 1, gene = 2)),
  error = conditionMessage
)
```

`validate_gson()` then asks about the content, and reports instead of refusing,
because an object that breaks a content rule still has to be constructible so
that it can be inspected. Duplicated identifiers, missing cross-references among
the three tables, empty tables, a `keytype` or `gsidtype` left at a placeholder
such as `UNKNOWN`, a `gsidtype` that the object's own gene set IDs contradict,
and a `schema_version` this package cannot read are all reported.

```{r}
#| label: gson-validate

defective <- gson(
  gsid2gene = data.frame(
    gsid = c("GO:1", "GO:1", "GO:2"),
    gene = c("1", "2", NA)
  ),
  gsid2name = data.frame(gsid = "GO:1", name = "one"),
  gene2name = data.frame(gene = "99", name = "dangling"),
  keytype = "UNKNOWN",
  gsidtype = "KEGG"
)
validate_gson(defective, error = FALSE)
validate_gson(wp)
```

There is deliberately no dictionary of accepted `keytype` or `gsidtype` values.
`MicrobiomeProfiler`, for example, writes `kegg_orthology`, `eggNOG_OG`, `taxid`,
and `hmdb_id` as `keytype`, none of which is a `keytypes()` value of any
Bioconductor annotation package, so a list held by `gson` would reject the
collections the ecosystem itself produces. What the package can verify is an
object against itself, so only a prefix the identifiers actually carry is
compared. Nothing is rewritten: a value the object declares is reported as it
stands, because the reader that knows what a file says is not the one that
decides what the producer meant.

## Writing and reading a collection {#gson-writing}

`write.gson()` stores a collection as JSON and `read.gson()` reads it back. The
round trip is exact, so a frozen file can be treated as the object it was written
from. The JSON keys the memberships by gene set ID and carries the names and the
metadata beside them; it is plain JSON, so a collection can also be read by tools
outside R.

```{r}
#| label: gson-roundtrip
#| cache: false

f <- tempfile(fileext = ".gson")
write.gson(custom, file = f)
identical(read.gson(f), custom)
writeLines(readLines(f))
```

GMT is the interoperable format: `read.gmt()` parses it and `write.gmt()`
produces it from a collection. The GMT form keeps the membership and the gene
set name and drops the rest of the metadata, so a `.gmt` file is for exchange
and a `.gson` file is for the versioned copy. `write.gmt()` also sorts the gene
sets alphabetically, so the three selected below are written in a different
order from the one the collection held them in; a ranked or curated order
survives only in the `.gson` file.

```{r}
#| label: gmt-roundtrip
gmt_file <- tempfile(fileext = ".gmt")
write.gmt(by_name, file = gmt_file)
writeLines(substr(readLines(gmt_file), 1, 60))
dim(read.gmt(gmt_file))
```

Memberships that carry no gene set ID or no gene are skipped when a collection is
written, with a warning that says how many were omitted, and `write.gmt()` falls
back to the gene set ID when `gsid2name` holds no usable name for a set. A `NA`
that reached a file would come back as a gene genuinely named `NA` and be handed
on to the enrichment routines.

A `.gson` file is written by other programs as often as by this one, so
`read.gson()` names the file and the shape it found instead of passing on what
the JSON parser happened to notice. A file whose memberships are not keyed by
gene set ID, whose tables hold a string where an array belongs, or that cannot be
parsed at all is an error that says which file and which member; a local path
that does not exist is caught before the parser is asked.

## Locating the source database

`urlpattern` is a claim that each gene set ID resolves to a web page.
`gson_url()` builds those URLs and `browseGS()` opens one in a browser, which is
the step that turns a row of an enrichment table back into evidence to read.

```{r}
#| label: gson-urls
gson_url(wp)[1:2]
gson_url(wp, c("WP100", "WP106"))
```

```{r}
#| label: gson-browse
#| eval: false

browseGS(wp, "WP4685")
```

The URL comes only from what the object declares. `gson` ships no table of other
databases' websites, so a producer that wants its gene sets browsable has to name
the pattern; `ReactomePA::gson_Reactome()` currently passes `urlpattern = NULL`.
An object that declares none gets a vector of `NA` from `gson_url()` and an error
from `browseGS()`.

```{r}
#| label: gson-no-urlpattern
#| warning: true

gson_url(custom)[1]
```

This is also where `gsidtype` earns its place. A consumer that knows the domain
of the identifiers can choose the right resource and the right pattern, whereas
one that does not has to infer the domain from the shape of the IDs, which is a
guess about someone else's data.

## Combining collections

Two operations answer two different questions, and choosing between them is a
design decision rather than a stylistic one.

`gsonList()` keeps the resources **separate** and names them, so one input is
tested against each knowledge base and the sources stay distinguishable in the
result.

```{r}
#| label: gson-list
gsons <- gsonList(wikipathways = wp, project = custom)
names(gsons)
class(gsons[1])
```

The component names come from the `gsname` slot of each object, not from the
argument labels of the call, so `gsonList(wikipathways = wp)` is still
`"WikiPathways"`. Set `gsname` when the identity of a resource must survive into
the result table. Subsetting keeps the `GSONList` class, so `gsons[1]` still
prints as a list of collections.

`gson_union()` merges several collections into **one** membership table, which is
what an analysis needs when a single test should see all the annotations at once,
for example GO, KEGG, and REAC as one `enricher()` input.

```{r}
#| label: gson-union-same-source
half <- gson_union(wp[1:50], wp[51:100])
length(half)
half@gsname
```

Two objects that name the same source are the same collection, so a gene set that
both of them carry is joined into one set holding every gene either gives it. Two
files of one download therefore merge rather than collide.

A gene set ID that two *different* sources both use is refused, because the
merged table would say that one ID names two sets of genes, and every result read
from it would be quietly wrong.

```{r}
#| label: gson-union-conflict
left <- gson(
  gsid2gene = data.frame(gsid = c("X1", "X2"), gene = c("1", "2")),
  gsid2name = data.frame(gsid = c("X1", "X2"), name = c("set one", "set two")),
  gsname = "SourceA",
  gsidtype = "X"
)
right <- gson(
  gsid2gene = data.frame(gsid = c("X1", "X3"), gene = c("7", "8")),
  gsid2name = data.frame(gsid = c("X1", "X3"), name = c("other one", "set three")),
  gsname = "SourceB",
  gsidtype = "X"
)
tryCatch(gson_union(left, right), error = conditionMessage)
```

`.conflict` decides how a collision is resolved. `"prefix"` qualifies every ID
with the source it comes from, `"rename"` qualifies only the colliding ones and
leaves IDs that already begin with their source alone, and the `gsid2name` rows
follow the IDs they describe.

```{r}
#| label: gson-union-prefix
merged <- gson_union(left, right, .conflict = "prefix")
names(merged)
merged@gsname
```

The metadata merges by what the merged object can still honestly say. `species`,
`keytype`, and `schema_version` have to agree, because one membership table cannot
mean two organisms, two identifier types, or two file schemas. `gsname`,
`gsidtype`, `version`, `accessed_date`, and `info` are joined with `" + "` when
the sources name themselves differently. `urlpattern` is kept only when every
object declares the same one and no ID moved, since a pattern that resolves part
of the collection to the wrong page is worse than no pattern.

## Passing a collection to the engines

The enrichment layer consumes a `GSON` through the `gson` argument of `enricher()`
and `GSEA()`, which takes the place of `TERM2GENE`. Over representation analysis
tests a thresholded gene vector against the collections.

```{r}
#| label: gson-enricher
#| message: false

gene <- unique(c(wp[["WP4685"]][1:25], wp[["WP5043"]]))
ora <- enricher(
  gene = gene,
  gson = wp,
  minGSSize = 5,
  maxGSSize = 500,
  pvalueCutoff = 0.05
)
head(as.data.frame(ora)[, c("ID", "Description", "Count", "pvalue")])
```

For the ranked-list case the separation is the same: the ranking is the
statistical input and the `GSON` object supplies the gene sets it is tested
against.

```{r}
#| label: gson-gsea
#| message: false
#| warning: false

data(geneList, package = "DOSE")
gsea <- GSEA(
  geneList = geneList,
  gson = wp,
  minGSSize = 10,
  maxGSSize = 500,
  eps = 0,
  pvalueCutoff = 0.05,
  seed = 123
)
head(as.data.frame(gsea)[, c("ID", "Description", "setSize", "NES", "pvalue")])
```

Passed a `gsonList`, `enricher()` returns one result per source, keyed by the
names of the list, and the table or figure can be faceted by knowledge base. The
KEGG-specific wrappers take the same objects through `organism`.

```{r}
#| label: gson-wrappers
#| eval: false

kk <- enrichKEGG(gene = gene, organism = kk_gson)
kk2 <- gseKEGG(geneList = geneList, organism = kk_gson)
res_multi <- enricher(gene = gene, gson = gsons)
```

The engine layer reads the same container directly: `enrichit` provides
`ora_gson()`, `gsea_gson()`, `nsea_gson()`, and `mnsea_gson()`. `GOSemSim`
accepts a `GSON` of GO annotations as the `annoDb` of `godata()` from version
2.39.3, so a frozen collection can enter the semantic-similarity layer as well as
the enrichment layer.

The wrapper form and the generalized form serve the same reproducibility goal:
the analysis uses a local, explicit knowledge resource instead of silently
retrieving a changing online database during every run.

## Versioning, metadata, and reproducibility

Knowledge bases change independently of the code used to analyze them. A
reproducible enrichment result therefore requires more than the R script and
p-value cutoff. At minimum, preserve:

- the GSON file or an immutable copy of the source tables;
- the resource name (`gsname`) and the release (`version`) or download date;
- the species;
- `keytype` for the genes and `gsidtype` for the gene set IDs;
- the source URL or the package that produced the resource, and `urlpattern` when
  the gene sets should be browsable;
- `schema_version` as the producer wrote it;
- any filtering, ID conversion, or deduplication applied to the annotation;
- the `clusterProfiler`, `gson`, and related package versions;
- the analysis method, background universe, thresholds, and random seeds where
  applicable; and
- the names of a `gsonList` when several resources were tested in parallel, or the
  merge policy when `gson_union()` joined them.

The object carries most of this itself, which is the point of the format: a
`.gson` file that states its source, organism, release, and identifier domains
does not depend on the notebook that produced it to explain what it is. Run
`validate_gson()` on a resource received from elsewhere before analyzing it, and
keep the report with the analysis record.

The simplest local-freezing workflow is to write the GSON object once and read it
during analysis:

```{r}
#| label: gson-reproducible-freeze
#| eval: false

kk_gson <- gson_KEGG(species = "hsa")
write.gson(kk_gson, file = "knowledge/kegg_hsa.gson")

kk_gson <- read.gson("knowledge/kegg_hsa.gson")
```

For larger projects, keep the GSON files under version control or in an archival
storage system alongside a small manifest. The manifest should state which file
was used for which analysis and should not rely only on a mutable database URL. If
the project uses `gsonList`, save every component and the list of names, not only
the final enrichment table.

One thing the format does not preserve is the row order of `gsid2gene` when a
collection is not already grouped by gene set ID, because a `.gson` file keys the
memberships by `gsid` and reading one regroups the table. The gene sets, and the
order of the genes within each set, survive; the interleaving of the rows does
not. Files written before `gsidtype` existed read it back as `NULL`, and a file
that carries it still opens in older releases of the package, which ignore the
unknown key.

## Where GSON fits in the toolchain

GSON makes the boundaries between the layers explicit:

```text
   custom table, GMT file, or remote knowledge source
                          |
        gson() / read.gmt() / read.gmt.wp() / gson_*() helpers
                          |
                GSON object  ---  validate_gson()
                          |
            +-------------+-------------+
            |                           |
   gsonList (parallel)         gson_union (merged)
            |                           |
   enricher() / GSEA() /  ora_gson() / gsea_gson() / godata()
            |                           |
   enrichment results           gson_url() / browseGS()
            |                           |
   plots and interpretation      manual evidence check
```

This architecture has three practical consequences:

1. **The analysis engine can be reused.** The same ORA or GSEA logic can work
   across multiple knowledge resources.
2. **The knowledge resource can be frozen.** A saved GSON file separates a
   publication or pipeline run from future changes in an online database, and
   reads back as the object it was written from.
3. **The resource can be shared.** A custom annotation can be distributed as a
   named, documented object rather than as an unexplained pair of data frames.

GSON is therefore the bridge between biological knowledge and the rest of the
biomedical knowledge-mining toolkit. It allows GO, KEGG, WikiPathways,
project-specific annotations, and other compatible resources to enter a common
workflow without pretending that they are the same biological evidence.

## Summary

- `gson` defines the `GSON` class for portable gene-set knowledge and metadata.
  Three mapping tables hold the knowledge; the metadata slots say what the
  identifiers, organism, release, and source mean.
- `gson_KEGG(species = )`, `gson_WP(organism = )`, and
  `gson_GO(OrgDb = , keytype = , ont = )` create resource-specific objects, and
  `read.gmt.wp(file, output = "GSON")` creates one from a file already on disk.
- `gson()` builds an object from `gsid2gene`, `gsid2name`, and `gene2name`, taking
  each table by position and storing every identifier as character.
- `length()`, `names()`, `as.data.frame()`, `x[[i]]`, and `x[i]` inspect and
  subset a collection; subsetting prunes the three tables together and keeps the
  metadata.
- `validate_gson()` reports what an object holds or claims without carrying it,
  while the class itself refuses a structurally invalid object at creation.
- `write.gson()` and `read.gson()` round-trip exactly; `write.gmt()` and
  `read.gmt()` exchange memberships and names without the metadata.
- `gson_url()` and `browseGS()` resolve gene sets to the pages of their source
  database, but only when the object declares a `urlpattern`.
- `gsonList()` tests one input against several separate resources; `gson_union()`
  merges several resources into one membership table and refuses an ambiguous
  gene set ID.
- `enricher()` and `GSEA()` take these objects through `gson`, the KEGG wrappers
  through `organism`, and `enrichit` and `GOSemSim` consume the same container.
- Resource identity, version, metadata, and the saved GSON files are part of a
  reproducible analysis record.

The next steps are to inspect the enrichment result, compare or reduce
redundant terms, and choose a visualization appropriate to the question. See
the chapters on result objects, semantic similarity, and `enrichplot` for those
operations.

## Next steps

- To analyze custom or unsupported annotations, see [universal enrichment](universal-enrichment.qmd#universal-api).
- For knowledge-source examples, see [GO](go-enrichment.qmd#clusterprofiler-go) or [KEGG](kegg-enrichment.qmd#clusterprofiler-kegg).
- To score the terms a collection holds, see [GO semantic similarity](go-semantic-similarity.qmd#GOSemSim), where `godata()` also accepts a `GSON` of GO annotations.
- For result visualization, continue to the [visualization map](enrichplot.qmd#sec-enrichplot).

## References {-}
