# GSON: the knowledge representation and exchange layer {#sec-gson}

```{r}
#| label: gson-load-packages
#| echo: false
#| message: false
#| warning: false
#| cache: false

source("_common.R")
```

A functional enrichment workflow has two different jobs:

1. **the analysis engine** tests whether gene sets are enriched; and
2. **the knowledge layer** defines which terms, pathways, diseases, or other
   biological concepts belong to which genes.

These jobs should not be confused. `enricher()` and `GSEA()` provide enrichment
interfaces, whereas a gene-set resource provides the biological knowledge that
those interfaces test. The `r Biocpkg("gson")` package supplies a standard way
to represent that knowledge and to move it between data sources, analysis
functions, and machines.

## Chapter overview

| Aspect | GSON knowledge layer |
|---|---|
| **Questions** | How can gene-set knowledge be represented, versioned, shared, combined, and passed to enrichment interfaces? |
| **Input** | Term/gene mappings plus metadata such as names, species, identifier type, source, and release/version. |
| **Methods** | Knowledge packaging, exchange, serialization, named multi-resource collections, and integration with ORA/GSEA; GSON is not a statistical test. |
| **Main functions** | `gson_KEGG()`, `gson_WP()`, `gson_GO()`, `gson()`, `write.gson()`, `read.gson()`, `gsonList()`, `enricher()`. |
| **Output** | Portable `GSON` objects/files and named `gsonList` collections for generalized enrichment. |
| **Main limitations** | Converter availability and metadata arguments vary by package version; verify API and retain provenance. Combined resources remain biologically distinct. |

## What GSON represents

GSON means **Gene Set Object Notation**. The `r Biocpkg("gson")` package
defines the `GSON` class for storing gene-set information together with the
metadata needed to interpret and reuse it. A GSON object is therefore more
than a two-column `TERM2GENE` table: it is a portable knowledge resource that
can retain information about the source, species, identifiers, and the version
or release from which the gene sets were obtained.

This separation is useful in practice. The same enrichment engine can be used
with GO, KEGG, WikiPathways, a disease collection, a cell-marker catalogue, or
a project-specific annotation table. Only the knowledge object changes; the
statistical question and the downstream result workflow can remain the same.

::: {.callout-note}
GSON is a **knowledge representation and exchange layer**, not a new
statistical enrichment test. ORA, GSEA, network-aware enrichment, and other
methods still determine how evidence is evaluated. GSON determines the
knowledge that is evaluated.
:::

## Obtaining a GSON object

`clusterProfiler` and related packages expose helper functions whose names
usually start with `gson_`. The available helpers depend on the installed
package versions. R's completion facilities and package help are the safest
way to discover the current set of helpers.

The following helpers are used in the existing knowledge-mining workflows:

- `gson_KEGG()` downloads KEGG pathway and module data for an organism;
- `gson_WP()` obtains WikiPathways data;
- `gson_GO()` obtains Gene Ontology data from an `OrgDb` and ontology choice;
- `gson::gson()` can be used to construct a GSON object from user-provided
  tables.

For example, a KEGG knowledge resource can be downloaded and frozen locally:

```{r}
#| label: gson-kegg-local
#| eval: false

library(gson)
library(clusterProfiler)

# Download KEGG data for human.
kk_gson <- gson_KEGG(species = "hsa")

# Write the GSON object to a portable file.
write.gson(kk_gson, file = "kegg_hsa.gson")

# Read the same knowledge resource back on another machine.
kk_gson <- read.gson("kegg_hsa.gson")
```

The download step may require network access, but the analysis step does not
have to repeat that download once the GSON file has been transferred to the
analysis machine. This is particularly useful on compute servers with limited
network access and in projects that need to preserve the knowledge release used
for a publication.

## The relationship with `enricher()` and `GSEA()`

A GSON object is consumed by the enrichment layer. The existing generalized
enrichment interface passes a GSON object, or a collection of GSON objects, to
`enricher()` through the `gson` argument:

```{r}
#| label: gson-enricher
#| eval: false

library(gson)
library(clusterProfiler)

data(geneList, package = "DOSE")
gene <- names(geneList)[abs(geneList) > 2]

kk_gson <- gson_KEGG(species = "hsa")

# The GSON object supplies the terms and gene membership tested by enricher().
res <- enricher(gene = gene, gson = kk_gson)
```

The same conceptual separation applies to GSEA: a ranked gene list is the
statistical input, while the GSON object supplies the gene sets against which
the ranking is tested. The existing KEGG workflow documents that a GSON object
or a saved GSON file can be used by `enricher()` and `GSEA()` through the
`gson` interface (or by parsing the resource in the installed version).

Because the exact generalized `GSEA()` argument names can vary with the
installed `clusterProfiler`/`gson` versions, this chapter does not hard-code an
unverified call signature. Check the local help before running a generalized
GSEA workflow:

```{r}
#| label: gson-gsea-api-check
#| eval: false

library(gson)
library(clusterProfiler)

?GSEA
?gson::gson

# The installed version documents how to pass a GSON object or file to GSEA.
# Do not replace this with an unverified TERM2GENE or gson argument without
# checking the help for the versions used in the analysis.
```

For a KEGG-specific workflow, the existing API also supports passing a local
GSON object to the KEGG wrappers:

```{r}
#| label: gson-kegg-wrappers
#| eval: false

# `kk_gson` is created with gson_KEGG() above, or read from a saved file.
kk <- enrichKEGG(gene = gene, organism = kk_gson)
kk2 <- gseKEGG(geneList = geneList, organism = kk_gson)
```

The wrapper form and the generalized form serve the same reproducibility goal:
the analysis uses a local, explicit knowledge resource instead of silently
retrieving a changing online database during every run.

## Combining knowledge bases with `gsonList`

Many biological questions are not well served by one database. GO may provide
broad functional coverage, KEGG may provide pathway and module structure, and
WikiPathways may contain a complementary collection of curated pathways. A
`gsonList` combines named GSON objects so that the same input can be evaluated
against several knowledge bases in one generalized enrichment workflow.

The following example is the multi-resource pattern documented in the existing
chapter. The exact download requirements depend on the current releases of the
individual resources:

```{r}
#| label: gson-list
#| eval: false

library(gson)
library(clusterProfiler)

# Prepare or download the individual knowledge resources.
kegg_gson <- gson_KEGG(species = "hsa")
wp_gson <- gson_WP(species = "Homo sapiens")
go_gson <- gson_GO(
    OrgDb = org.Hs.eg.db,
    keyType = "ENTREZID",
    ont = "BP"
)

# Keep stable names so that the output can be traced back to its source.
gsons <- gsonList(
    kegg = kegg_gson,
    wp = wp_gson,
    go = go_gson
)

data(geneList, package = "DOSE")
gene <- names(geneList)[abs(geneList) > 2]

# Run one generalized enrichment request across the named resources.
res <- enricher(gene = gene, gson = gsons)
```

When `enricher()` receives a `gsonList`, the result is organized by the named
input resources. Each component corresponds to one knowledge base and can be
accessed by name or by position. Naming the resources explicitly is preferable
to relying on numeric positions because it makes reports and later audits
readable.

The results can also be visualized together. `enrichplot` provides
`autofacet()` for faceting a multi-resource result by knowledge base:

```{r}
#| label: gson-list-plot
#| eval: false

library(enrichplot)

dotplot(res) + autofacet()
```

A combined display is useful for comparison, but it does not make the
knowledge bases interchangeable. A GO term, a KEGG pathway, and a WikiPathways
pathway may differ in granularity, annotation coverage, update schedule, and
biological interpretation. The database identity should remain visible in the
result and in the figure.

## Custom knowledge bases

The main benefit of a generalized knowledge layer is that the same workflow can
also use annotations that are not shipped as a standard organism database.
Examples include:

- a custom annotation generated for a non-model organism;
- a project-specific collection of marker genes;
- a locally curated pathway catalogue;
- a versioned export from an external resource; and
- a reduced or specialised ontology used for a particular experiment.

The `gson::gson()` constructor is the documented entry point for manually
creating a GSON object from user data. The exact constructor arguments should
be checked against the installed package because the accepted metadata fields
are part of the package API and may evolve:

```{r}
#| label: custom-gson
#| eval: false

library(gson)

# Build TERM2GENE and (optionally) TERM2NAME tables from the project resource.
# term2gene <- data.frame(term = ..., gene = ...)
# term2name <- data.frame(term = ..., name = ...)

# The constructor is intentionally shown as a checked API boundary. Consult
# ?gson::gson for the current argument names and required metadata fields.
# custom_gson <- gson::gson(
#     TERM2GENE = term2gene,
#     TERM2NAME = term2name,
#     ...
# )
```

If the current `gson::gson()` interface is not suitable for a particular
custom table, the same annotations can still be used through the standard
`TERM2GENE`/`TERM2NAME` interface of `enricher()` or `GSEA()`. In that case,
record the source, identifier type, organism, filtering rules, and release in
the analysis metadata even though the table is not yet packaged as GSON.

A useful rule is:

> Use a plain annotation table for a one-off exploratory test; promote it to a
> versioned GSON object when it will be shared, reused, combined with other
> resources, or cited in a report.

## Versioning, metadata, and reproducibility

Knowledge bases change independently of the code used to analyze them. A
reproducible enrichment result therefore requires more than the R script and
p-value cutoff. At minimum, preserve:

- the GSON file or an immutable copy of the source tables;
- the resource name and release or download date;
- species and identifier type;
- the source URL or package that produced the resource;
- any filtering, ID conversion, or deduplication applied to the annotation;
- the `clusterProfiler`, `gson`, and related package versions;
- the analysis method, background universe, thresholds, and random seeds where
  applicable; and
- the names used in a `gsonList` when multiple resources are combined.

The simplest local-freezing workflow is to write the GSON object once and read
it during analysis:

```{r}
#| label: gson-reproducible-freeze
#| eval: false

library(gson)

# On a machine with access to the remote resource:
kk_gson <- gson_KEGG(species = "hsa")
write.gson(kk_gson, file = "knowledge/kegg_hsa.gson")

# On the analysis server or in a later rerun:
kk_gson <- read.gson("knowledge/kegg_hsa.gson")
```

For larger projects, keep the GSON files under version control or an archival
storage system alongside a small manifest. The manifest should state which file
was used for which analysis and should not rely only on a mutable database URL.
If the project uses `gsonList`, save every component and the list of names, not
only the final enrichment table.

## Where GSON fits in the toolchain

GSON makes the boundaries between the layers explicit:

```text
custom or remote knowledge source
                ↓
       GSON object / GSON file
                ↓
       gsonList (optional)
                ↓
  enricher() / GSEA() / package wrappers
                ↓
 enrichment result objects and plots
                ↓
  interpretation and reproducible report
```

This architecture has three practical consequences:

1. **The analysis engine can be reused.** The same ORA or GSEA logic can work
   across multiple knowledge resources.
2. **The knowledge resource can be frozen.** A saved GSON file separates a
   publication or pipeline run from future changes in an online database.
3. **The resource can be shared.** A custom annotation can be distributed as a
   named, documented object rather than as an unexplained pair of data frames.

GSON is therefore the bridge between biological knowledge and the rest of the
biomedical knowledge-mining toolkit. It allows GO, KEGG, WikiPathways,
project-specific annotations, and other compatible resources to enter a common
workflow without pretending that they are the same biological evidence.

## Summary

- `gson` defines the `GSON` class for portable gene-set knowledge and metadata.
- Helper functions such as `gson_KEGG()`, `gson_WP()`, and `gson_GO()` create
  resource-specific GSON objects.
- `write.gson()` and `read.gson()` support local freezing and offline reuse.
- `gsonList()` combines named knowledge bases for generalized enrichment.
- `enricher()` consumes a GSON object or `gsonList` through the generalized
  knowledge interface; GSEA uses the same separation between ranked evidence
  and gene-set knowledge, with the exact installed-version call documented by
  `GSEA` help.
- `gson::gson()` provides the entry point for custom knowledge resources, but
  its constructor arguments should be checked in the installed version.
- Resource identity, version, metadata, and the saved GSON files are part of a
  reproducible analysis record.

The next steps are to inspect the enrichment result, compare or reduce
redundant terms, and choose a visualization appropriate to the question. See
the chapters on result objects, semantic similarity, and `enrichplot` for those
operations.

## Next steps

- To analyze custom or unsupported annotations, see [universal enrichment](universal-enrichment.qmd#universal-api).
- For knowledge-source examples, see [GO](go-enrichment.qmd#clusterprofiler-go) or [KEGG](kegg-enrichment.qmd#clusterprofiler-kegg).
- For result visualization, continue to the [visualization map](enrichplot.qmd#sec-enrichplot).

## References {-}
