# Welcome {.unnumbered}

```{r}
#| label: setup-badges
#| cache: false
#| include: false

library(emojifont)

badge_version <- function(pkg, color = "green") {
    v <- packageVersion(pkg)
    url <- paste0("https://github.com/YuLab-SMU/", pkg)
    badger::badge_custom(pkg, v, color, url)
}
```

```{r}
#| label: load-common
#| echo: false
#| message: false
#| cache: false

source("_common.R")
```

# Introduction {-}

Biomedical knowledge mining is not a single test and it is not a collection of unrelated R packages. It is a workflow that turns molecular evidence into organized, inspectable biological knowledge:

```text
experiment → identifiers and evidence → enrichment → knowledge sources
          → evidence integration → figures → biological interpretation
```

This book documents that workflow as a toolkit. The packages remain visible because they provide the implementation, but the main navigation follows the decisions a researcher makes rather than the history of package development.

## The toolkit at a glance {-}

The system has six cooperating layers:

1. **Prepare the evidence** — gene vectors, ranked lists, backgrounds, identifiers, and custom annotations.
2. **Choose an analysis engine** — ORA, GSEA, comparison, network-aware, weighted, or multi-omics enrichment.
3. **Connect biological knowledge** — GO, KEGG, Reactome, disease, MeSH, pathway collections, and user-defined gene sets.
4. **Organize the evidence** — result objects, semantic similarity, redundancy reduction, representative terms, and PPI networks.
5. **Communicate the result** — summary plots, gene–term networks, GSEA views, pathway views, and comparative figures.
6. **Interpret and report** — evidence-guided biological interpretation, reproducible reports, and optional AI assistance.

The common objects passed between these layers are deliberately simple: a vector of identifiers, a named numeric ranking, a `TERM2GENE` table, a GSON knowledge object, or an enrichment result object.

## Start with the route that matches your question {-}

| If you have… | Start with… | The question it answers |
|---|---|---|
| A differential-gene list | [A complete first workflow](first-workflow.qmd) | How do I run and inspect a first ORA/GSEA analysis? |
| A list but are unsure how to prepare it | [The data contract](data-contract.qmd) | Which IDs, ranking, background, and annotation format are required? |
| Several possible analysis designs | [Choose a route](choose-a-route.qmd) | Should I use ORA, GSEA, `compareCluster`, network, or custom enrichment? |
| Several biological databases | [GSON knowledge packaging](gson.qmd) | How can I combine knowledge sources without changing the statistical engine? |
| An existing enrichment result | [Organize and integrate evidence](03-evidence-integration.qmd) | How do I reduce redundancy, compare terms, and find supporting genes? |
| A figure or report to make | [Visualize and communicate results](04-visualization.qmd) | Which plot answers which biological question? |
| A package or function name | [Package and capability map](capability-map.qmd) | Where does this function fit in the system? |

## The first complete workflow {-}

A useful first pass is deliberately short:

```text
prepare IDs and universe
        ↓
choose ORA or GSEA
        ↓
connect GO, KEGG, or another knowledge source
        ↓
inspect the result object
        ↓
plot the strongest themes
        ↓
check evidence before writing a biological conclusion
```

The [first workflow chapter](first-workflow.qmd) uses stable demonstration data and links each step to the deeper chapters. It is the recommended entry point for readers who want to get a result before learning every available method.

## Package roles in the system {-}

| Role in the workflow | Main packages or modules | What the reader gets |
|---|---|---|
| High-level workflow and knowledge-aware entry points | `clusterProfiler` | Consistent functions and result objects |
| Statistical enrichment engine | `enrichit` | ORA, GSEA, network-aware, weighted, and multi-omics methods |
| Knowledge representation and exchange | `gson` | Versioned, metadata-aware knowledge objects and combined knowledge bases |
| Specialized knowledge connectors | `DOSE`, `ReactomePA`, `meshes` | Disease, Reactome, and MeSH analysis |
| Semantic similarity | `GOSemSim`, `DOSE`, `meshes` | Similarity among terms, genes, and gene clusters |
| Result visualization | `enrichplot` | Reusable plots for ORA, GSEA, comparison, and networks |
| Genomic-coordinate entry point | `ChIPseeker` | Peak annotation and genomic coordination enrichment |
| Optional interpretation layer | `interpret()` and supported model providers | Evidence synthesis and narrative drafting, with explicit uncertainty |

The packages are therefore modules in one tool system. A user can enter through a package-specific function, but should still be able to follow the same path from input to result, figure, and interpretation.

## Version badges {-}

`r badge_version("clusterProfiler")`
`r badge_version("DOSE")`
`r badge_version("enrichplot")`
`r badge_version("GOSemSim")`
`r badge_version("meshes")`
`r badge_version("ReactomePA")`

## Book structure {-}

- **Part 0 — Start here:** data contract, first workflow, and route selection.
- **Part 1 — Choose the analysis engine:** ORA, GSEA, comparison, network-aware, weighted, and multi-omics enrichment.
- **Part 2 — Connect biological knowledge:** GSON, GO, KEGG, Reactome, disease, MeSH, pathway collections, and custom annotations.
- **Part 3 — Organize and integrate evidence:** result objects, semantic similarity, term selection, and PPI networks.
- **Part 4 — Visualize and communicate:** summary plots, networks, GSEA views, comparisons, and publication-ready figures.
- **Part 5 — Interpret and report:** evidence-guided interpretation and optional AI-assisted reporting.
- **Part 6 — Recipes and reference:** task-oriented workflows, troubleshooting, capability maps, and reference material.

## Citation {-}

If you use this software suite in published research, cite the papers most relevant to the methods and packages used:

1. __G Yu__<sup>\*</sup>. [Background bias in functional enrichment analysis: Insights from clusterProfiler](https://doi.org/10.59717/j.xinn-life.2025.100181). __*The Innovation Life*__. 2026, 4(1):100181.
2. __G Yu__<sup>\*</sup>. [Thirteen years of clusterProfiler](https://doi.org/10.1016/j.xinn.2024.100722). __*The Innovation*__. 2024, 5(6):100722.
3. S Xu<sup>#</sup>, E Hu<sup>#</sup>, Y Cai<sup>#</sup>, Z Xie<sup>#</sup>, X Luo<sup>#</sup>, L Zhan, W Tang, Q Wang, B Liu, R Wang, W Xie, T Wu, L Xie, __G Yu__<sup>\*</sup>. [Using clusterProfiler to characterise Multi-Omics Data](https://www.nature.com/articles/s41596-024-01020-z). __*Nature Protocols*__. 2024, 19(11):3292–3320.
4. Q Wang<sup>#</sup>, M Li<sup>#</sup>, T Wu, L Zhan, L Li, M Chen, W Xie, Z Xie, E Hu, S Xu, __G Yu__<sup>\*</sup>. [Exploring epigenomic datasets by ChIPseeker](https://doi.org/10.1002/cpz1.585). __*Current Protocols*__. 2022, 2(10):e585.
5. T Wu<sup>#</sup>, E Hu<sup>#</sup>, S Xu, M Chen, P Guo, Z Dai, T Feng, L Zhou, W Tang, L Zhan, X Fu, S Liu, X Bo<sup>\*</sup>, **G Yu**<sup>\*</sup>. clusterProfiler 4.0: A universal enrichment tool for interpreting omics data. **_The Innovation_**. 2021, 2(3):100141. [doi:10.1016/j.xinn.2021.100141](https://doi.org/10.1016/j.xinn.2021.100141)
6. __G Yu__<sup>\*</sup>. Gene Ontology Semantic Similarity Analysis Using GOSemSim. In: Kidder B. (eds) Stem Cell Transcriptional Networks. __*Methods in Molecular Biology*__. 2020, 2117:207–215.
7. __G Yu__<sup>\*</sup>. Using meshes for MeSH term enrichment and semantic analyses. __*Bioinformatics*__. 2018, 34(21):3766–3767. [doi:10.1093/bioinformatics/bty410](https://doi.org/10.1093/bioinformatics/bty410)
8. __G Yu__, QY He<sup>\*</sup>. ReactomePA: an R/Bioconductor package for reactome pathway analysis and visualization. __*Molecular BioSystems*__. 2016, 12(2):477–479. [doi:10.1039/C5MB00663E](https://doi.org/10.1039/C5MB00663E)
9. __G Yu__<sup>\*</sup>, LG Wang, and QY He<sup>\*</sup>. ChIPseeker: an R/Bioconductor package for ChIP peak annotation, comparison and visualization. __*Bioinformatics*__. 2015, 31(14):2382–2383. [doi:10.1093/bioinformatics/btv145](https://doi.org/10.1093/bioinformatics/btv145)
10. __G Yu__<sup>\*</sup>, LG Wang, GR Yan, QY He<sup>\*</sup>. DOSE: an R/Bioconductor package for Disease Ontology Semantic and Enrichment analysis. __*Bioinformatics*__. 2015, 31(4):608–609. [doi:10.1093/bioinformatics/btu684](https://doi.org/10.1093/bioinformatics/btu684)
11. __G Yu__, LG Wang, Y Han and QY He<sup>\*</sup>. clusterProfiler: an R package for comparing biological themes among gene clusters. __*OMICS: A Journal of Integrative Biology*__. 2012, 16(5):284–287. [doi:10.1089/omi.2011.0118](https://doi.org/10.1089/omi.2011.0118)
12. __G Yu__, F Li, Y Qin, X Bo<sup>\*</sup>, Y Wu, S Wang<sup>\*</sup>. GOSemSim: an R package for measuring semantic similarity among GO terms and gene products. __*Bioinformatics*__. 2010, 26(7):976–978. [doi:10.1093/bioinformatics/btq064](https://doi.org/10.1093/bioinformatics/btq064)

## Contributing {-}

The source code is hosted at <https://github.com/YuLab-SMU/biomedical-knowledge-mining-book/>. Feedback, bug reports, examples, and improvements to the workflow are welcome through issues and pull requests.
