# Evidence-guided biological interpretation {#evidence-guided-interpretation}

Statistical enrichment results are evidence to inspect, not biological conclusions by themselves. This chapter describes how to move from enriched terms, core genes, networks, expression trends, and prior knowledge to a transparent interpretation. AI assistance is presented as an optional evidence-synthesis layer after those checks, not as a replacement for the enrichment analysis or literature review.

<a id="ai-assisted-biological-interpretation"></a>

```{r}
#| label: load-packages
#| echo: false
#| message: false
#| cache: false

source("_common.R")

## Load the project-local credentials before deciding which LLM chunks to run.
## The key itself is never printed or written to the rendered book.
if (file.exists(".env") && requireNamespace("dotenv", quietly = TRUE)) {
    dotenv::load_dot_env(".env")
}

## The LLM examples below run only when an API key is available. GitHub does
## not expose repository secrets to workflows triggered from forked
## repositories, and a local build has no key either, so the key doubles as the
## switch: no key, examples skipped, book still renders.
has_ai <- nzchar(Sys.getenv("DEEPSEEK_API_KEY"))
if (has_ai && requireNamespace("aisdk", quietly = TRUE)) {
    aisdk::set_model("deepseek:deepseek-v4-flash")
}

## interpret() does not raise when the API call fails: it warns and returns a
## placeholder whose `confidence` is "None". opts_chunk$set(warning = FALSE)
## hides that warning, so without an explicit check a missing or invalid key
## would silently publish an empty report. Fail loudly instead.
##
## A second, quieter failure mode exists: when the model answers but the reply
## cannot be parsed (for example a reasoning model that exhausts `max_tokens`
## and returns truncated JSON), interpret() falls back to a placeholder with
## `confidence == "Low"` and no parsed fields. A successful call always carries
## a `narrative`, so requiring it separates the two cases.
interpret_failed <- function(res) {
    identical(res$confidence, "None") ||
        is.null(res$overview) ||
        grepl("^LLM call failed", res$overview) ||
        (identical(res$confidence, "Low") && is.null(res$narrative))
}

annotation_failed <- function(res) {
    is.null(res$cell_type) ||
        is.null(res$confidence) ||
        is.null(res$reasoning) ||
        identical(res$confidence, "None")
}
```

Enrichment results identify terms and genes associated with the input, but they do not by themselves establish a mechanism. Interpretation therefore starts with the result object, core genes, background, database provenance, and any supporting expression or network evidence.

## The `interpret` function

`interpret()` accepts `enrichResult`, `gseaResult`, or `compareClusterResult` objects and asks an LLM to organize the supplied evidence into a structured report. The report is a draft for evidence review, not a replacement for the enrichment test or literature validation.

The function requires an API key for a provider supported by `aisdk`. If `model` is not supplied, it uses the current model from `aisdk::get_model()` / `aisdk::set_model()`; an explicit `model` argument overrides that default for one call. The examples below use DeepSeek through `DEEPSEEK_API_KEY`, but other `aisdk` providers can be used.

```{r}
#| label: interpret-setup
#| eval: !expr has_ai
library(aisdk)

# Credentials come from the environment; a local .env file works too
if (file.exists(".env") && requireNamespace("dotenv", quietly = TRUE)) {
    dotenv::load_dot_env()
}

# Set a session-wide default model. Any provider supported by aisdk works here;
# in CI the key is supplied as the DEEPSEEK_API_KEY repository secret.
set_model("deepseek:deepseek-v4-flash")

# Check the current default model
get_model()
```

```{r}
#| label: interpret-basic
#| eval: !expr has_ai
#| results: asis
library(clusterProfiler)
library(DOSE)

# A self-contained enrichment result: `edo` is built offline from the demo gene
# list shipped with DOSE, so the example depends neither on a previous chapter
# (`edo` is defined in the enrichplot chapter) nor on a network service.
data(geneList, package = "DOSE")
de <- names(geneList)[abs(geneList) > 1.5]
edo <- enrichDO(de)

# Basic usage
# 'edo' is your enrichment result object
res <- interpret(edo)

if (interpret_failed(res)) {
    stop("interpret() returned no report -- check the API key and quota:\n",
         res$overview)
}
cat(render_interpretation_blocks(res), sep = "\n")
```

### Tasks and Inputs

`interpret()` supports three tasks:

+ **`task = "interpretation"`**: organize enrichment evidence into a biological narrative;
+ **`task = "annotation"`**: combine marker genes and enrichment terms for cell-type annotation;
+ **`task = "phenotyping"`**: propose a state or phenotype label for a group.

The input can be one result object, a list of compatible objects, or a `compareCluster` result processed by cluster. For example, `list(go_res, kegg_res)` combines compatible knowledge sources, while a `compareCluster` result returns one report per cluster. The returned fields are structured so that reports can be checked, compared, and rendered consistently.

For single-cell data, the marker table can come from `Seurat::FindAllMarkers()` or another marker-calling method. Keep the cluster and gene columns, use the same identifier namespace in the marker and CellMarker tables, and pass the grouped markers through `compareCluster()` before annotation. The executable example below uses the repository-local CellMarker table so that it does not depend on a separate count matrix or a pasted annotation report. A complete Seurat marker workflow is available in the [pbmc3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html#assigning-cell-type-identity-to-clusters).

### Cell Type Annotation

The cell-type example reads the local `datasets/Cell_marker_Human.xlsx` marker table, builds four marker profiles, runs `compareCluster()` with `enricher()`, and passes the result to `interpret(task = "annotation")` when `DEEPSEEK_API_KEY` is available.

```{r}
#| eval: !expr has_ai
#| label: interpret-annotation
#| cache: true
#| results: hide

library(clusterProfiler)
library(readxl)
library(dplyr)

cell_marker_path <- file.path("datasets", "Cell_marker_Human.xlsx")
if (!file.exists(cell_marker_path)) {
    stop("The cell-marker dataset is missing: ", cell_marker_path)
}

cell_marker <- readxl::read_excel(cell_marker_path)
TERM2GENE_cell <- cell_marker |>
    transmute(
        cell_name = .data[["Cell name"]],
        marker = toupper(.data[["Marker"]])
    ) |>
    filter(
        !is.na(cell_name), !is.na(marker),
        nzchar(cell_name), nzchar(marker)
    ) |>
    distinct()

# A compact marker panel keeps the example transparent and the prompt small.
marker_sets <- list(
    T_cell = c("CD3D", "CD3E", "TRBC1", "IL7R", "LTB", "MAL", "CCR7", "LST1"),
    B_cell = c("CD79A", "MS4A1", "CD37", "CD79B", "HLA-DRA", "CD74", "BANK1"),
    Monocyte = c("LYZ", "S100A8", "S100A9", "CTSS", "FCN1", "LGALS3", "TYROBP", "LST1"),
    NK_cell = c("NKG7", "GNLY", "PRF1", "GZMB", "KLRD1", "FCGR3A", "XCL1", "XCL2")
)
marker_table <- do.call(
    rbind,
    lapply(names(marker_sets), function(cluster) {
        data.frame(
            cluster = cluster,
            gene = marker_sets[[cluster]],
            stringsAsFactors = FALSE
        )
    })
)

cell_marker_enrichment <- compareCluster(
    gene ~ cluster,
    data = marker_table,
    fun = enricher,
    TERM2GENE = TERM2GENE_cell,
    pvalueCutoff = 1,
    qvalueCutoff = 1,
    minGSSize = 2,
    maxGSSize = 500
)

annotation_reports <- interpret(
    cell_marker_enrichment,
    task = "annotation",
    context = paste(
        "Human single-cell marker panels. Each cluster is a compact set of",
        "top marker genes and was tested against the local CellMarker database.",
        "Use the enrichment terms as supporting evidence, but prefer markers",
        "that distinguish closely related cell types. Return a concise label",
        "and explain which supplied markers and cell-marker terms support it."
    ),
    n_pathways = 12,
    max_tokens = 4096,
    temperature = 0.1
)

if (inherits(annotation_reports, "interpretation")) {
    annotation_reports <- list(Default = annotation_reports)
}
if (!is.list(annotation_reports) || length(annotation_reports) == 0) {
    stop("DeepSeek returned no cell-type annotation reports.")
}
for (report in annotation_reports) {
    if (annotation_failed(report)) {
        stop(
            "DeepSeek cell-type annotation failed for cluster ",
            if (!is.null(report$cluster)) report$cluster else "(unknown)",
            ": ",
            if (!is.null(report$overview)) report$overview else "no diagnostic returned"
        )
    }
}
```

The result below is generated from the live object returned by DeepSeek. It is
rendered as an HTML block, not Markdown headings: the preview is readable
in the chapter, while the raw `print()` output remains available in a collapsed
section without entering the book's table of contents.

```{r}
#| eval: !expr has_ai
#| label: print-annotation-result
#| results: asis

cat(render_interpretation_blocks(annotation_reports), sep = "\n")
```

::: {.callout-note}
When the book is rendered without `DEEPSEEK_API_KEY`, the two chunks above are
skipped and the rest of the chapter still renders. When the key is present, a
render performs a real DeepSeek call; keep the cache enabled for ordinary builds
and clear `interpretation_cache/` deliberately when a fresh report is wanted.
:::

<!-- The previous pasted annotation report was removed; the live DeepSeek output above is the maintained example. -->

## Extended evidence inputs

`interpret_agent()` applies three sequential stages to an interpretation request:

1. **Term curation** selects and organizes the supplied enrichment terms in the
   stated biological context; it does not rerun the enrichment test.
2. **Evidence inspection** combines the result object with optional gene-level
   changes and interaction data to identify genes or modules that support each
   term.
3. **Narrative drafting** writes the checked inputs into the structured
   interpretation fields used by the report and plotting methods.

The implementation refers to these stages as **Cleaner**, **Detective**, and **Storyteller**. They are components of one shared-model pipeline, not separate statistical analyses. The stages share one evidence bundle, and their output remains a model-generated draft that should be checked against the original result object.

```{r}
#| label: interpret-agent
#| eval: false
context <- "scRNA-seq analysis of CD8+ T cells in tumor microenvironment, comparing exhausted and naive states."
res <- interpret_agent(edo, context = context)
```

## PPI and expression evidence

Set `add_ppi = TRUE` to request protein interactions from STRING. Pass a named `gene_fold_change` vector to retain expression direction in the interpretation. These inputs add network and expression evidence; they do not establish pathway activity or causality.

```r
# Prepare a named vector of fold changes
gene_list <- c("CD8A" = 2.5, "PDCD1" = 1.8, "GZMB" = 3.2)

res <- interpret(
    edo,
    task = "interpretation",
    add_ppi = TRUE,
    gene_fold_change = gene_list
)
```

## Prior labels and hierarchical inputs

For cell-type annotation, `prior` supplies an existing label for comparison. The returned annotation can agree with, refine, or disagree with that label; inspect the markers and enrichment terms before accepting a change.

```{r}
#| label: interpret-prior-knowledge
#| eval: false
my_priors <- c("Cluster1" = "T cells")
res <- interpret(edo, prior = my_priors, task = "annotation")
```

`interpret_hierarchical()` accepts major- and minor-cluster enrichment objects together with a mapping. It constrains each minor-cluster interpretation to the supplied major-cluster context; the mapping itself must be checked against the experimental design.

```r
cluster_mapping <- c(
    "SubCluster1_1" = "MajorCluster1",
    "SubCluster1_2" = "MajorCluster1",
    "SubCluster2_1" = "MajorCluster2"
)

res_hier <- interpret_hierarchical(
    x_minor = enrich_minor,
    x_major = enrich_major,
    mapping = cluster_mapping
)
```

## When enrichment returns no terms

When a result has no enriched terms, `interpret_agent()` can be given the input genes and optional PPI data. The implementation warns that no significant pathways were found, passes the raw genes to the Detective and Storyteller stages, and marks the returned object with `data_source = "gene_list_only"`. The output should therefore be reported as a gene-based hypothesis; its confidence is taken from the returned object rather than assigned a fixed value.

```r
res <- interpret_agent(tough_genes, add_ppi = TRUE)
```

## Visualization of interpretation objects

The `plot()` method displays relationships stored in an interpretation object. When the returned object contains them, the plot can distinguish key drivers, activation or inhibition labels, and physical interactions. These labels are model outputs and require review against the supplied expression and interaction evidence.

```r
plot(res)
```

The complete hand-off is:

```text
enrichment result → optional fold-change/PPI/prior evidence
                  → interpretation object → report or plot
```

The final report and plot are views of the interpretation object. They do not
constitute a second enrichment test, and interaction edges remain external
supporting evidence.

## Operational checks

`context` should identify the experiment, biological setting, contrast direction, and any relevant cell-type or tissue resolution. `interpret()` cannot infer these details from pathway names alone.

A weak context names only the assay:

```r
context <- "scRNA-seq of T cells"
```

A strong context answers six questions in one or two sentences:

| Question | Example |
| :--- | :--- |
| Species and tissue | human breast tumour, bulk tissue |
| Disease / biological setting | breast carcinoma, tumour grade |
| Experimental design | 200 samples from the MAINZ cohort |
| Contrast tested, and its direction | grade III vs. grade I; positive values are higher in grade III |
| Cell type / composition | bulk tissue, no cell-type resolution |
| Hypothesis under test | progression involves loss of epithelial differentiation and ECM remodelling |

Written out, that is still only a few lines:

```r
context <- paste(
  "Human breast-cancer transcriptome (MAINZ cohort, bulk tumour tissue, 200",
  "samples). Genes are ranked by the log2 ratio of geometric-mean expression in",
  "grade III versus grade I tumours, so positive values are higher in the",
  "higher-grade tumours. No cell-type resolution is available. Hypothesis:",
  "progression from grade I to grade III involves loss of epithelial",
  "differentiation and increased extracellular-matrix remodelling."
)
```

Two rules of thumb follow. First, state the **direction of the contrast** --
"A vs. B" without saying which side is up is a common way to get a report with
the regulation backwards. Second, **do not put the answer in the context**: if
you write "we expect apoptosis to be suppressed", the model will often agree
with you regardless of the data. Describe the design, not the conclusion you are
hoping for. (If your data are from single cells, add the cell type or the
compartment the cluster belongs to; for bulk tissue, saying that no cell-type
resolution exists is more useful than omitting the point.)

### Choosing a Model and Switching Provider

Model selection is delegated to `aisdk`. `interpret()` itself has no model
registry and no provider logic: when `model` is left at its default (`NULL`) it
uses whatever `aisdk` currently considers the default, and `model = "provider:model"`
overrides that default for a single call.

```r
library(aisdk)

# Session-wide default, used by every interpret() call that omits `model`
set_model("deepseek:deepseek-v4-flash")
get_model()
```

Calls that omit `model` use the session-wide setting selected with `set_model()`. Pass `model` for a one-off provider or model.

```r
res <- interpret(edo, model = "openai:gpt-4o")
```

Credentials follow the same rule. `aisdk` reads each provider's key from an
environment variable: `OPENAI_API_KEY`, `ANTHROPIC_API_KEY` and `GEMINI_API_KEY`
for the three providers bundled with `aisdk`, and the corresponding variable for
the adapters in `aisdk.providers` (for example `DEEPSEEK_API_KEY`, which is what
this chapter uses). Put the key in a project `.Renviron` or export it in the
shell; `interpret()` never takes a key argument. If you never call
`set_model()`, `aisdk` falls back to its built-in default, `openai:gpt-4o` --
which is why the examples you may have seen elsewhere appear to use GPT-4o by
default.

`aisdk` ships a catalogue of the providers and models it knows about:

```r
list_models()            # every provider and model aisdk knows about
list_models("openai")    # just one provider
```

Model choice affects latency, cost, context capacity, and output variability. Fix the model when comparing reports, because different models can produce different summaries from the same input.

### Multi-Model and Consensus Workflows

There is **no built-in consensus or ensemble mode** in `interpret()` or
`interpret_agent()`. A call goes to exactly one model. `interpret_agent()` is a
multi-*agent* pipeline, not a multi-*model* one: its three agents share a single
model.

What you *can* do today is run the same input through several models and compare
the reports yourself. Because `model =` is a per-call argument, this needs no
extra machinery:

```r
res_deepseek <- interpret(edo, model = "deepseek:deepseek-v4-flash")
res_gpt      <- interpret(edo, model = "openai:gpt-4o")
```

Each call is a separate, billable request, and the two reports will not be
identical even for the same model (see below). A pragmatic middle ground is to
use one model to write the report and a second, independent call only to
*criticise* it: ask a different model whether every gene and pathway named in
the narrative is actually present in the input table. That is a manual
workflow, but it catches the most common failure mode -- plausible-sounding
genes that were never in the data. For provider failover, `aisdk` provides
`generate_text_fallback()` for its own generation calls, but `interpret()` does
not expose a fallback argument.

### Practical Guidance and Pitfalls

**Non-determinism.** `interpret()` samples at `temperature = 0.3` by default, so
two calls with identical input will differ in wording and occasionally in
emphasis. This is expected, not a bug. It has a direct consequence for
reproducible builds: Quarto's chunk cache (`cache: true`) is what keeps the
published page stable -- the LLM call runs once and its output is reused on
later renders. Clear the cache and you will get a different report. If you need
the exact same text in two places, cache the object and reuse it.

**Unsupported entities.** The model is instructed to ground claims in the supplied table, but it can still produce incorrect output. Before quoting a report, check that every gene named in the narrative appears in the `geneID` (or, for GSEA, `core_enrichment`) column and every pathway named appears in `Description`. If `add_ppi = TRUE`, identify the cited edges as STRING evidence, not measurements from the experiment.

**`n_pathways`.** Only the top `n_pathways` terms (default 20) are sent to the
model, ordered by adjusted p-value (`p.adjust`, falling back to `pvalue`).
Raising it gives the model more to work with but dilutes the signal and enlarges
the prompt; lowering it forces a tighter focus. For a `compareClusterResult`,
`interpret()` splits the result by cluster and returns one report per cluster,
so the number of calls -- and the cost -- scales with the number of clusters,
not with `n_pathways`.

**`max_tokens`.** The default is 8192. Reasoning models spend part of that
budget deliberating before they emit the JSON, and if they run out, the reply is
truncated and cannot be parsed; the symptom is a warning mentioning
`finish_reason: length` and a placeholder report. If you hit it, raise
`max_tokens` (the diagnostic suggests 16384 or more) or switch to a model that
does not reason before answering.

**Seeing the cost.** `interpret()` returns only the parsed report; token usage
is not part of the returned object. Run with `verbose = TRUE` to print the raw
response and the prompt/completion/total token counts -- the practical way to
watch spend while you tune `n_pathways` and `max_tokens`.

**`add_ppi`.** Setting `add_ppi = TRUE` queries STRING for the genes in the top
terms and injects the edge list into the prompt, which lets the model identify
functional modules from individual genes. It requires network access, adds
latency, and is a second failure point; the query is wrapped so that a STRING
outage degrades to "no network evidence" instead of an error. Leave it at its
default (`FALSE`) unless module-level reasoning is what you need.

**`gene_fold_change`.** Passing a named numeric vector of log fold changes adds
the direction of regulation to the prompt. Note the limitation: the helper
intersects your vector with the genes in the result's `geneID` column, and a
`gseaResult` table has no `geneID` column (it carries `core_enrichment`
instead). So `gene_fold_change` is used for over-representation results
(`enrichResult`, `compareClusterResult`) and silently ignored for GSEA results
-- which is harmless, because a GSEA result already encodes direction in the
sign of its enrichment score.

**Ranking.** For over-representation results the terms are ordered by
`p.adjust`; for GSEA the direction the model sees is the sign of `NES`. If your
GSEA ranked list was built so that positive means "up in the condition of
interest", say so in `context` -- otherwise the model has to guess the
orientation.

Treat the output as an unverified draft. Check every gene and pathway named in the narrative against the input table; then assess whether each mechanistic statement has supporting evidence.

## Worked Example: From Enrichment to a Report

This example builds a GSEA result offline from the `geneList` shipped with `DOSE`, passes a defined `context` to `interpret()`, and renders the returned report. The `interpret_agent()` example above is disabled because it makes multiple model calls.

```{r}
#| label: interpret-worked-example
#| eval: !expr has_ai
#| results: asis
library(clusterProfiler)
library(org.Hs.eg.db)

data(geneList, package = "DOSE")

# A real enrichment result, computed offline: GSEA over GO cellular components.
# geneList is the DOSE demo list -- log2 ratios of grade III vs. grade I tumours
# from the MAINZ breast-cancer cohort.
gse <- gseGO(
    geneList      = geneList,
    OrgDb         = org.Hs.eg.db,
    keyType       = "ENTREZID",
    ont           = "CC",
    minGSSize     = 15,
    maxGSSize     = 100,
    eps           = 0,
    pvalueCutoff  = 0.05,
    pAdjustMethod = "BH"
)

# A context that names the species, tissue, design, contrast and hypothesis.
context <- paste(
    "Human breast-cancer transcriptome (MAINZ cohort, bulk tumour tissue).",
    "Genes are ranked by the log2 ratio of geometric-mean expression in grade",
    "III versus grade I tumours, so positive values are higher in the",
    "higher-grade tumours. Hypothesis: progression from grade I to grade III",
    "involves loss of epithelial differentiation and increased",
    "extracellular-matrix remodelling."
)

report <- interpret(gse, context = context, n_pathways = 15)

if (interpret_failed(report)) {
    stop("interpret() returned no report -- check the API key and quota:\n",
         report$overview)
}

cat(render_interpretation_blocks(report), sep = "\n")
```

The printed report follows the schema described at the top of this chapter. Read
it in this order:

1.  **1. Overview** -- the model's one-paragraph summary. If it is vague, the
    `context` was probably too vague.
2.  **2. Regulatory Drivers** and **3. Key Mechanisms** -- the "who" and the
    "how". Check every named gene against your input table.
3.  **5. Hypothesis** -- the "so what". This is the part worth arguing with.
4.  **6. Narrative Draft** -- a paragraph written for a Results or Discussion
    section. It is a draft: edit it, do not paste it.
5.  **7. Refined Regulatory Network** -- populated only when the model can
    support edges; `plot(report)` draws the same network with `ggtangle` when it
    is available.

## Next steps

- For the preceding evidence objects, see [evidence integration](03-evidence-integration.qmd).
- For a reproducible hand-off, record the analysis in the [reference appendix](appendix.qmd).
- Keep optional AI assistance after statistical checking and external evidence review.

