# AI-Assisted Biological Interpretation

```{r}
#| label: load-packages
#| echo: false
#| message: false
#| cache: false

source("_common.R")

## The LLM examples below run only when an API key is available. GitHub does
## not expose repository secrets to workflows triggered from forked
## repositories, and a local build has no key either, so the key doubles as the
## switch: no key, examples skipped, book still renders.
has_ai <- nzchar(Sys.getenv("DEEPSEEK_API_KEY"))

## interpret() does not raise when the API call fails: it warns and returns a
## placeholder whose `confidence` is "None". opts_chunk$set(warning = FALSE)
## hides that warning, so without an explicit check a missing or invalid key
## would silently publish an empty report. Fail loudly instead.
##
## A second, quieter failure mode exists: when the model answers but the reply
## cannot be parsed (for example a reasoning model that exhausts `max_tokens`
## and returns truncated JSON), interpret() falls back to a placeholder with
## `confidence == "Low"` and no parsed fields. A successful call always carries
## a `narrative`, so requiring it separates the two cases.
interpret_failed <- function(res) {
    identical(res$confidence, "None") ||
        is.null(res$overview) ||
        grepl("^LLM call failed", res$overview) ||
        (identical(res$confidence, "Low") && is.null(res$narrative))
}
```

Traditional enrichment analysis typically results in a list of significant pathways or GO terms. While statistically sound, these lists often leave researchers asking, "**So what?**" What is the underlying biological mechanism? Who are the key drivers? Is this a pro-survival or pro-death signal?

To bridge the gap between statistical results and biological insights, `clusterProfiler` introduces an AI-powered interpretation module. By leveraging Large Language Models (LLMs) and a multi-agent system, `clusterProfiler` can now act as a virtual bioinformatician, converting dry enrichment lists into coherent, evidence-based biological narratives.

## The `interpret` Function

The core function for this feature is `interpret()`. It accepts enrichment results (e.g., from `enrichKEGG`, `enrichGO`, or `compareCluster`) and uses an LLM to generate a structured report.

To use this feature, you need to configure an API key for a provider supported by `aisdk`. Model selection is now delegated to `aisdk`: if `model` is not specified, `interpret()` uses the current default model managed by `aisdk::get_model()` / `aisdk::set_model()`, while an explicit `model = ...` argument overrides that default for a single call. In practice, this usually means loading credentials from your `.env` file or setting the relevant environment variables before calling `set_model()`. For provider configuration details, see the `aisdk` documentation. The examples on this page use DeepSeek, whose key `aisdk` reads from the `DEEPSEEK_API_KEY` environment variable; any other provider supported by `aisdk` can be used instead.

```{r}
#| label: interpret-setup
#| eval: !expr has_ai
library(aisdk)

# Credentials come from the environment; a local .env file works too
if (file.exists(".env") && requireNamespace("dotenv", quietly = TRUE)) {
    dotenv::load_dot_env()
}

# Set a session-wide default model. Any provider supported by aisdk works here;
# in CI the key is supplied as the DEEPSEEK_API_KEY repository secret.
set_model("deepseek:deepseek-v4-flash")

# Check the current default model
get_model()
```

```{r}
#| label: interpret-basic
#| eval: !expr has_ai
library(clusterProfiler)
library(DOSE)

# A self-contained enrichment result: `edo` is built offline from the demo gene
# list shipped with DOSE, so the example depends neither on a previous chapter
# (`edo` is defined in the enrichplot chapter) nor on a network service.
data(geneList, package = "DOSE")
de <- names(geneList)[abs(geneList) > 1.5]
edo <- enrichDO(de)

# Basic usage
# 'edo' is your enrichment result object
res <- interpret(edo)

if (interpret_failed(res)) {
    stop("interpret() returned no report -- check the API key and quota:\n",
         res$overview)
}
print(res)
```

### Tasks and Inputs

`interpret()` is not just for explaining enrichment results. It breaks down LLM capabilities into three distinct tasks:

+ **`task = "interpretation"`**: (Default) Converts enrichment results into a mechanistic narrative suitable for publication (What -> So What).
+ **`task = "annotation"`**: Performs cell type annotation for single-cell clusters using both marker genes and enrichment terms as evidence.
+ **`task = "phenotyping"`**: Assigns a "state/phenotype label" to a group (e.g., "Pro-inflammatory" or "Senescent-like").

To strengthen the evidence, `interpret()` supports "evidence synthesis" from multiple sources:

+ **Single Object**: `enrichResult`, `gseaResult`, or `compareClusterResult`.
+ **Multiple Objects**: A `list()` of results (e.g., `list(kegg_res, go_res)` or `list(cellmarker_res, go_res)`).
+ **Batch Processing**: If the input is a `compareCluster` result, it automatically splits by cluster and generates a report for each.

The key features of `interpret()` include:

+ **Prompt Skeleton**: A fixed structure to guide the LLM.
+ **Structured Output**: Enforced structure for parsing, comparison, and batch processing.
+ **Reasoning First**: Encourages "deduction before writing" to avoid merely listing pathway names.

### Cell Type Annotation

For example, we can use `Seurat` to identify marker genes for each cluster in a single-cell RNA-seq dataset. Then we can use `compareCluster` to perform enrichment analysis for each cluster. Finally, we can use `interpret` to annotate cell types based on the enrichment results and marker genes.

```{r}
#| eval: false
#| label: interpret-annotation
## Not run in the book build: needs Seurat plus the pbmc3k 10x matrix
## (data/filtered_gene_bc_matrices/hg19) and the CellMarker xlsx file.
library(Seurat)
dir = "data/filtered_gene_bc_matrices/hg19"
pbmc.data <- Read10X(data.dir = dir)
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", 
                          min.cells=3, min.features=200)
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc <- subset(pbmc,
  subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5
)
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize",
                      scale.factor = 10000)
pbmc <- ScaleData(pbmc)
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst",
                             nfeatures = 2000)
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
pbmc <- RunUMAP(pbmc, dims = 1:10)
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE)

library(dplyr)
topN_marker <- function(markers, n) {
    markers %>%
        group_by(cluster) %>%
        dplyr::filter(avg_log2FC > 1) %>%
        slice_head(n = n) %>%
        ungroup()
}
top20 <- topN_marker(pbmc.markers, 20)
# downloaded from: http://www.bio-bigdata.center/CellMarker_download_files/file/Cell_marker_Human.xlsx
# note: `gene` from FindAllMarkers() is a gene symbol, and so is CellMarker's
# `marker` column -- both must be the same identifier type for the join to work.
cm <- rio::import("Cell_marker_Human.xlsx") 
x <- compareCluster(gene~cluster, data=top20, fun=enricher, TERM2GENE=cm[,c("cell_name", "marker")])
y <- interpret(x, task="annotation")
```

The output `y` is a list of interpretation results, one for each cluster. We can extract the inferred cell types.

```
> sapply(y, \(x) x$cell_type)
                                        0 
                           "Naive T Cell"
                                        1
                     "Classical Monocyte"
                                        2
                            "CD4+ T cell"
                                        3
                      "Follicular B cell"
                                        4
                  "CD8+ Cytotoxic T Cell" 
                                        5
"CD16+ monocyte (Non-classical monocyte)"
                                        6
               "Natural Killer (NK) cell"
                                        7
      "Plasmacytoid Dendritic Cell (pDC)"
                                        8 
                          "Megakaryocyte"
```

This result is highly consistent with the manual annotation from the [Seurat pbmc3k tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html#assigning-cell-type-identity-to-clusters):

| Cluster ID | Markers | Cell Type |
| :--- | :--- | :--- |
| 0 | IL7R, CCR7 | Naive CD4+ T |
| 1 | CD14, LYZ | CD14+ Mono |
| 2 | IL7R, S100A4 | Memory CD4+ |
| 3 | MS4A1 | B |
| 4 | CD8A | CD8+ T |
| 5 | FCGR3A, MS4A7 | FCGR3A+ Mono |
| 6 | GNLY, NKG7 | NK |
| 7 | FCER1A, CST3 | DC |
| 8 | PPBP | Platelet |

The full report provides detailed reasoning, confidence levels, and supporting evidence (markers/pathways) for each cluster assignment, offering transparency and explainability that simple label transfer methods lack.

```{r}
#| eval: false
#| label: print-annotation-result
## Output of the annotation example above, which is not run in the book build.
print(y)
```

::: {.callout-note icon=false}
# Enrichment Interpretation / Annotation Report {.unnumbered .unlisted}

## Cell Type Annotation {.unnumbered .unlisted}

### Cluster: 0 {.unnumbered .unlisted}

**Cell Type:** Naive T Cell
**Confidence:** High

**Reasoning:**
 The cluster is definitively identified as a naive T cell based on the co-expression of canonical pan-T cell markers (CD3D, CD3E) and the master regulator of naive T cell identity, TCF7 (cited in top terms: Naive CD8+ T cell, Naive CD4+ T cell, etc.). The high expression of CCR7, a critical homing receptor for naive and central memory T cells, and LEF1, another Wnt-pathway TF co-operating with TCF7, further solidifies this identity. The enrichment list is dominated by naive and central memory T cell subtypes, with effector/cytotoxic terms ranking lower and lacking their specific markers (e.g., GZMB, PRF1). The presence of both CD4+ and CD8+ associated terms suggests a mixed population or a shared naive state before lineage commitment, but the core identity is naive T cell.

**Supporting Markers/Pathways:**
- CD3D
- CD3E
- TCF7
- CCR7
- LEF1
- NOSIP
- MAL

---

## Cell Type Annotation {.unnumbered .unlisted}

### Cluster: 1 {.unnumbered .unlisted}

**Cell Type:** Classical Monocyte
**Confidence:** High

**Reasoning:**
 The enrichment list contains many related myeloid cell types, but the specific marker gene profile is definitive. The cluster expresses the core classical monocyte signature: high expression of CD14, S100A8, S100A9, FCN1, and LYZ (Top Specific/Marker Genes). While 'Myeloid cell' and 'Macrophage' are top-ranked by p-value, they are broad categories. The specific 'Classical monocyte' term (GeneRatio: 5/20, p.adjust: 1.637326e-07) is strongly supported by its gene list (S100A9/FCN1/CD14/S100A8/LYZ), which perfectly matches the top markers. The cluster lacks definitive markers to distinguish it as a Dendritic Cell (e.g., no FLT3, CD1C, CLEC9A), Macrophage (e.g., low/absent MRC1/CD163), or Neutrophil (e.g., absent MPO, ELANE). The presence of FCN1 and CD14 together is a hallmark of classical monocytes, and the absence of FCGR3A (CD16) argues against non-classical monocytes. 

**Supporting Markers/Pathways:**
- CD14
- S100A8
- S100A9
- FCN1
- LYZ
- CST3
- TYROBP
- MS4A6A

---

## Cell Type Annotation {.unnumbered .unlisted}

### Cluster: 2 {.unnumbered .unlisted}

**Cell Type:** CD4+ T cell
**Confidence:** High

**Reasoning:**
 The cluster is definitively a T cell, as the top enriched term is 'T cell' (p.adjust: 2.86e-17) and the marker list includes the core T-cell receptor complex genes CD3D, CD3E, CD3G, and CD247 (LAT). Among T-cell subtypes, the evidence strongly favors a CD4+ lineage over CD8+. The second most significant term is 'CD4+ T cell' (p.adjust: 2.40e-14), and its gene list (IL32, CD3E, IL7R, CD27, CD3D, TNFRSF4, MAL, CD2, LTB, CD40LG, CD3G) is almost entirely contained within the top 'T cell' markers. Key CD4+ T-cell markers IL7R and CD27 are among the top specific genes. While 'CD8+ T cell' is also enriched, its signature genes (like AQP3) are present but lower in the marker list, and definitive cytotoxic CD8+ markers (e.g., GZMB, PRF1) are absent. The presence of CD40LG and TNFRSF4 (OX40), which are associated with CD4+ T helper and regulatory functions, further supports this assignment. The cluster lacks exclusive markers for NK cells (e.g., NCAM1, KLR genes) or Tregs (FOXP3), though it shows some regulatory association.

**Supporting Markers/Pathways:**
- CD3D
- CD3E
- CD3G
- IL7R
- CD27
- CD40LG
- IL32
- LTB
- TNFRSF4

---

## Cell Type Annotation {.unnumbered .unlisted}

### Cluster: 3 {.unnumbered .unlisted}

**Cell Type:** Follicular B cell
**Confidence:** High

**Reasoning:**
 The top enriched term is 'Follicular B cell' (p.adjust: 2.95e-16), and its gene list contains definitive B cell lineage markers (CD79A, CD79B, MS4A1, BANK1, FCER2, TCL1A) that are also present in the cluster's top specific genes. While other top terms like 'Secretory cell' and 'Classical monocyte' are enriched, they are driven almost exclusively by MHC Class II genes (HLA-DRA, HLA-DRB1, etc.), which are not specific to those cell types but are also expressed by antigen-presenting B cells. The presence of core B cell receptor components (CD79A/B) and mature B cell markers (MS4A1, FCER2, TCL1A) that are absent from monocyte/dendritic cell definitions, combined with the lack of specific monocyte (e.g., CD14, FCGR3A) or secretory cell markers, confirms the identity as a Follicular B cell.

**Supporting Markers/Pathways:**
- CD79A
- MS4A1
- CD79B
- TCL1A
- FCER2
- BANK1
- CD37
- HLA-DRA
- HLA-DRB1
- CD74

---

## Cell Type Annotation {.unnumbered .unlisted}

### Cluster: 4 {.unnumbered .unlisted}

**Cell Type:** CD8+ Cytotoxic T Cell
**Confidence:** High

**Reasoning:**
 The top enriched terms are a mixture of 'Natural killer cell' and various T cell subtypes, indicating shared cytotoxic function. However, the specific marker gene list is definitive. It includes the core T cell receptor complex genes CD3D, CD8A, and CD8B (present in 'CD8+ T cell' and 'Cytotoxic T cell' enrichments), which are lineage-defining for CD8+ T cells and absent from NK cells. While NKG7, PRF1, GZMA, GZMK, and GZMH are shared cytotoxic molecules, the co-expression of CD3D with CD8A/CD8B specifically identifies a cytotoxic T cell lineage. The absence of definitive NK-specific markers (e.g., NCAM1/CD56, KLRD1/CD94, FCGR3A/CD16) from the top marker list, and the presence of the T cell-specific signaling adaptor HCST (DAP10), supports a T cell identity. The 'Cytotoxic CD8+ T cell' enrichment term (GeneRatio: 8/20) provides the most precise functional and lineage match.

**Supporting Markers/Pathways:**
- CD3D
- CD8A
- CD8B
- GZMK
- NKG7
- CCL5
- PRF1
- GZMA
- GZMH
- CST7
- LAG3
- KLRG1

---

## Cell Type Annotation {.unnumbered .unlisted}

### Cluster: 5 {.unnumbered .unlisted}

**Cell Type:** CD16+ monocyte (Non-classical monocyte)
**Confidence:** High

**Reasoning:**
 The top enriched term is 'CD1C-CD141- dendritic cell' (p.adjust: 3.68e-23), but this is likely a misannotation due to shared myeloid markers. The gene list for this term (LST1, CKB, HCK, CSF1R, IFITM3, MS4A7, SERPINA1, LILRB1, CDKN1C, PILRA, FCGR3A, HMOX1, RHOC, LRRC25, SIGLEC10, MS4A4A) is a composite of pan-myeloid and monocyte-specific genes, and lacks definitive dendritic cell markers (e.g., CD1C, CLEC9A, BATF3). The cluster's specific marker list is dominated by canonical markers for non-classical monocytes: FCGR3A (CD16) is the defining marker, supported by MS4A7, CSF1R, and HES4 (enriched in 'CD16+ monocyte' and 'Non-classical monocyte' terms). The absence of T/NK/B cell markers and the presence of macrophage/pan-myeloid genes (e.g., CST3, CTSL) confirm a myeloid lineage, while the high expression of FCGR3A, MS4A7, and HES4 specifically distinguishes the non-classical monocyte subset from classical monocytes, macrophages, and dendritic cells. 

**Supporting Markers/Pathways:**
- FCGR3A (CD16)
- MS4A7
- HES4
- CSF1R
- LST1
- IFITM3
- SIGLEC10
- PILRA

---

## Cell Type Annotation {.unnumbered .unlisted}

### Cluster: 6 {.unnumbered .unlisted}

**Cell Type:** Natural Killer (NK) cell
**Confidence:** High

**Reasoning:**
 The top enriched terms include both 'Natural killer cell' (17/20 genes, p=2.79e-27) and 'Cytotoxic CD4+ T cell' (14/20 genes, p=5.24e-28). While the latter has a slightly better p-value, the discriminatory marker analysis strongly favors NK cells. The cluster's top specific genes include canonical NK markers FGFBP2, SPON2, XCL2, XCL1, SH2D1B, and FCGR3A (CD16a), which are not specific to T cells. Critically, the cluster lacks definitive T-cell lineage markers (e.g., CD3D, CD3E, CD4, CD8A). The shared cytotoxic genes (PRF1, GNLY, GZMB, GZMA, NKG7) are expressed by both NK cells and cytotoxic T cells, but the presence of NK-specific markers and absence of T-cell receptor complex genes confirms NK cell identity.

**Supporting Markers/Pathways:**
- FGFBP2
- SPON2
- XCL2
- XCL1
- SH2D1B
- FCGR3A (CD16a)
- KLRD1 (CD94)
- GNLY
- PRF1
- GZMB
- GZMA
- NKG7

---

## Cell Type Annotation {.unnumbered .unlisted}

### Cluster: 7 {.unnumbered .unlisted}

**Cell Type:** Plasmacytoid Dendritic Cell (pDC)
**Confidence:** High

**Reasoning:**
 The enrichment list contains two distinct dendritic cell lineages: conventional/myeloid DCs (cDC2, CD1C+ DCs) and plasmacytoid DCs (pDC). While the top-ranked term is 'CD1C+_A dendritic cell' (a cDC2 subtype), the cluster's specific marker genes are definitive for pDC identity. The pDC-specific markers LILRA4, CLEC4C (BDCA-2), SERPINF1, and P2RY6 are present in the top marker list and are the defining genes for the 'Plasmacytoid dendritic cell(pDC)' and 'Plasmacytoid dendritic cell' enrichment terms. Critically, the cluster lacks the core, non-overlapping markers for cDC2s: while it expresses CD1C and FCER1A (which can be expressed at low levels in some pDCs), it does NOT express the definitive cDC2 markers CLEC10A (in the cDC2 enrichment term but not in the pDC-specific marker set from the top genes) and CD1C at high specificity relative to pDC markers. The rule of exclusion applies: the top term is a cDC2 type, but the specific marker gene list is dominated by pDC markers and lacks exclusive cDC2 commitment.


**Supporting Markers/Pathways:**
- LILRA4 (ILT7)
- CLEC4C (BDCA-2)
- SERPINF1
- P2RY6
- CLIC2
- SCT
- LRRC26

---

## Cell Type Annotation {.unnumbered .unlisted}

### Cluster: 8 {.unnumbered .unlisted}

**Cell Type:** Megakaryocyte
**Confidence:** High

**Reasoning:**
 The assignment is based on the definitive convergence of enrichment terms and marker genes. The top enriched term is 'Megakaryocyte' (p.adjust: 3.58e-10) with genes SPARC, GNG11, PF4, GP9, ITGA2B, GP1BA. The related terms 'Platelet' and 'Progenitor cell' are lower-ranked and share subsets of these genes (e.g., PF4, GP9, ITGA2B), which is expected as platelets are anucleate fragments of megakaryocytes. The top specific marker gene list is dominated by canonical megakaryocyte/platelet markers (GP9, ITGA2B, GP1BA, PF4, ITGB3, SPARC, GNG11) and lacks definitive markers for other hematopoietic lineages that could challenge this identity (e.g., no CD3, CD19, CD14, ELANE). The 'Progenitor cell' term is likely reflective of the shared SPARC and PF4 expression in some progenitor states, but the presence of terminal differentiation markers like GP1BA and ITGA2B confirms a mature megakaryocyte identity. 

**Supporting Markers/Pathways:**
- GP9
- ITGA2B
- GP1BA
- PF4
- SPARC
- GNG11
- ITGB3
- CLDN5
- CMTM5
- SDPR

---
:::

## The Multi-Agent System

Instead of relying on a single prompt, `clusterProfiler` employs a **Multi-Agent System (MAS)** to ensure accuracy and depth. This system consists of three specialized agents that work in a pipeline:

1.  **Agent Cleaner**: Acts as a curator. It filters out "housekeeping" pathways (e.g., Ribosome, Spliceosome) that may be statistically significant but irrelevant to the specific biological context (e.g., tumor immunology), reducing noise.
2.  **Agent Detective**: Acts as a systems biologist. It analyzes the gene list, looks for **Hub Genes** in Protein-Protein Interaction (PPI) networks, and combines this with Fold Change data to identify **Key Drivers** and infer regulatory mechanisms.
3.  **Agent Storyteller**: Acts as a scientific writer. It synthesizes the findings from the Cleaner and Detective into a logical narrative, distinguishing between observations ("What"), mechanisms ("How"), and implications ("So What").

You can activate this deep mode using the `interpret_agent()` function.

```{r}
#| label: interpret-agent
#| eval: false
## Kept off: the multi-agent pipeline runs three LLM calls per cluster (Cleaner,
## Detective, Storyteller) on top of the single-call examples above, which would
## push this chapter's cost per build well past its budget. To enable it once the
## build is known to be green, replace `eval: false` with `eval: !expr has_ai` and
## wrap the result in `interpret_failed()` as the examples above do.
# Provide biological context to help Agent Cleaner
context <- "scRNA-seq analysis of CD8+ T cells in Tumor Microenvironment, comparing Exhausted vs. Naive states."

res <- interpret_agent(edo, context = context)
```

## Knowledge-Guided Interpretation

Enrichment analysis often treats genes as a "bag of words," ignoring their interactions and expression changes. The **Knowledge-Guided Interpretation** mode injects external knowledge to empower the AI's reasoning.

### 1. PPI Networks and Hub Genes
By setting `add_ppi = TRUE`, the system fetches protein-protein interaction data (from STRING). The AI can then identify functional modules (e.g., a TCR signaling complex) rather than just isolated genes.

### 2. Expression Trends
By providing `gene_fold_change`, the AI can infer the direction of pathway activity. For example, if the "Apoptosis" pathway is enriched but pro-apoptotic genes are downregulated while anti-apoptotic genes are upregulated, the AI will correctly interpret this as "Apoptosis Resistance."

```r
# Prepare a named vector of fold changes
gene_list <- c("CD8A" = 2.5, "PDCD1" = 1.8, "GZMB" = 3.2)

res <- interpret(edo, 
                 task = "cell_type",
                 add_ppi = TRUE,                # Enable PPI network analysis
                 gene_fold_change = gene_list   # Inject expression data
)
```

## Reference-Guided Interpretation

For cell type annotation, LLMs can sometimes hallucinate. To prevent this, `clusterProfiler` supports **Reference-Guided Interpretation**.

### Prior Knowledge Injection
You can provide "prior knowledge" (e.g., results from SingleR, scGPT, or manual rough annotation) to the AI. The AI acts as a validator and refiner:
*   **Validation**: Checks if pathway evidence supports the prior label.
*   **Refinement**: Refines a broad label (e.g., "T cell") into a specific state (e.g., "Proliferating CD8+ T cell") based on pathway activity.
*   **Correction**: Flags potential misannotations if the evidence contradicts the prior.

```{r}
#| label: interpret-prior-knowledge
#| eval: false
## Kept off: assigns to `res` and prints nothing, so running it would only add
## an API call per build without changing the rendered page.
# Prior knowledge from SingleR
my_priors <- c("Cluster1" = "T cells")

res <- interpret(edo, prior = my_priors, task = "cell_type")
```

### Hierarchical Interpretation
For complex datasets, `interpret_hierarchical()` mimics the human thought process of annotating major lineages first (e.g., Myeloid) and then subtypes (e.g., M1 Macrophage). It enforces lineage constraints to prevent impossible annotations (e.g., a T cell subtype appearing within a Myeloid cluster).

This approach is also highly applicable to **Single-cell Trajectory Inference**. Developmental processes inherently follow a hierarchical structure (e.g., Stem Cell -> Progenitor -> Terminally Differentiated Cell). By utilizing this hierarchical relationship, `interpret_hierarchical()` can provide context-aware interpretations that respect the biological differentiation path, ensuring that downstream states are interpreted within the context of their upstream progenitors.

```r
# Mapping between minor and major clusters
cluster_mapping <- c(
  "SubCluster1_1" = "MajorCluster1", 
  "SubCluster1_2" = "MajorCluster1",
  "SubCluster2_1" = "MajorCluster2"
)

# Hierarchical interpretation
res_hier <- interpret_hierarchical(
    x_minor = enrich_minor, 
    x_major = enrich_major, 
    mapping = cluster_mapping
)
```

## Gene-Based Fallback Mode

In real-world research, enrichment analysis sometimes fails to return significant pathways due to small gene sets or background noise.

`clusterProfiler` introduces a **Gene-Based Fallback Mode**. When no enriched pathways are found, the agent does not simply error out. Instead, it:
1.  Directly analyzes the function of the input genes.
2.  Retrieves PPI networks for these genes.
3.  Infers biological function based on gene connectivity and function, providing a "Medium Confidence" report instead of an empty result.

```r
# When enrichment fails, this still works
res <- interpret_agent(tough_genes, add_ppi = TRUE)
```

## Visualization: From Story to Figure

Text reports are great, but graphical representations are often preferred for presentations. The `plot()` method for interpretation results uses `ggtangle` (a grammar of graphics for networks) to visualize the AI-inferred regulatory network.

The resulting plot highlights:
*   **Key Drivers**: Central nodes identified by the AI.
*   **Activation (Green)** / **Inhibition (Red)**: Regulatory relationships inferred from data.
*   **Interactions (Grey)**: Physical associations.

```r
# Visualize the interpretation result
plot(res)
```

This feature creates a closed loop: from **Enrichment** (Statistics) to **Interpretation** (Insight) and finally to **Visualization** (Communication).

## Best Practices and Common Questions

The examples above are deliberately small. This section collects the guidance
that matters once you point `interpret()` at a real analysis: how to write
`context`, how to choose and switch models, what multi-model workflows are (and
are not) supported, and how to keep the report honest.

### How Specific Should `context` Be?

`context` is the only place where the model learns *what experiment produced the
gene list*, and it is the highest-leverage input you control. An enrichment
table is ambiguous on its own: "extracellular matrix organization" is
up-regulated in fibrosis, in wound healing and in tumour invasion alike. The
model cannot tell those apart, so it averages over them and returns a bland
report. A specific context removes the ambiguity.

A weak context names only the assay:

```r
context <- "scRNA-seq of T cells"
```

A strong context answers six questions in one or two sentences:

| Question | Example |
| :--- | :--- |
| Species and tissue | human breast tumour, bulk tissue |
| Disease / biological setting | breast carcinoma, tumour grade |
| Experimental design | 200 samples from the MAINZ cohort |
| Contrast tested, and its direction | grade III vs. grade I; positive values are higher in grade III |
| Cell type / composition | bulk tissue, no cell-type resolution |
| Hypothesis under test | progression involves loss of epithelial differentiation and ECM remodelling |

Written out, that is still only a few lines:

```r
context <- paste(
  "Human breast-cancer transcriptome (MAINZ cohort, bulk tumour tissue, 200",
  "samples). Genes are ranked by the log2 ratio of geometric-mean expression in",
  "grade III versus grade I tumours, so positive values are higher in the",
  "higher-grade tumours. No cell-type resolution is available. Hypothesis:",
  "progression from grade I to grade III involves loss of epithelial",
  "differentiation and increased extracellular-matrix remodelling."
)
```

Two rules of thumb follow. First, state the **direction of the contrast** --
"A vs. B" without saying which side is up is a common way to get a report with
the regulation backwards. Second, **do not put the answer in the context**: if
you write "we expect apoptosis to be suppressed", the model will often agree
with you regardless of the data. Describe the design, not the conclusion you are
hoping for. (If your data are from single cells, add the cell type or the
compartment the cluster belongs to; for bulk tissue, saying that no cell-type
resolution exists is more useful than omitting the point.)

### Choosing a Model and Switching Provider

Model selection is delegated to `aisdk`. `interpret()` itself has no model
registry and no provider logic: when `model` is left at its default (`NULL`) it
uses whatever `aisdk` currently considers the default, and `model = "provider:model"`
overrides that default for a single call.

```r
library(aisdk)

# Session-wide default, used by every interpret() call that omits `model`
set_model("deepseek:deepseek-v4-flash")
get_model()
```

The setup chunk on this page calls `set_model()` once, so every example that
omits `model` uses DeepSeek. To send one call elsewhere, pass `model` directly:

```r
res <- interpret(edo, model = "openai:gpt-4o")
```

Credentials follow the same rule. `aisdk` reads each provider's key from an
environment variable: `OPENAI_API_KEY`, `ANTHROPIC_API_KEY` and `GEMINI_API_KEY`
for the three providers bundled with `aisdk`, and the corresponding variable for
the adapters in `aisdk.providers` (for example `DEEPSEEK_API_KEY`, which is what
this chapter uses). Put the key in a project `.Renviron` or export it in the
shell; `interpret()` never takes a key argument. If you never call
`set_model()`, `aisdk` falls back to its built-in default, `openai:gpt-4o` --
which is why the examples you may have seen elsewhere appear to use GPT-4o by
default.

`aisdk` ships a catalogue of the providers and models it knows about:

```r
list_models()            # every provider and model aisdk knows about
list_models("openai")    # just one provider
```

Which one should you use? `interpret()` only needs a model that can follow the
structured-output schema. In practice the useful split is between small, cheap
models -- fine for first-pass triage -- and larger reasoning models, which tend
to produce more coherent mechanistic narratives but cost more and consume more
of the output budget (see `max_tokens` below). Whatever you choose, keep it
fixed once you start comparing reports: different models write different
reports, and that is the subject of the next subsection.

### Multi-Model and Consensus Workflows

There is **no built-in consensus or ensemble mode** in `interpret()` or
`interpret_agent()`. A call goes to exactly one model. `interpret_agent()` is a
multi-*agent* pipeline, not a multi-*model* one: its three agents share a single
model.

What you *can* do today is run the same input through several models and compare
the reports yourself. Because `model =` is a per-call argument, this needs no
extra machinery:

```r
res_deepseek <- interpret(edo, model = "deepseek:deepseek-v4-flash")
res_gpt      <- interpret(edo, model = "openai:gpt-4o")
```

Each call is a separate, billable request, and the two reports will not be
identical even for the same model (see below). A pragmatic middle ground is to
use one model to write the report and a second, independent call only to
*criticise* it: ask a different model whether every gene and pathway named in
the narrative is actually present in the input table. That is a manual
workflow, but it catches the most common failure mode -- plausible-sounding
genes that were never in the data. If what you need is provider *failover*
rather than consensus, note that `aisdk` provides `generate_text_fallback()` for
its own generation calls, but `interpret()` does not expose a fallback argument.

### Practical Guidance and Pitfalls

**Non-determinism.** `interpret()` samples at `temperature = 0.3` by default, so
two calls with identical input will differ in wording and occasionally in
emphasis. This is expected, not a bug. It has a direct consequence for
reproducible builds: Quarto's chunk cache (`cache: true`) is what keeps the
published page stable -- the LLM call runs once and its output is reused on
later renders. Clear the cache and you will get a different report. If you need
the exact same text in two places, cache the object rather than calling twice.

**Hallucination.** The system prompt instructs the model to ground every claim
in the supplied table and to cite supporting terms, and the structured output
forces it to name genes in `refined_network` and `network_evidence`. That
reduces hallucination; it does not eliminate it. Before you quote a report,
check the two things that are cheap to check: every gene named in the narrative
appears in the `geneID` (or, for GSEA, `core_enrichment`) column of the input
table, and every pathway named appears in its `Description` column. If you set
`add_ppi = TRUE`, remember that the edges the model cites come from a STRING
query, not from your data.

**`n_pathways`.** Only the top `n_pathways` terms (default 20) are sent to the
model, ordered by adjusted p-value (`p.adjust`, falling back to `pvalue`).
Raising it gives the model more to work with but dilutes the signal and enlarges
the prompt; lowering it forces a tighter focus. For a `compareClusterResult`,
`interpret()` splits the result by cluster and returns one report per cluster,
so the number of calls -- and the cost -- scales with the number of clusters,
not with `n_pathways`.

**`max_tokens`.** The default is 8192. Reasoning models spend part of that
budget deliberating before they emit the JSON, and if they run out, the reply is
truncated and cannot be parsed; the symptom is a warning mentioning
`finish_reason: length` and a placeholder report. If you hit it, raise
`max_tokens` (the diagnostic suggests 16384 or more) or switch to a model that
does not reason before answering.

**Seeing the cost.** `interpret()` returns only the parsed report; token usage
is not part of the returned object. Run with `verbose = TRUE` to print the raw
response and the prompt/completion/total token counts -- the practical way to
watch spend while you tune `n_pathways` and `max_tokens`.

**`add_ppi`.** Setting `add_ppi = TRUE` queries STRING for the genes in the top
terms and injects the edge list into the prompt, which lets the model identify
functional modules rather than isolated genes. It requires network access, adds
latency, and is a second failure point; the query is wrapped so that a STRING
outage degrades to "no network evidence" instead of an error. Leave it at its
default (`FALSE`) unless module-level reasoning is what you need.

**`gene_fold_change`.** Passing a named numeric vector of log fold changes adds
the direction of regulation to the prompt. Note the limitation: the helper
intersects your vector with the genes in the result's `geneID` column, and a
`gseaResult` table has no `geneID` column (it carries `core_enrichment`
instead). So `gene_fold_change` is used for over-representation results
(`enrichResult`, `compareClusterResult`) and silently ignored for GSEA results
-- which is harmless, because a GSEA result already encodes direction in the
sign of its enrichment score.

**Ranking.** For over-representation results the terms are ordered by
`p.adjust`; for GSEA the direction the model sees is the sign of `NES`. If your
GSEA ranked list was built so that positive means "up in the condition of
interest", say so in `context` -- otherwise the model has to guess the
orientation.

**Validating a report before trusting it.** Treat the output as a first draft
from a competent but over-eager collaborator. The minimum checks are the two
above (every gene and pathway exists in the input); beyond that, ask whether the
narrative is *falsifiable* by your data -- if it claims a mechanism, is there a
term or gene in the table that could have contradicted it? A report that only
restates the pathway names in longer sentences has added no information, and
sharpening `context` or raising `n_pathways` is usually the fix.

## Worked Example: From Enrichment to a Report

This example runs end to end: it builds a GSEA result offline from the `geneList`
shipped with `DOSE`, interprets it with a specific `context`, and prints the
report. It is the `interpret()` counterpart of the `interpret_agent()` call that
users often start from; `interpret_agent()` itself is kept disabled on this page
because it makes several LLM calls per build.

```{r}
#| label: interpret-worked-example
#| eval: !expr has_ai
library(clusterProfiler)
library(org.Hs.eg.db)

data(geneList, package = "DOSE")

# A real enrichment result, computed offline: GSEA over GO cellular components.
# geneList is the DOSE demo list -- log2 ratios of grade III vs. grade I tumours
# from the MAINZ breast-cancer cohort.
gse <- gseGO(
    geneList      = geneList,
    OrgDb         = org.Hs.eg.db,
    keyType       = "ENTREZID",
    ont           = "CC",
    minGSSize     = 15,
    maxGSSize     = 100,
    eps           = 0,
    pvalueCutoff  = 0.05,
    pAdjustMethod = "BH"
)

# A context that names the species, tissue, design, contrast and hypothesis.
context <- paste(
    "Human breast-cancer transcriptome (MAINZ cohort, bulk tumour tissue).",
    "Genes are ranked by the log2 ratio of geometric-mean expression in grade",
    "III versus grade I tumours, so positive values are higher in the",
    "higher-grade tumours. Hypothesis: progression from grade I to grade III",
    "involves loss of epithelial differentiation and increased",
    "extracellular-matrix remodelling."
)

report <- interpret(gse, context = context, n_pathways = 15)

if (interpret_failed(report)) {
    stop("interpret() returned no report -- check the API key and quota:\n",
         report$overview)
}

report
```

The printed report follows the schema described at the top of this chapter. Read
it in this order:

1.  **1. Overview** -- the model's one-paragraph summary. If it is vague, the
    `context` was probably too vague.
2.  **2. Regulatory Drivers** and **3. Key Mechanisms** -- the "who" and the
    "how". Check every named gene against your input table.
3.  **5. Hypothesis** -- the "so what". This is the part worth arguing with.
4.  **6. Narrative Draft** -- a paragraph written for a Results or Discussion
    section. It is a draft: edit it, do not paste it.
5.  **7. Refined Regulatory Network** -- populated only when the model can
    support edges; `plot(report)` draws the same network with `ggtangle` when it
    is available.

