# Miscellaneous topics {-}

## Leading Edge Analysis

Leading edge analysis is a powerful feature in GSEA that identifies the core set of genes driving the enrichment signal. It reports three key metrics:

- **Tags**: Percentage of genes contributing to the enrichment score
- **List**: Position in the ranked list where the enrichment score is attained
- **Signal**: Strength of the enrichment signal

DOSE, clusterProfiler, and ReactomePA all support leading edge analysis and can report the core enriched genes that contribute to the enrichment.

### Core Enriched Genes Extraction

After performing GSEA, the results object contains a `core_enrichment` column that lists the core genes responsible for each enriched term:

```{r}
library(DOSE)
data(geneList)
x <- gseDO(geneList)
head(x)
```

The output includes the core enriched genes in Entrez ID format for each significant term.

### Enhancing Readability with setReadable

To make the results more interpretable, use `setReadable()` to convert Entrez IDs to gene symbols:

```{r}
library(clusterProfiler)
y <- setReadable(x, 'org.Hs.eg.db')
head(y, 2)
```

This transformation makes the core enrichment results much more readable and biologically meaningful.

For visualization of leading edge analysis results using `cnetplot`, please refer to the [enrichplot chapter](enrichplot.html).

## Non-Model Plant Annotation with clusterProfiler

For non-model plants and other organisms lacking standard annotation packages, clusterProfiler can be used with custom annotation data obtained from tools like eggNOG.

### Workflow Overview

1. **Annotation with eggNOG**: Use the eggNOG web server to annotate protein sequences
2. **Parse eggNOG Results**: Extract GO and KEGG annotations using custom scripts
3. **Enrichment Analysis**: Use clusterProfiler's `enricher()` function with custom annotation data

### Key Steps

#### 1. eggNOG Annotation
Upload protein sequences to the [eggNOG mapper](http://eggnog-mapper.embl.de/) with appropriate parameters for your organism.

#### 2. Parsing eggNOG Results
Use [Python scripts](https://github.com/Hua-CM/HuaSmallTools/tree/master/parse) to process eggNOG output files:

```python
# Parse GO ontology file
python parse_go_obofile.py -i go-basic.obo -o go.tb

# Parse eggNOG annotations with reference species filtering
python parse_eggNOG.py -i panax_ginseng.annotations \
                       -g go.tb \
                       -O ath,osa \
                       -o output_directory
```

This generates two key files:
- `GOannotation.tsv`: GO term annotations
- `KOannotation.tsv`: KEGG pathway annotations

#### 3. Enrichment Analysis with clusterProfiler

```{r}
#| eval: false

library(clusterProfiler)

# Read annotation files
KOannotation <- read.delim("KOannotation.tsv", stringsAsFactors=FALSE)
GOannotation <- read.delim("GOannotation.tsv", stringsAsFactors=FALSE)
GOinfo <- read.delim("go.tb", stringsAsFactors=FALSE)

# Your gene list
gene_list <- c("gene1", "gene2", "gene3")  # Replace with your actual gene list

# GO enrichment (Molecular Function as example)
GOannotation_split <- split(GOannotation, GOannotation$level)
enricher(gene_list,
          TERM2GENE = GOannotation_split[['molecular_function']][c(2,1)],
          TERM2NAME = GOinfo[1:2])

# KEGG enrichment
enricher(gene_list,
          TERM2GENE = KOannotation[c(3,1)],
          TERM2NAME = KOannotation[c(3,4)])
```

### Advantages
- Works for any organism with protein sequences
- Uses reliable eggNOG annotation pipeline
- Flexible reference species filtering for KEGG

### Considerations
- Requires intermediate Python scripting
- Performance may vary with dataset size
- Manual integration of annotation and analysis steps

This approach enables comprehensive functional enrichment analysis for non-model organisms using clusterProfiler's powerful enrichment capabilities combined with custom annotation data.

<!--
Planned additions:
- GO utilities and redundancy reduction methods
- Data frame interface and accessor tips
- Extracting genes by term/pathway (geneInCategory)
- Practical troubleshooting notes and best practices
-->
