# Protein-Protein Interaction Network


```{r}
#| echo: false
#| message: false
#| cache: false

library(clusterProfiler)
library(org.Hs.eg.db)
```

`clusterProfiler` provides the `getPPI()` function to retrieve Protein-Protein Interaction (PPI) information from the STRING database. This allows users to explore the functional modules and interactions among genes of interest or genes enriched in specific pathways.

To visualize these networks, we recommend using the `ggtangle` package, which provides a grammar of graphics for networks, allowing for flexible and aesthetically pleasing visualizations. It is worth noting that network visualization functions in `enrichplot`, such as `cnetplot` and `emapplot`, are implemented based on `ggtangle`.

## Retrieving PPI for a Gene List

You can retrieve the PPI network for a specific list of genes using `getPPI()`. You need to specify the taxonomy ID (e.g., `9606` for human).

```{r}
library(clusterProfiler)
library(ggtangle)
library(ggplot2)

genes <- c("TP53", "BRCA1", "BRCA2", "S100A6")
g <- getPPI(genes, taxID="9606")
```

## Retrieving PPI for Enrichment Results

A more common scenario is to explore the interactions among genes enriched in a specific pathway. `getPPI()` works seamlessly with enrichment result objects (e.g., from `enrichKEGG`).

```{r}
data(geneList, package="DOSE")
de <- names(geneList)[1:100]
x <- enrichKEGG(de)
y <- setReadable(x, 'org.Hs.eg.db', 'ENTREZID')

# Get PPI for the first pathway
g <- getPPI(y, 1)
```

You can also specify the pathway ID directly or retrieve PPIs for multiple pathways.

```{r}
# Get PPI for a specific pathway ID
g2 <- getPPI(y, "hsa04110")

# Get PPI for the top 3 pathways
g3 <- getPPI(y, 1:3)
```

## Adding Neighbors

Sometimes it is useful to include neighbors of the query genes to see the broader context. You can use the `add_nodes` parameter.

```{r}
# Add 10 neighbor nodes
g_extended <- getPPI(y, 1, add_nodes = 10)
```

## Visualization with ggtangle

Instead of using the default `plot()` method, `ggtangle` allows us to visualize the network using `ggplot2` syntax. This offers greater flexibility in customizing the appearance of nodes and edges.

```{r}
# Basic visualization
library(ggtangle)
p <- ggplot(g) + geom_edge() + geom_point()
p
```

### Mapping Fold Change

One of the powerful features of `ggtangle` is the ability to map external data, such as fold changes, onto the network. This helps in understanding the expression patterns within the interaction network.
 
We can create a data frame containing the fold change information and attach it to the plot using the `%<+%` operator. This operator is originally introduced in the `ggtree` package [@yu_two_2018] for attaching data to a phylogenetic tree object.

```{r}
# Prepare fold change data
# geneList contains log2FC values, names are Entrez IDs
# Since 'y' was setReadable, the nodes in 'g' are gene symbols.
# We need to match the IDs.

library(org.Hs.eg.db)
fc <- geneList
names(fc) <- mapIds(org.Hs.eg.db, keys=names(fc), column="SYMBOL", keytype="ENTREZID")

# Create a data frame for mapping
# 'name' column should match the node names in the network
d <- data.frame(name = names(fc), logFC = fc)

# Map data to the plot
ggplot(g) %<+% d + 
  geom_edge() + 
  geom_point(aes(color = logFC), size = 8) + 
  shadowtext::geom_shadowtext(aes(label = name), color="black", bg.color="white") +
  #scale_color_gradient2(low = "blue", mid = "white", high = "red") +
  enrichplot::set_enrichplot_color(reverse=F) +
  theme_void()
```

This visualization allows you to simultaneously see the interaction structure (from STRING) and the functional state (up/down-regulation) of the genes, providing a more comprehensive view of the biological context.
